# Differences in the gut microbiota community and funciton of canids

## Filter data

```{r load_data_community, comment="", message=FALSE, warning=FALSE}
load("resources/data.Rdata")
load("output/alpha_all.RData")
```

## Sample overview
```{r}
sample_metadata %>% 
  summarise(total_samples=n())
```

```{r}
sample_metadata %>% 
  group_by(dog_type, season) %>% 
  summarise(n=n())
```

## Individuals

```{r}
sample_metadata  %>% 
  filter(country=="Greenland") %>% 
  group_by(dog_type, season, day) %>% 
  summarise(n=n())
```

```{r}
remove_sled <- sample_metadata %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample)

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")
genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)

phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()
phylo_genome <- genome_counts_filt %>%
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  filter(genome %in% rownames(phylo_genome)) %>%
  mutate(genome2 = genome) %>% 
  column_to_rownames("genome2") %>%
  dplyr::select(domain, phylum, class, order, family, genus, species, genome) %>% 
  as.matrix() %>%
  tax_table()

physeq_genome <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
```

## Taxonomy overview 

### Stacked barplot
```{r taxonomy_barplot_canids, fig.height=8, fig.width=20, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
    geom_bar(stat="identity", colour="white", linewidth=0.1) + 
    scale_fill_manual(values=phylum_colors) +
  facet_grid(~ country, scale="free", space="free")+
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 8, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```
```{r taxonomy_barplot, fig.height=10, fig.width=15, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(., genome_metadata, by = join_by(genome == genome)) %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
  geom_bar(stat="identity", colour="white", linewidth=0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_grid(~factor(dog_type, labels=c("companion" = "Companion dogs", "sled_dog" = "Sled dogs", "wolf" = "Wolves")), scale="free", space="free")+
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

**Number of bacteria phyla**

```{r phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Number of Archaea phyla**

```{r arch, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

### Genus and species annotation

**Number of MAGs without species-level annotation**
```{r nonspe, comment="", message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "") %>%
  summarize(Mag_nospecies = n())%>%
  select(Mag_nospecies) %>% 
  pull()
```
```{r nonspe_phylum, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage_nosp=100*count_nospecies/count_total) %>% 
  arrange(percentage_nosp) %>% 
  tt()
```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", message=FALSE, warning=FALSE}
nmags <- nrow(genome_metadata)
nonspecies <- genome_metadata %>%
    filter(species == "") %>%
    nrow()
perct <- nonspecies*100/nmags
perct
```

**Number of MAGs without genera-level annotation**
```{r nongenera, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
    filter(genus == "") %>%
    nrow()
cat(nongenera)
```

**Percentage of MAGs without genera-level annotation**
```{r gen_percet, comment="", message=FALSE, warning=FALSE}
perct_g <- nongenera*100/nmags
perct_g
```

### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,dog_type) %>%
  summarise(relabun=sum(count))
```

#### Origin: Dogs vs Sled dogs vs Wolves
```{r taxonomy_phylum_summary_origin, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T),
    Dog_mean = mean(relabun[dog_type == "companion"] * 100, na.rm = T),
    Dog_sd = sd(relabun[dog_type == "companion"] * 100, na.rm = T),
    Sled_dog_mean = mean(relabun[dog_type == "sled_dog"] * 100, na.rm = T),
    Sled_dog_sd = sd(relabun[dog_type == "sled_dog"] * 100, na.rm = T),
    Wolf_mean = mean(relabun[dog_type == "wolf"] * 100, na.rm = T),
    Wolf_sd = sd(relabun[dog_type == "wolf"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    "Companion dogs" = str_c(round(Dog_mean, 3), "±", round(Dog_sd, 3)),
    "Sled dogs" = str_c(round(Sled_dog_mean, 3), "±", round(Sled_dog_sd, 3)),
    Wolves = str_c(round(Wolf_mean, 3), "±", round(Wolf_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(phylum, Total, "Companion dogs", "Sled dogs", Wolves) %>%
  rename(Phylum=phylum) %>% 
  tt()
```

**Using CLR transformation**

```{r}
set.seed(0206)
phylo_counts <- genome_counts_filt %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_geno_meta <- genome_metadata %>%
  column_to_rownames(var="genome")%>%
  as.matrix()%>%
  tax_table()

physeq_genome_filt <- phyloseq(phylo_counts,phylo_geno_meta)
physeq_phylum <- tax_glom(physeq_genome_filt, taxrank = "phylum")
physeq_phylum_clr <- microbiome::transform(physeq_phylum, "clr")

df <- psmelt(physeq_phylum_clr)

results <- df %>%
  left_join(., sample_metadata, by=join_by("Sample"=="sample")) %>% 
  group_by(phylum) %>%
  do({
    model <- lm(Abundance ~ dog_type, data = .)
    as.data.frame(pairs(emmeans(model, ~ dog_type), adjust = "BH"))
  }) %>%
  ungroup()%>% 
  filter(p.value < 0.05) 

```
```{r}
results %>% 
  filter(contrast=="companion - sled_dog")
```

```{r}
results %>% 
  filter(contrast=="companion - wolf")
```

```{r}
results %>% 
  filter(contrast=="sled_dog - wolf")
```

## Alpha diversity
```{r alpha_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  filter(genome %in% genome_metadata_bacteria$genome) %>% 
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample))
```

```{r save_alpha, eval=FALSE}
save(alpha_div, file="output/alpha_all.RData")
```

```{r alpha_div_diets_summary_all, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
     total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Dog_mean = mean(value[dog_type == "companion"], na.rm = T),
    Dog_sd = sd(value[dog_type == "companion"], na.rm = T),
    Sled_dog_mean = mean(value[dog_type == "sled_dog"], na.rm = T),
    Sled_dog_sd = sd(value[dog_type == "sled_dog"], na.rm = T),
    Wolf_mean = mean(value[dog_type == "wolf"], na.rm = T),
    Wolf_sd = sd(value[dog_type == "wolf"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    "Companion dogs" = str_c(round(Dog_mean, 3), "±", round(Dog_sd, 3)),
    "Sled dogs" = str_c(round(Sled_dog_mean, 3), "±", round(Sled_dog_sd, 3)),
    Wolves = str_c(round(Wolf_mean, 3), "±", round(Wolf_sd, 3))
  ) %>%
  arrange(-Sled_dog_mean) %>%
  dplyr::select(alpha, Total, "Companion dogs", "Sled dogs", Wolves) %>%
  rename(Alpha=alpha) %>% 
  tt()
```

```{r alpha_plots, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
type_colors <- c('#41B6C0','#E5D388',"#bb6749")
new_labels <- c(
    "companion" = "Companion dogs",
    "sled_dog" = "Sled dogs",
    "wolf" = "Wolves"
    )

alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(
    y = value,
    x = dog_type,
    group = dog_type,
    color = dog_type,
    fill = dog_type
  )) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_color_manual(
    name = "Canids",
    values = type_colors,
    labels = new_labels
  ) +
  scale_fill_manual(
    name = "Canids",
    values = type_colors,
    labels = new_labels
  ) +
facet_wrap(
  ~ factor(
    metric,
    levels = c("richness","neutral","phylogenetic"),
    labels = c("Richness","Neutral","Phylogenetic")
  ),
  scales = "free",
  ncol = 5
) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  kruskal_test(value ~ dog_type) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))%>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  wilcox_test(value ~ dog_type, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

## Gut microbiota composition based on beta diversity

```{r}
type_colors <- c('#41B6C0','#E5D388',"#bb6749")
new_labels <- c(
    "companion" = "Companion dogs",
    "sled_dog" = "Sled dogs",
    "wolf" = "Wolves"
    )

load("output/beta.RData")
```

```{r hillpair, comment="", message=FALSE, warning=FALSE, eval=FALSE}
hillpair <- function(data,q=c(0,1,2),metric=c("S","C","U","V"),tree,dist,tau,out="dist"){

  ###
  # Type of analysis detection
  ###

  #Hill numbers type definition
  if(missing(tree) && missing(dist)){hilltype="neutral"}
  if(!missing(tree) && missing(dist)){hilltype="phylogenetic"}
  if(missing(tree) && !missing(dist)){hilltype="functional"}
  if(!missing(tree) && !missing(dist)) stop("Phylogenetic and functional trait information cannot be added at once. Use either phylogenetic (tree) or functional (fun) information.")

  ###
  # Create pairwise tables
  ###

  # List of pairwise combinations
  pairs <- list()
  for (i in 1:(ncol(data)-1)) {
    for (j in (i + 1):ncol(data)) {
      pair <- data[, c(i, j), drop = FALSE]
      pairs <- append(pairs, list(pair))
    }
  }

  # List of pairwise sample names
  pairs_names <- list()
  for (i in 1:(ncol(data)-1)) {
    for (j in (i + 1):ncol(data)) {
      pair_names <- colnames(data[, c(i, j), drop = FALSE])
      pairs_names <- append(pairs_names, list(pair_names))
    }
  }

  ###
  # Estimate processing time
  ###
  pairs_length <- length(pairs_names)
  start.time <- Sys.time()
  if(hilltype == "neutral"){suppressMessages(pairs_dist <- lapply(pairs[1:10], hilldiss, q=q, metric=metric))}
  if(hilltype == "phylogenetic"){suppressMessages(pairs_dist <- lapply(pairs[1:10], hilldiss, q=q, metric=metric, tree=tree))}
  if(hilltype == "functional"){suppressMessages(pairs_dist <- lapply(pairs[1:10], hilldiss, q=q, metric=metric, dist=dist, tau=tau))}
  end.time <- Sys.time()
  time_estimate <- round((end.time - start.time)/10*pairs_length)
  options(warn=1)
  message(paste0("The estimated time for calculating the ",pairs_length," pairwise combinations is ",time_estimate+2, " seconds."))

  ###
  # Compute all pairwise distances
  ###

   if(hilltype == "neutral"){suppressMessages(pairs_dist <- lapply(pairs, hilldiss, q=q, metric=metric))}
   if(hilltype == "phylogenetic"){suppressMessages(pairs_dist <- lapply(pairs, hilldiss, q=q, metric=metric, tree=tree))}
   if(hilltype == "functional"){suppressMessages(pairs_dist <- lapply(pairs, hilldiss, q=q, metric=metric, dist=dist, tau=tau))}

  ###
  # Arrange output
  ###

  #Flatten list into matrix
  flatten_matrix <- function(matrix){
      matrix %>%
      as.data.frame() %>%
      pivot_longer(cols = everything(), names_to = "Variable") %>%
      select(value) %>%
      pull()
      }

    pairs_dist_matrix <- pairs_dist %>%
      lapply(flatten_matrix) %>%
      do.call(rbind,.) %>%
      as.data.frame()

    #List metric combinations
    if(length(q)>1){
      metric_combinations <- expand.grid(colnames(pairs_dist[[1]]),rownames(pairs_dist[[1]])) %>%
          mutate(name = paste(Var2, Var1, sep = "") ) %>%
          select(name)  %>%
          pull()
    }else{
      metric_combinations <- metric
    }

    #Assign column names to different metrics
    colnames(pairs_dist_matrix) <- metric_combinations

    #Add sample names
    pairs_dist_matrix <- pairs_dist_matrix %>%
      mutate(first=unlist(lapply(pairs_names, function(x) x[1]))) %>%
      relocate(first, .before = 1) %>%
      mutate(second=unlist(lapply(pairs_names, function(x) x[2]))) %>%
      relocate(second, .after = 1)

    #Convert table into distance object
    if (out == "dist"){
      create_distance_matrix <- function(column_name) {
        sample_names <- colnames(data)
        matrix_df <- matrix(
          0,
          nrow = length(sample_names),
          ncol = length(sample_names),
          dimnames = list(sample_names, sample_names)
        )
        pair_index <- cbind(
          match(pairs_dist_matrix$first, sample_names),
          match(pairs_dist_matrix$second, sample_names)
        )
        matrix_df[pair_index] <- pairs_dist_matrix[[column_name]]
        matrix_df[pair_index[, 2:1, drop = FALSE]] <- pairs_dist_matrix[[column_name]]
        matrix_df <- as.dist(matrix_df)
        return(matrix_df)
      }

      pairs_dist_matrix <- lapply(colnames(pairs_dist_matrix[3:ncol(pairs_dist_matrix)]), create_distance_matrix)
      names(pairs_dist_matrix) <- metric_combinations

      if(length(pairs_dist_matrix) == 1){
        pairs_dist_matrix <- pairs_dist_matrix[[1]]
      }
    }

  return(pairs_dist_matrix)

}
```

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

# beta_q1p <- genome_counts_filt %>%
#   filter(genome %in% genome_metadata_bacteria$genome) %>% 
#   column_to_rownames(., "genome") %>%
#   filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
#   select_if(~!all(. == 0)) %>%
#   hillpair(., q = 1, tree = genome_tree)

save(beta_q0n, 
     beta_q1n, 
     #beta_q1p, 
     file = "output/beta.Rdata")
```

```{r q1, comment="", message=FALSE, warning=FALSE, eval=FALSE}
# table <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
#   select(where(~ !all(. == 0)))
# labs <- colnames(table)
# 
# beta_q1n <- hillpair(table, q = 1)
# attr(beta_q1n$S, "Labels") <- labs
# beta_q1n <- beta_q1n$S
# save(beta_q1n,file="output/beta_q1n.RData")
```

```{r}
### CLR
# beta_q1n <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(.) > 0) %>%
#   select(where(~ sum(.) > 0)) %>%
#   t() %>%
#   (\(x) x + 1)() %>%
#   clr() %>%
#   dist()
```

```{r}
### Bray curtis
# beta_q1n_bray <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(.) > 0) %>%
#   select(where(~ sum(.) > 0)) %>%
#   t() %>%
#   vegdist(method = "bray")
```

```{r}
## Hellinger
# mat <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(.) > 0) %>%
#   select(where(~ sum(.) > 0)) %>%
#   t()
# 
# mat_hel <- decostand(mat, method = "hellinger")
# 
# beta_hel <- dist(mat_hel, method = "euclidean")
```

### Richness
```{r permanova__q0, comment="", message=FALSE, warning=FALSE}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]
beta_q0n_dis <- as.dist(beta_q0n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q0n_dis))) %>% 
  column_to_rownames(., "sample")

betadisper(beta_q0n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q0n_dis, sample_metadata_ordered$dog_type) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q0n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q0n_dis ~ dog_type,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

***Locality***

Significant
```{r}
pair_results <- pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$locality, perm = 999)
pair_results %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pair_results %>% 
  filter(p.adjusted>0.05)
```

***Dog group***

Significant
```{r}
pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$dog_type, perm = 999) %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$dog_type, perm = 999) %>% 
  filter(p.adjusted>0.05)
```

Centroids
```{r}
betadisper_obj <- betadisper(beta_q0n_dis, sample_metadata_ordered$dog_type)
dist_centroids <- dist(betadisper_obj$centroids)
```

### Neutral
```{r permanova__q1, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)
beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]
beta_q1n_dis <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1n_dis))) %>% 
  column_to_rownames(., "sample")

all(rownames(sample_metadata_ordered) == labels(beta_q1n_dis))

#betadisper(beta_q1n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q1n_dis, sample_metadata_ordered$dog_type)%>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n_dis ~ dog_type + country + locality,
        data = sample_metadata_ordered,
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q1n_dis ~ dog_type + country + locality,
        data = sample_metadata_ordered,
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q1n_dis ~ dog_type,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

***Pairwise***
Significant
```{r}
pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$dog_type, perm = 999) %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$dog_type, perm = 999) %>% 
  filter(p.adjusted>0.05)
```

Centroids
```{r}
betadisper_obj <- betadisper(beta_q1n_dis, sample_metadata_ordered$dog_type)
dist_centroids <- dist(betadisper_obj$centroids)
dist_centroids
```

***NMDS***
```{r generate_nmds, comment="", message=FALSE, warning=FALSE, eval=FALSE}
nmds_beta <- beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup()
save(nmds_beta,file="output/nmds_beta_1206.RData")

nmds_beta_country <- beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type, country) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup()
save(nmds_beta_country,file="output/nmds_beta_country_1206.RData")
```

```{r nmds_plot_type, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
load("output/nmds_beta.RData")
nmds_beta %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = 'Canids') 
```

```{r nmds_plot_country, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
load("output/nmds_beta_country.RData")

nmds_beta_country %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels)+
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = 'Canids')
```


#### PCoa country
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100
```

```{r pcoa_neutral_loc, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
centroids <- pcoa_df %>%
  group_by(dog_type, country) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )
centroids_loc <- centroids %>%
  mutate(group = paste(dog_type, country, sep = "_"))

pcoa_df1 <- pcoa_df %>%
  mutate(group = paste(dog_type, country, sep = "_"))

ggplot(pcoa_df1, aes(Axis.1, Axis.2, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(
    data = centroids_loc,
    aes(Axis.1, Axis.2, fill = group),
    color = "black",
    shape = 21,
    size = 6,
    show.legend = FALSE
  ) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Group"
  )  +
  theme_minimal() +
  theme(legend.text = element_text(size = 8)) +
  guides(color = guide_legend(ncol = 1))
```

***Centroid distances***

```{r centroid_distances_type_locality, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=12, fig.fullwidth=TRUE}
centroids_loc <- pcoa_df %>%
  group_by(dog_type, country) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )

centroids_loc <- centroids_loc %>%
  mutate(group = paste(dog_type, country, sep = "_"))


coords <- centroids_loc[, c("Axis.1", "Axis.2")] %>%
  as.data.frame()
rownames(coords) <- centroids_loc$group

dist_loc <- dist(coords)

dist_loc_df <- as.data.frame(as.table(as.matrix(dist_loc))) %>%
  rename(group1 = Var1,
         group2 = Var2,
         distance = Freq) %>%
  filter(group1 != group2)

ggplot(dist_loc_df, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank()) +
  labs(fill = "Centroids distance")
```
```{r}
 dist_loc_df %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(group1, group2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) %>% 
  arrange(distance)
```

#### PCoa dog type
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100

centroids <- pcoa_df %>%
  group_by(dog_type) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )
```
```{r pcoa_neutral_dogtype, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
ggplot(pcoa_df, aes(Axis.1, Axis.2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(data = centroids,
             aes(Axis.1, Axis.2, fill = dog_type),
             color = "black",
             shape = 21,
             size = 6,
             show.legend = FALSE) +
    scale_color_manual(
  name = "Canid",
  values = type_colors,
  labels = new_labels)+
scale_fill_manual(
  name = "Canid",        
  values = type_colors,
  labels = new_labels) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Canid"
  ) +
  theme_classic(base_size = 14)
```

Show samples closest to domestic dogs
```{r pcoa_neutral_type_sample, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
domestic_centroid <- centroids %>% filter(dog_type == "companion")
sled_centroid <- centroids %>% filter(dog_type == "sled_dog")

pcoa_df <- pcoa_df %>%
  mutate(dist_to_domestic = sqrt((Axis.1 - domestic_centroid$Axis.1)^2 +
                                 (Axis.2 - domestic_centroid$Axis.2)^2),
    dist_to_sled = sqrt((Axis.1 - sled_centroid$Axis.1)^2 +
                        (Axis.2 - sled_centroid$Axis.2)^2)
  )


closest_sled_to_domestic <- pcoa_df %>%
  filter(dog_type == "sled_dog") %>%
  arrange(dist_to_domestic) %>%
  slice_head(n = 5)

closest_samples <- closest_sled_to_domestic$sample

ggplot(pcoa_df, aes(Axis.1, Axis.2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(aes(color = dog_type), level = 0.95, linewidth = 1.2) +
  geom_point(
    data = centroids,
    aes(x = Axis.1, y = Axis.2, fill = dog_type),
    color = "black",
    shape = 21,
    size = 6,
    show.legend = FALSE
  ) +
  geom_point(
    data = pcoa_df %>% filter(sample %in% closest_samples),
    aes(x = Axis.1, y = Axis.2),
    shape = 21,
    fill = NA,
    color = "black",
    size = 5,
    stroke = 1.5
  ) +
  scale_color_manual(name = "Canid",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canid",
                    values = type_colors,
                    labels = new_labels) +
  geom_text_repel(
    data = closest_sled_to_domestic,
    aes(x = Axis.1, y = Axis.2, label = locality),
    size = 4,
    box.padding = 0.5,
    point.padding = 0.5,
    segment.color = "grey50"
  ) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Dog type"
  ) +
  theme_classic(base_size = 14)
```

**Centroid distances**
```{r cent, comment="", message=FALSE, warning=FALSE}
centroid_coords <- centroids[, c("Axis.1", "Axis.2")]
rownames(centroid_coords) <- centroids$dog_type

centroid_dist <- dist(centroid_coords)
centroid_dist_mat <- as.matrix(centroid_dist)

dist_table <- as.data.frame(as.table(centroid_dist_mat)) %>%
  rename(group1 = Var1, group2 = Var2, distance = Freq) %>%
  filter(group1 != group2) %>%
  arrange(distance)
```

```{r centroid_distances_type_figure, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.fullwidth=TRUE}
ggplot(dist_table, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_blank()) +
  labs(fill="Centroids distance")
```

#### Distance comparison test

This tests whether samples of the same dog type are more similar than different ones
```{r}
dist_mat <- as.matrix(beta_q1n_dis)

dist_df <- as.data.frame(dist_mat) %>%
  mutate(sample1 = rownames(.)) %>%
  pivot_longer(-sample1, names_to = "sample2", values_to = "distance") %>%
  filter(sample1 < sample2)

meta <- sample_metadata_ordered %>%
  mutate(sample = rownames(.))

dist_df <- dist_df %>%
  left_join(meta, by = c("sample1" = "sample")) %>%
  rename(dog1 = dog_type, loc1 = locality) %>%
  left_join(meta, by = c("sample2" = "sample")) %>%
  rename(dog2 = dog_type, loc2 = locality)

dist_df <- dist_df %>%
  mutate(
    same_dog = dog1 == dog2,
    same_locality = loc1 == loc2
  )
```

```{r}
dist_cross_loc <- dist_df %>%
  filter(!same_locality)

wilcox.test(distance ~ same_dog, data = dist_cross_loc)

ggplot(dist_cross_loc, aes(x = same_dog, y = distance)) +
  geom_boxplot() +
  labs(
    x = "Same dog type",
    y = "Beta diversity distance",
    title = "Distance comparison (across localities only)"
  )
```
If same_dog == TRUE has lower distances: suggests similarity within dog types across locations


## Differential abundance analysis
```{r}
load("output/ancombc_phylum_sled_companion_wolf.Rdata")
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
```

### Phylum level
```{r pair_all_phylum, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_types <- lapply(region_pairs_types, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

   fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome),
    fix_formula = "dog_type",
    tax_level = "phylum", 
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )
   
  res_table <- fit$res

  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})
names(results_types) <- sapply(region_pairs_types, paste, collapse = "_vs_")

save(results_types,
     file = "output/ancombc_phylum_sled_companion_wolf.Rdata")
```

#### Domestic vs sled_dog

**Enriched in companion dogs**
```{r}
companion <- results_types$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
companion %>%
  nrow() %>% 
  cat()
```

```{r}
companion
```

**Enriched in sled dogs**

```{r}
sled <- results_types$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)

sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled
```


#### Sled dogs vs wolves

**Enriched in sled dogs**
```{r}
sled <- results_types$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)
```

```{r}
sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled
```

**Enriched in wolves**

```{r}
wolf <- results_types$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)

wolf %>%
  nrow() %>%
  cat()
```

```{r}
wolf
```

#### All plots
```{r generate_all, comment="", message=FALSE,warning=FALSE, eval=FALSE}
taxa <- as.data.frame(physeq_genome@tax_table)  %>% 
  distinct(phylum)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxa, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (name in names(results_types)) {

  parts <- strsplit(name, "_vs_")[[1]]
  
  name1 <- parts[1]
  name2 <- parts[2]
  
  title <- paste0("Enriched in ", name1, "  ( − LFC )   vs   ", "Enriched in ", name2, "  ( + LFC )")
  
  tab <- results_types[[name]]

  coef <- grep("^lfc_dog_type", colnames(tab), value = TRUE)[1]
  qcol <- sub("^lfc_", "q_", coef)
  
  plot_data <- tab %>%
    filter(.data[[qcol]] < 0.05) %>%
    mutate(logFC = .data[[coef]])
  
  tax_color <- plot_data %>% 
    rename(phylum=1) %>% 
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()

  plot <- ggplot(plot_data, aes(x = reorder(taxon, logFC), y = logFC, fill = taxon)) +
    geom_col() +
    scale_fill_manual(values = tax_color) +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    labs(
      title = title,
      x = "Phylum",
      y = "Log-Fold Change",
      fill = "Phylum"
    )

  ggsave(paste0("output/figures/Fig_phylum_", name, ".png"),
         plot = plot,
         width = 6,
         height = 8)
}
```

### MAG level
```{r load_all, comment="", message=FALSE, warning=FALSE}
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
load("output/ancombc_pairwise_mag_sled_companion_sepa.Rdata")
```

```{r pair_all_mag, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_mags <- lapply(region_pairs_types, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

  fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome),
    fix_formula = "dog_type",
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )

  res_table <- fit$res

  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})

names(results_mags) <- sapply(region_pairs_types, paste, collapse = "_vs_")
save(results_mags,
     file = "output/ancombc_pairwise_mag_sled_companion_sepa.Rdata")
```

#### Companion vs Sled dogs

**Total number of enriched MAGs**
```{r}
results_mags$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  nrow
```

**Enriched in companion dogs**
```{r}
companion <- results_mags$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)

all <- companion %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

```{r}
perc <- 100-(noanno*100/all)
perc %>% 
  cat()
```

```{r}
companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, genus, family, everything()) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, genus, family, phylum, everything()) %>% 
  filter(species=="") %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count)
```

```{r}
companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

**Enriched in sled dogs**

```{r}
sled <- results_mags$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)

all <- sled %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

Percentage of bacteria annotated to species level
```{r}
perc <- 100-(noanno*100/all)
perc %>% 
  cat()
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, genus, family, everything())
```
```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, genus, family, phylum, everything()) %>% 
  filter(species=="") %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count)
```
```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)

sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

#### Sled dogs vs wolves

**Total number of enriched MAGs**
```{r}
results_mags$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  nrow
```

**Enriched in sled dogs**
```{r}
sled <- results_mags$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)

all <- sled %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```
```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, family, everything())
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

**Enriched in wolves**

```{r}
wolf <- results_mags$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)

all <- wolf %>%
  nrow()
all %>% 
  cat()
```

how many of the enriched bacteria were not annotated to species
```{r}
noanno <- wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, family, everything())
```

```{r}
wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)

wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

### All plots
```{r plots_diff, comment="", message=FALSE, warning=FALSE, eval=FALSE}
taxonomy <- data.frame(physeq_genome@tax_table) %>%
  distinct(phylum) %>% 
  rownames_to_column(., "taxon")

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (name in names(results_mags)) {

  parts <- strsplit(name, "_vs_")[[1]]
  
  name1 <- parts[1]
  name2 <- parts[2]
  
  title <- paste0("Enriched in ", name1, "  ( − LFC )   vs   ", "Enriched in ", name2, "  ( + LFC )")
  
  tab <- results_mags[[name]]

  coef <- grep("^lfc_dog_type", colnames(tab), value = TRUE)[1]
  qcol <- sub("^lfc_", "q_", coef)
  
  plot_data <- tab %>%
    filter(.data[[qcol]] < 0.05) %>%
    mutate(logFC = .data[[coef]]) %>% 
    left_join(genome_metadata, by=join_by("taxon"=="genome"))
  
  tax_color <- plot_data %>% 
    distinct(phylum) %>% 
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()

  plot <- ggplot(plot_data, aes(x = reorder(taxon, logFC), y = logFC, fill = phylum)) +
    geom_col() +
    scale_fill_manual(values = tax_color) +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 6),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    labs(
      title = title,
      x = "MAGs",
      y = "Log-Fold Change",
      fill = "Phylum"
    )

  ggsave(paste0("output/figures/Fig_mag_", name, ".png"),
         plot = plot,
         width = 10,
         height = 25)
}
```

### Species level
```{r}
genome_tax <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(genus=="",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus),
    species  = if_else(species=="",
                     if_else(str_detect(genus, "Unclassified"), genus, paste0(genus, "_Unclassified")),
                     species)
  )
```

```{r phy_sp, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  left_join(genome_tax[c(1,3,4,6,7,8)], by = "genome") %>% 
  select(-genome) %>%
  group_by(phylum, class, family, genus, species) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class, family, genus))%>%
  column_to_rownames("species") %>%
  otu_table(., taxa_are_rows = TRUE)

physeq_genome_sp <- phyloseq(phylo_genome, phylo_samples)
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
load("output/ancombc_species_sled_companion_wolf.Rdata")
```

```{r pair_all_dogtype, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_species <- lapply(region_pairs_types, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

   fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome_sp),
    fix_formula = "dog_type",
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )
   # Extract main results table
  res_table <- fit$res

  # Add comparison label
  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})
names(results_species) <- sapply(region_pairs_types, paste, collapse = "_vs_")
 save(results_species,
     file = "output/ancombc_species_sled_companion_wolf.Rdata")
```

#### Companion vs sled dogs

**Enriched in companion dogs**
```{r}
companion <- results_species$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)

all <- companion %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```
```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
companion
```

**Enriched in sled dogs**

```{r}
sled <- results_species$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)

all <- sled %>%
  nrow()
all %>% 
  cat()
```

```{r}
sled
```


#### Sled dogs vs wolves

**Enriched in sled dogs**
```{r}
sled <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)

sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled
```

**Enriched in wolves**

```{r}
wolf <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)

wolf %>%
  nrow() %>%
  cat()
```

```{r}
wolf
```


## Functional

```{r gift_analysis_func_all, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts), ]
genome_counts_filt <- genome_counts_filt %>%
  select(genome, all_of(sample_metadata$sample)) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome, ]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]

#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome, ]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>%
  select_if( ~ !is.numeric(.) || sum(.) != 0)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered, GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions, GIFT_db)

#Get community-weighed average GIFTs per sample
genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  column_to_rownames(., "genome")

GIFTs_elements_community <- to.community(GIFTs_elements_filtered, genome_counts_row, GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions, genome_counts_row, GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains, genome_counts_row, GIFT_db)
```


### MCI
```{r gitfs_functional_type}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type) %>%
  summarise(mean = round(mean(value), 3), sd = round(sd(value), 3)) %>%
  mutate(MCI = str_c(round(mean, 3), "±", round(sd, 3))) %>%
  select(dog_type, MCI) %>%
  rename("Canid" = "dog_type") %>%
  tt()
```

```{r}
MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample))

shapiro.test(residuals(lm(value ~ dog_type, data = MCI)))
bartlett.test(value ~ dog_type, data = MCI)

kruskal.test(value ~ dog_type, data = MCI)
pw <- pairwise.wilcox.test(x = MCI$value,
                           g = MCI$dog_type,
                           p.adjust.method = "BH")

wilcox_tbl <- pw$p.value %>%
  as.data.frame() %>%
  rownames_to_column("group1") %>%
  pivot_longer(-group1, names_to = "group2", values_to = "p_adjust") %>%
  filter(!is.na(p_adjust)) %>%
  mutate(
    comparison = paste(group1, "vs", group2),
    p_adjust = round(p_adjust, 3),
    sig = case_when(
      p_adjust < 0.001 ~ "***",
      p_adjust < 0.01  ~ "**",
      p_adjust < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(comparison, p_adjust, sig)

wilcox_tbl %>%
  rename("Compared groups" = comparison,
         "Adjusted p-value" = p_adjust) %>%
  tt()
```

### Community elements differences:

```{r}
function_matrix <- GIFTs_functions_community %>%
  as.data.frame() %>%
  as.matrix()

metadata <- sample_metadata %>%
  select(sample, dog_type) %>%
  column_to_rownames("sample")

metadata <- metadata[rownames(function_matrix), , drop=FALSE]
stopifnot(all(rownames(function_matrix) %in% rownames(metadata)))
stopifnot(all(rownames(function_matrix) == rownames(metadata)))
meta_row <- rownames_to_column(metadata, "sample")

adonis2(function_matrix ~ dog_type,
        data = metadata,
        method = "euclidean",
        permutations = 999)
```

```{r commun_wilcox_elem_type, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample")

uniqueGIFT_db <- unique(GIFT_db[c(2, 4, 5, 6)]) %>%
  unite("Function", Function:Element, sep = "_", remove = FALSE)

significant_elements <- element_gift %>%
  pivot_longer(-c(sample, dog_type),
               names_to = "trait",
               values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ dog_type)$p.value,
            .groups = "drop") %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05)

selected_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type),
               names_to = "trait",
               values_to = "value") %>%
  filter(trait %in% significant_elements$trait)

pairwise_results <- selected_data %>%
  group_by(trait) %>%
  dunn_test(value ~ dog_type, p.adjust.method = "BH") %>%
  ungroup() %>%
  filter(p.adj < 0.05) %>%
  mutate(comparison = paste(group1, "vs", group2)) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element))
```

```{r ele_plot_type, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
plot_elem <- ggplot(pairwise_results, aes(x = comparison, y = fct_rev(Function))) +
  geom_tile(aes(fill = p.adj), color = "white") +
  geom_text(aes(label = p.adj.signif), size = 4, color = "black") +
  scale_fill_gradient(low = "#b30000", high = "#fef0d9" , name = "Adjusted p-value") +
  theme_minimal() +
  labs(title = "Significant Pairwise Comparisons by Trait", x = "Locality Comparison", y = "Trait") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

#ggsave("Figures/plot_element_dog_type.png", plot = plot_elem, width = 10, height = 25, dpi = 300)

plot_elem
```

```{r barplot_ele_type2, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>%  select(sample, dog_type) %>% filter(), by = "sample")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% pairwise_results$trait) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element))

plot_data %>%
  ggplot(., aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal() +
  theme(strip.text = element_text(size = 18))
```

**Starch**
```{r starch_plot, fig.height=4, fig.width=6, fig.fullwidth=TRUE}
plot_data %>%
  filter(Element == "Starch") %>%
#  left_join(sample_metadata_origin %>% select(sample, origin), by = "sample") %>% 
  ggplot(., aes(x = dog_type, y = value, fill = dog_type, color=dog_type)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  theme_minimal()  +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  #geom_text_repel(aes(label = sample), size = 3, max.overlaps = Inf, show.legend = FALSE) +
  theme_classic()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  ) 
```
```{r}
selected_data %>%
  group_by(trait) %>%
  dunn_test(value ~ dog_type, p.adjust.method = "BH") %>% 
  ungroup()%>% 
  filter(trait=="D0203") 
```

Differences between companion dogs and sled dogs
```{r}
pairwise_results %>%
  filter(comparison == "companion vs sled_dog") %>%
  nrow
```

```{r barplot_ele_type, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
# element_gift <- GIFTs_elements_community %>% 
#   as.data.frame() %>% 
#   rownames_to_column(., "sample") %>% 
#   left_join(sample_metadata %>%  select(sample,dog_type), by="sample")

dog_sled <- pairwise_results %>%
  filter(comparison == "companion vs sled_dog")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sled$trait) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element)) 

plot_data %>%
  filter(dog_type != "wolf") %>%
  ggplot(., aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal() +
  theme(strip.text = element_text(size = 18))
```

```{r}
enrichment_results <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  filter(sample %in% sample_metadata$sample) %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample") %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sled$trait) %>%
  filter(dog_type != "wolf") %>%
  group_by(trait) %>%
  summarise(
    median_sled = median(value[dog_type == "sled_dog"]),
    median_domestic = median(value[dog_type == "companion"]),
    log2FC = log2((median_sled + 1) / (median_domestic + 1)),
    difference = median_sled - median_domestic
  )

enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  select(log2FC, Function) %>%
  arrange(desc(Function)) %>%
  tt()
```

```{r plot3_sled_dom, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,3,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

enrichment_code <- enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by=join_by(trait==Code_element))

unique_codes <- unique(enrichment_code$Code_function)

gift_colors <- read_tsv("resources/gift_colors.tsv") %>%
  filter(Code_function %in% unique_codes) %>%
  mutate(legend = str_c(Code_function, " - ", Function))
gift_color_values <- setNames(gift_colors$Color, gift_colors$legend)
```

```{r final_plot_sled_dom, comment="", message=FALSE, warning=FALSE, fig.height=12, fig.width=12, fig.fullwidth=TRUE}
enrichment_code %>%
  left_join(
    gift_colors %>% select(Code_function, Color, legend),
    by = join_by(Code_function == Code_function)
  ) %>%
  ggplot(aes(x = forcats::fct_reorder(Function, log2FC), y = log2FC, fill = legend)) +
  geom_col() +
  scale_fill_manual(values = gift_color_values) +
  geom_hline(yintercept = 0) +
  coord_flip() +
  geom_text(aes(75, 0.040), label = "Enriched in\n sled dogs", color = "#666666") +
  geom_text(aes(11, -0.09), label = "Enriched in\n companion dogs ", color = "#666666") +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_blank(),
    panel.background = element_blank()
  ) +
  xlab("Microbial Functional Capacity") +
  ylab("log2FC")
```


Number of enriched functions in each group
```{r}
sled_enr <- enrichment_code %>%
  left_join(., unique(GIFT_db[c(2, 5)]), by = join_by(trait == Code_element)) %>%
  filter(log2FC > 0) %>%
  group_by(Function.y) %>%
  summarise(Sled = n())

enrichment_code %>%
  left_join(., unique(GIFT_db[c(2, 5)]), by = join_by(trait == Code_element)) %>%
  filter(log2FC < 0) %>%
  group_by(Function.y) %>%
  summarise(Companion = n()) %>%
  full_join(sled_enr, by = "Function.y") %>%
  arrange(-Sled) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .))) %>%
  mutate(difference = Companion - Sled)
```

Differences between sled dogs and wolves
```{r barplot_ele_type2_wolf, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>%  select(sample, dog_type) %>% filter(), by =
              "sample")

dog_sledwolf <- pairwise_results %>%
  filter(comparison == "sled_dog vs wolf")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sledwolf$trait) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element))

plot_data %>%
  filter(dog_type != "domestic") %>%
  ggplot(., aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal() +
  theme(strip.text = element_text(size = 18))
```

```{r}
enrichment_results <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  filter(sample %in% sample_metadata$sample) %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample") %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sledwolf$trait) %>%
  filter(dog_type != "domestic") %>%
  group_by(trait) %>%
  summarise(
    median_sled = median(value[dog_type == "sled_dog"]),
    median_wolves = median(value[dog_type == "wolf"]),
    log2FC = log2((median_sled + 1) /
                    (median_wolves + 1)),
    difference = median_sled - median_wolves
  )

enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  nrow()

enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  select(log2FC, Function) %>%
  tt()
```

```{r plot3_sled_domest, comment="", message=FALSE, warning=FALSE}

uniqueGIFT_db<- unique(GIFT_db[c(2,3,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

enrichment_code <- enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element))

unique_codes <- unique(enrichment_code$Code_function)

gift_colors <- read_tsv("resources/gift_colors.tsv") %>%
  filter(Code_function %in% unique_codes) %>%
  mutate(legend = str_c(Code_function, " - ", Function))

gift_color_values <- setNames(gift_colors$Color, gift_colors$legend)
```

```{r final_plot_sled_domest, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=12, fig.fullwidth=TRUE}
enrichment_code %>%
  left_join(
    gift_colors %>% select(Code_function, Color, legend),
    by = join_by(Code_function == Code_function)
  ) %>%
  ggplot(aes(x = forcats::fct_reorder(Function, log2FC), y = log2FC, fill = legend)) +
  geom_col() +
  scale_fill_manual(values = gift_color_values) +
  geom_hline(yintercept = 0) +
  coord_flip() +
  geom_text(aes(30, 0.040), label = "Enriched in\n sled dogs", color = "#666666") +
  geom_text(aes(9, -0.05), label = "Enriched in\n wolves ", color = "#666666") +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.position = "right",
    panel.background = element_blank()
  ) +
  xlab("Microbial Functional Capacity") +
  ylab("log2FC")
```

Higher in wolves
```{r}
enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  select(log2FC, Function) %>%
  filter(log2FC < 0) %>%
  tt()
```

Higher in sled dogs
```{r}
enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  select(log2FC, Function) %>%
  filter(log2FC > 0) %>%
  tt()
```

### Community function differences:

```{r commun_wilcox_func_type, comment="", message=FALSE, warning=FALSE}
function_gift <- GIFTs_functions_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample")

uniqueFuncGIFT_db <- unique(GIFT_db[c(3, 4, 5)])

significant_functions <- function_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ dog_type)$p.value) %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05) %>%
  rownames_to_column(., "functions")  %>%
  select(-functions)

selected_data <- function_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% significant_functions$trait)

pairwise_results_func <- selected_data %>%
  group_by(trait) %>%
  dunn_test(value ~ dog_type, p.adjust.method = "BH") %>%
  filter(p.adj < 0.05) %>%
  mutate(comparison = paste(group1, "vs", group2)) %>%
  left_join(., uniqueFuncGIFT_db, by = join_by(trait == Code_function))
```

```{r pairwise_plot_func_type, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
plot_funct <- ggplot(pairwise_results_func, aes(x = comparison, y = fct_rev(Function))) +
  geom_tile(aes(fill = p.adj), color = "white") +
  geom_text(aes(label = p.adj.signif), size = 4, color = "black") +
  scale_fill_gradient(low = "#b30000", high = "#fef0d9" , name = "Adjusted p-value") +
  theme_minimal() +
  labs(title = "Significant Pairwise Comparisons by Trait", x = "Locality Comparison", y = "Trait") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

#ggsave("output/figures/plot_function_dog_type.png", plot = plot_funct, width = 8, height = 8, dpi = 300)

plot_funct
```

```{r barplot_func_dog_type, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=30, fig.fullwidth=TRUE}
function_gift <- GIFTs_functions_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample")

plot_data <- function_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% pairwise_results_func$trait) %>%
  left_join(., uniqueFuncGIFT_db, by = join_by(trait == Code_function))

ggplot(plot_data, aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_data %>%
  group_by(dog_type, Function) %>%
  summarise(mean = mean(value)) %>%
  pivot_wider(names_from = dog_type, values_from = mean)
```
