# Management-associated ecology drives functional restructuring of the gut microbiome in Qimmit

```{r load_data_beta_sled_dogs, comment="", message=FALSE, warning=FALSE}
set.seed(1234)
load("resources/data.Rdata")
load("output/beta.Rdata")
load("output/alpha_all.Rdata")


sample_metadata <- sample_metadata %>% 
  filter(country=="Greenland") %>%
  filter(season != "winter") %>% 
  mutate(
    dog_group = case_when(
      locality == "Daneborg" ~ "military",
      TRUE                   ~ "civilian"
    )
  )

alpha_div <- alpha_div %>% 
  filter(sample %in% sample_metadata$sample)

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata_bacteria %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

## Taxonomy
**Number of bacteria phyla**
```{r phyla_sled, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

```{r taxonomy_G_barplot_sled, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(., genome_metadata, by = join_by(genome == genome)) %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(count > 0) %>%
  ggplot(., aes(x = sample, y = count, fill = phylum, group = phylum)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = phylum_colors) +
  facet_grid(~ locality, scale = "free", space = "free") +
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 8, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_text(size = 7, hjust = 1, angle = 90),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

### Phylum relative abundances

```{r taxonomy_phylum_summary_sled, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, locality) %>%
  summarise(relabun = sum(count))
```

```{r taxonomy_phylum_summary_origin_sled, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T),
    Daneborg_mean = mean(relabun[locality == "Daneborg"] * 100, na.rm = T),
    Daneborg_sd = sd(relabun[locality == "Daneborg"] * 100, na.rm = T),
    Disko_mean = mean(relabun[locality == "Disko"] * 100, na.rm = T),
    Disko_sd = sd(relabun[locality == "Disko"] * 100, na.rm = T),
    Illulissat_mean = mean(relabun[locality == "Illulissat"] * 100, na.rm = T),
    Illulissat_sd = sd(relabun[locality == "Illulissat"] * 100, na.rm = T),
    Ittoqqortoormiit_mean = mean(relabun[locality == "Ittoqqortoormiit"] * 100, na.rm = T),
    Ittoqqortoormiit_sd = sd(relabun[locality == "Ittoqqortoormiit"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 4), "±", round(total_sd, 3)),
    Daneborg = str_c(round(Daneborg_mean, 4), "±", round(Daneborg_sd, 3)),
    Disko = str_c(round(Disko_mean, 4), "±", round(Disko_sd, 3)),
    Illulissat = str_c(round(Illulissat_mean, 4), "±", round(Illulissat_sd, 3)),
    Ittoqqortoormiit = str_c(round(Ittoqqortoormiit_mean, 4), "±", round(Ittoqqortoormiit_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(phylum, Total, Daneborg, Disko, Illulissat, Ittoqqortoormiit) %>%
  rename(Phylum = phylum) %>%
  tt()
```

### Upset: shared MAGs
```{r chart1_sled, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
genome_row <- genome_counts_filt %>%
  column_to_rownames(., "genome")

sample_metadata_row <- sample_metadata %>%
  column_to_rownames(., "sample")

all(colnames(genome_row) == sample_metadata$sample)
common_samples <- intersect(colnames(genome_row), sample_metadata$sample)
genome_row_aligned <- genome_row[, common_samples, drop = FALSE] %>%
  rownames_to_column(., "genome")
sample_metadata_aligned <- sample_metadata_row[common_samples, , drop = FALSE]

genome_counts_rel <- genome_row_aligned %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames(., "genome")

genome_counts_rel_pa = 1 * (genome_counts_rel > 0)
table_upset_analysis_cont = t(aggregate(
  t(genome_counts_rel_pa),
  by = list(sample_metadata_aligned$locality),
  FUN = sum
)[, -1])
colnames(table_upset_analysis_cont) = levels(as.factor(sample_metadata_aligned$locality))
table_upset_analysis = (table_upset_analysis_cont > 0) * 1
table_upset_analysis = data.frame(table_upset_analysis)
table_upset_analysis = apply(table_upset_analysis, 2, as.integer)
rownames(table_upset_analysis) <- rownames(genome_counts_rel_pa)
country_colors <- c('#41B6C0', '#E5D388', "#834630", "#76b183")

upset(
  as.data.frame(table_upset_analysis),
  keep.order = T,
  sets = rev(c("Daneborg", "Ittoqqortoormiit", "Illulissat", "Disko")),
  sets.bar.color = rev(country_colors),
  mb.ratio = c(0.55, 0.45),
  order.by = "freq"
)
```

## Alpha diversity

```{r alpha_plots_sled, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
#  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))  %>%
#  mutate(locality=factor(locality,levels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island")))%>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  ggplot(aes(y = value, x = locality, group=locality, color=locality, fill=locality)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_color_manual(name="Locality",
          breaks=c("Daneborg","Ittoqqortoormiit","Illulissat","Disko"),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=c('#41B6C0','#E5D388',"#834630","#76b183"))+
  scale_fill_manual(name="Locality",
          breaks=c("Daneborg","Ittoqqortoormiit","Illulissat","Disko"),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=c('#41B6C0','#E5D388',"#834630","#76b183")) +
  facet_wrap(. ~factor(metric, labels=c("richness" = "Richness", "neutral" = "Neutral", "phylogenetic" = "Phylogenetic")), scales = "free", ncol = 5) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```

```{r alpha_values_season_sled, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Daneborg_mean = mean(value[locality == "Daneborg"], na.rm =T),
    Daneborg_sd = sd(value[locality == "Daneborg"], na.rm =T),
    Disko_mean = mean(value[locality == "Disko"], na.rm = T),
    Disko_sd = sd(value[locality == "Disko"], na.rm =T),
    Illulissat_mean = mean(value[locality == "Illulissat"], na.rm = T),
    Illulissat_sd = sd(value[locality == "Illulissat"], na.rm = T),
    Ittoqqortoormiit_mean = mean(value[locality == "Ittoqqortoormiit"], na.rm = T),
    Ittoqqortoormiit_sd = sd(value[locality == "Ittoqqortoormiit"], na.rm = T)) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    "Daneborg-Sirius" = str_c(round(Daneborg_mean, 2), "±", round(Daneborg_sd, 2)),
    "Disko Island" = str_c(round(Disko_mean, 2), "±", round(Disko_sd, 2)),
    Illulissat = str_c(round(Illulissat_mean, 2), "±", round(Illulissat_sd, 2)),
    Ittoqqortoormiit = str_c(round(Ittoqqortoormiit_mean, 2), "±", round(Ittoqqortoormiit_sd, 2))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(alpha,Total, "Daneborg-Sirius",Ittoqqortoormiit, Illulissat, "Disko Island") %>% 
  rename(Alpha=alpha) %>% 
  tt()
```

```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  kruskal_test(value ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))%>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  wilcox_test(value ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

Non-significant
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))%>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  wilcox_test(value ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj > 0.05)
```


## Beta diversity

### Richness

**PERMANOVA**
```{r permanova_greenland_q0_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]
beta_q0n_dis <- as.dist(beta_q0n_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q0n_dis)) %>%
  arrange(match(sample, labels(beta_q0n_dis))) %>%
  column_to_rownames(., "sample")
            
all(rownames(sample_metadata_ordered) == labels(beta_q0n_dis))

betadisper(beta_q0n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n_dis ~ locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$locality, perm = 999)
```

**NMDS**
```{r beta_div_nmds_neutral_plot_G_sled, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
beta_q0n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = locality)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +labs(color='Locality')
```

### Neutral

**PERMANOVA**

```{r permanova_greenland_q1_sled, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)

beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]

beta_q1n_dis <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q1n_dis)) %>%
  arrange(match(sample, labels(beta_q1n_dis))) %>%
  column_to_rownames(., "sample")
            
all(rownames(sample_metadata_ordered) == labels(beta_q1n_dis))

betadisper(beta_q1n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n_dis ~ locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$locality, perm = 999)
```

**NMDS**
```{r beta_div_neutral_G_sled, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=12, fig.fullwidth=TRUE}
beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = locality)) +
    geom_point(size = 3) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    scale_color_manual(name="Locality",
          breaks=c("Daneborg","Ittoqqortoormiit","Illulissat","Disko"),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=c('#41B6C0','#E5D388',"#834630","#76b183"))+
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Locality')
```

***Centroid distance based on the NMDS***
```{r}
betadisper_obj <- betadisper(beta_q1n_dis, sample_metadata_ordered$locality)
dist(betadisper_obj$centroids)
```

**PCoA**
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100
```

```{r pcoa_neutral, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
centroids <- pcoa_df %>%
  group_by(locality) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )

ggplot(pcoa_df, aes(Axis.1, Axis.2, color = locality)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(data = centroids,
             aes(Axis.1, Axis.2, fill = locality),
             color = "black",
             shape = 21,
             size = 6,
             show.legend = FALSE) +
    scale_color_manual(name="Locality",
          breaks=c("Daneborg","Ittoqqortoormiit","Illulissat","Disko"),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=c('#41B6C0','#E5D388',"#834630","#76b183"))+
  scale_fill_manual(name="Locality",
          breaks=c("Daneborg","Ittoqqortoormiit","Illulissat","Disko"),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=c('#41B6C0','#E5D388',"#834630","#76b183")) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Dog type"
  ) +
  theme_classic(base_size = 14)
```

***Centroid distances based on the PCoA***
```{r}
centroid_coords <- centroids %>%
  select(locality, Axis.1, Axis.2)

dist_mat <- dist(centroid_coords[, c("Axis.1", "Axis.2")]) %>% as.matrix()

dist_table <- as.data.frame(as.table(as.matrix(dist_mat))) %>%
  mutate(
    group1 = centroids$locality[Var1],
    group2 = centroids$locality[Var2],
    distance = Freq
  ) %>% 
  select(-Var1, -Var2, -Freq)
```

```{r centroid_distances_locality, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
ggplot(dist_table, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_blank()) +
  labs(fill="Centroids distance")
```

### Phylogenetic

**PERMANOVA**
```{r permanova_greenland_qP_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q1p_mat <- as.matrix(beta_q1p$S)

beta_q1p_filtered <- beta_q1n_mat[rownames(beta_q1p_mat) %in% sample_metadata$sample,
                        colnames(beta_q1p_mat) %in% sample_metadata$sample]

beta_q1p_dis <- as.dist(beta_q1p_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q1p_dis)) %>%
  arrange(match(sample, labels(beta_q1n_dis))) %>%
  column_to_rownames(., "sample")

betadisper(beta_q1p_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1p_dis ~ locality,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

pairwise.adonis(beta_q1p_dis, sample_metadata_ordered$locality, perm = 999)
```

**NMDS**
```{r beta_div_nmds_phylo1_plot_G, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
beta_q1p_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  mutate(x_cen = median(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = median(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = locality)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
  scale_color_manual(
    name = "Locality",
    breaks = c("Daneborg", "Ittoqqortoormiit", "Illulissat", "Disko"),
    values = c('#41B6C0', '#E5D388', "#834630", "#76b183")
  ) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(color='Locality')
```

### Is there any location closer to domestic dogs?
```{r load_data_beta_dan, comment="", message=FALSE, warning=FALSE}
set.seed(1234)
load("resources/data.Rdata")
load("output/beta.Rdata")

remove_sled <- sample_metadata %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_class = case_when(
      breed == "Wolf" ~ "wolf",
      locality == "Disko" ~ "Disko",
      locality == "Daneborg" ~ "Daneborg",
      locality == "Illulissat" ~ "Illulissat",
      locality == "Ittoqqortoormiit" ~ "Ittoqqortoormiit",
      TRUE ~ "Companion"
    )
  ) %>%
  mutate(dog_group = case_when(locality == "Daneborg" ~ "military", TRUE ~ "civilian"))

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata_bacteria %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

**Neutral**
```{r permanova_greenland_q1_dan, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)

beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]

beta_q1n_distance <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1n_distance))) %>% 
  column_to_rownames(., "sample")
            
all(rownames(sample_metadata_ordered) == labels(beta_q1n_distance))

betadisper(beta_q1n_distance, sample_metadata_ordered$dog_class) %>% permutest(., pairwise = TRUE) 
anova(betadisper_obj)

adonis2(beta_q1n_distance ~ dog_class, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

pairwise.adonis(beta_q1n_distance, sample_metadata_ordered$dog_class, perm = 999)
```

```{r}
betadisper_obj <- betadisper(beta_q1n_distance, sample_metadata_ordered$dog_class)
dist(betadisper_obj$centroids)
```

```{r beta_div_neutral_dan, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=14, fig.fullwidth=TRUE}
beta_q1n_distance %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_class) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_class)) +
    geom_point(size = 3) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    scale_color_manual(name="Dog group",
          breaks=c("Companion","Daneborg","Ittoqqortoormiit","Illulissat","Disko", "wolf"),
          labels=c("Companion","Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island", "Wolf"),
          values=c('#41B6C0',"#f5c564","#c49c4b","#9c7c3b", "#614e27","#bb6749"))+
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    )
```


## Differential abundance
```{r phylo_sled, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample) %>%
  filter(country == "Greenland")

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

phylo_samples <- sample_metadata %>%
  filter(country=="Greenland") %>% 
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>%
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_taxonomy <- genome_metadata %>%
  filter(genome %in% rownames(phylo_genome)) %>%
  mutate(genome2 = genome) %>% 
  column_to_rownames("genome2") %>%
  dplyr::select(domain, phylum, class, order, family, genus, species, genome) %>% 
  as.matrix() %>%
  tax_table()

physeq_genome_green <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
```

### Differences across localities

#### MAG level
```{r load_all_sled, comment="", message=FALSE, warning=FALSE}
load("output/ancombc_sled_greenland.Rdata")
region_pairs_sled <- combn(unique(sample_metadata$locality), 2, simplify = FALSE)
```

```{r pair_all_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_sled_mag <- lapply(region_pairs_sled, function(pair) {
  subset_idx <- sample_metadata$locality %in% pair
  
  sled <-  ancombc2(
    data = prune_samples(subset_idx, physeq_genome_green),
    fix_formula = "locality",
    group = "locality",
    p_adj_method = "BH",
    pseudo_sens = TRUE,
    prv_cut = 0.05,
    lib_cut = 1000,
    s0_perc = 0.05,
    struc_zero = TRUE,
    neg_lb = FALSE,
    alpha = 0.05,
    n_cl = 2,
    verbose = TRUE,
    global = FALSE,
    pairwise = FALSE,
    dunnet = FALSE,
    trend = FALSE,
    iter_control = list(
      tol = 1e-5,
      max_iter = 50,
      verbose = FALSE
    ),
    em_control = list(tol = 1e-5, max_iter = 100),
    mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
    trend_control = NULL
  )
  sled_table <- sled$res
  
  sled_table$comparison <- paste(pair, collapse = "_vs_")
  
  return(sled_table)
})

names(results_sled_mag) <- sapply(region_pairs_sled, paste, collapse = "_vs_")

save(results_sled_mag, file = "output/ancombc_sled_greenland.Rdata")
```

```{r extract_sled, comment="", message=FALSE, warning=FALSE}
extract_pair <- function(res, pair) {

  tab <- res
  if (is.null(tab) || nrow(tab) == 0) return(NULL)

  # Identify country coefficient column
  coef_cols <- grep("^lfc_locality", colnames(tab), value = TRUE)
  coef_cols <- coef_cols[!grepl("\\(Intercept\\)", coef_cols)]
  if (length(coef_cols) == 0) return(NULL)

  coef <- coef_cols[1]  # only one since pairwise subset
  country_nonref <- sub("^lfc_locality", "", coef)
  country_ref <- setdiff(pair, country_nonref)

  tibble(
    taxon       = tab$taxon,
    model_term  = coef,
    non_reference = country_nonref,
    reference     = country_ref,
    logFC       = tab[[coef]],
    q_value     = tab[[sub("^lfc_", "q_", coef)]],
    diff_abn    = tab[[sub("^lfc_", "diff_", coef)]],
    enriched_in = ifelse(tab[[coef]] > 0, country_nonref, country_ref),
    depleted_in = ifelse(tab[[coef]] > 0, country_ref, country_nonref),
    comparison  = paste(pair, collapse = "_vs_")
  )
}
pairwise_tables_greenland <- map2(
  results_sled_mag,
  region_pairs_sled,
  extract_pair
)
pairwise_tables_greenland <- discard(pairwise_tables_greenland, is.null)

final_table_tables_greenland <- bind_rows(pairwise_tables_greenland)
nrow(final_table_tables_greenland)
table(final_table_tables_greenland$comparison)

sig_taxa_tables_greenland <- final_table_tables_greenland %>%
  filter(q_value < 0.05) %>% 
  left_join(genome_metadata %>% select(genome, phylum), by=join_by("taxon"=="genome"))
```

```{r loop_plots_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
taxonomy <- data.frame(physeq_genome_green@tax_table) %>%
  distinct(phylum) %>% 
  rownames_to_column(., "taxon")

unique_contrasts <- unique(sig_taxa_tables_greenland$comparison)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (c in unique_contrasts) {

  subset_data <- sig_taxa_tables_greenland %>%
    filter(comparison == c) %>%
    arrange(enriched_in, logFC) %>%
    mutate(taxon_ordered = factor(taxon, levels = unique(taxon)))
  
  non_ref <- unique(subset_data$non_reference)
  ref <- unique(subset_data$reference)
  title <- paste0("Enriched in ", ref, " (− LFC) vs  ", "Enriched in ", non_ref, " (+ LFC)")

  tax_color <- as.data.frame(unique(subset_data$phylum)) %>%
    rename(phylum = 1) %>%
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()
  
  p <- ggplot(subset_data, aes(x = taxon_ordered, y = logFC, fill = phylum)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 6),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    scale_fill_manual(values = tax_color) +
    labs(
      title = title,
      x = "MAGs",
      y = "Log-Fold Change",
      fill = "Phylum"
    )
  
  ggsave(paste0("output/figures/Fig_", gsub(" ", "_", c), ".png"), plot = p, width = 10, height = 25, dpi = 300)
}
```

**Daneborg-Sirius vs Disko Island**

***Enriched in Daneborg-Sirius***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC < 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC < 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

***Enriched in Disko Island***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC > 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC > 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

**Daneborg-Sirius vs Illulissat**

***Enriched in Daneborg-Sirius***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC < 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC < 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

***Enriched in Illulissat***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC > 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC > 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

**Daneborg-Sirius vs Ittoqqortoormiit**

***Enriched in Daneborg-Sirius***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC < 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC < 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

***Enriched in Ittoqqortoormiit***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC > 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC > 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

### Differences between military and civilian-owned dogs

#### MAG level
```{r ancom_phylum_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
ancom_rand_output = ancombc2(
  data = physeq_genome_green,
  assay_name = "counts",
  fix_formula = "dog_group",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "dog_group",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 50,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
  trend_control = NULL
)

save(ancom_rand_output, 
     file = "output/ancombc_daneborg_others.Rdata")
```

```{r ancom_rand_sled, comment="", message=FALSE, warning=FALSE}
load("output/ancombc_daneborg_others.Rdata")
taxonomy <- data.frame(physeq_genome_green@tax_table) %>%
  rownames_to_column(., "taxon")

ancombc_rand_table_mag <- ancom_rand_output$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  dplyr::arrange(q_dog_groupmilitary) %>%
  merge(., taxonomy, by="taxon") %>%
  dplyr::arrange(lfc_dog_groupmilitary)
  
colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>%
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```

```{r ancombc_rand_plot_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=8, fig.fullwidth=TRUE}
ancombc_rand_table_mag %>%
  mutate(genome=factor(taxon,levels=ancombc_rand_table_mag$taxon)) %>%
  ggplot(., aes(x=lfc_dog_groupmilitary, y=forcats::fct_reorder(genome,lfc_dog_groupmilitary), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  geom_text(aes(3, 445), label = "Enriched in\n military dogs", color="#666666") +
  geom_text(aes(-5, 145), label = "Enriched in\n civilian-owned\n dogs", color="#666666") +
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Genome")+
  guides(fill=guide_legend(title="Phylum"))
```

```{r dif_sled_others_genus, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
genus_matrix <- ancombc_rand_table_mag%>%
      mutate(taxon=factor(taxon,levels=ancombc_rand_table_mag$taxon)) %>%
      filter(!is.na(phylum))

genus_matrix <- genus_matrix %>%
  arrange(desc(-lfc_dog_groupmilitary)) %>%
  slice_head(n = 30) %>%
  bind_rows(genus_matrix %>%
              arrange(-lfc_dog_groupmilitary) %>%
              slice_head(n = 30)) 

tax_table <- as.data.frame(unique(genus_matrix$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by = "phylum") %>%
  dplyr::arrange(phylum) %>%
  dplyr::select(colors) %>%
  pull()

genus_matrix %>%
  ggplot(., aes(x=lfc_dog_groupmilitary, y=forcats::fct_reorder(genome,lfc_dog_groupmilitary), fill=phylum)) + 
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  geom_text(aes(4.3, 35), label = "Enriched in\n military dogs", color="#666666") +
  geom_text(aes(-5.6, 20), label = "Enriched in\n civilian-owned\n dogs", color="#666666") +
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Genome") +
  guides(fill=guide_legend(title="Phylum"))      
```

***Military dogs***
```{r}
ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  arrange(desc(species)) %>% 
  select(species) %>% 
  nrow %>% 
  cat()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  group_by(phylum) %>% 
  summarise(count=n()) %>% 
  arrange(-count) %>% 
  tt()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count)%>% 
  tt()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  group_by(phylum, family, genus, species) %>% 
  summarise(count=n()) %>% 
  arrange(-count) %>% 
  tt()
```

***Civilian-owned dogs***
```{r}
ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  arrange(desc(species)) %>% 
  select(species) %>% 
  nrow()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  group_by(phylum) %>% 
  summarise(count=n()) %>% 
  arrange(-count)%>% 
  tt()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count)%>% 
  tt()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  group_by(phylum, family, genus, species) %>% 
  summarise(count=n()) %>% 
  arrange(-count)%>% 
  tt()
```

#### Species level
```{r}
genome_tax <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order), if_else(
      str_detect(class, "Unclassified"),
      class,
      paste0(class, "_Unclassified")
    ), order),
    family = if_else(is.na(family), if_else(
      str_detect(order, "Unclassified"),
      order,
      paste0(order, "_Unclassified")
    ), family),
    genus  = if_else(genus == "", if_else(
      str_detect(family, "Unclassified"),
      family,
      paste0(family, "_Unclassified")
    ), genus),
    species  = if_else(species == "", if_else(
      str_detect(genus, "Unclassified"),
      genus,
      paste0(genus, "_Unclassified")
    ), species)
  )
```

```{r}
phylo_samples <- sample_metadata %>%
  filter(country=="Greenland") %>% 
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  left_join(genome_tax[c(1,3,4,6,8)], by = "genome") %>% 
  select(-genome) %>%
  group_by(phylum, class, family, species) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class, family))%>%
  column_to_rownames("species") %>%
  otu_table(., taxa_are_rows = TRUE)

physeq_genome_sp <- phyloseq(phylo_genome, phylo_samples)

load("output/ancombc_danenborg_greenland_sp.Rdata")
```

```{r ancom_sp_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
ancom_rand_output_sp = ancombc2(
  data = physeq_genome_sp,
  assay_name = "counts",
  fix_formula = "dog_group",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "dog_group",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 50,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
  trend_control = NULL
)

save(ancom_rand_output_sp, 
     file = "output/ancombc_danenborg_greenland_sp.Rdata")
```

```{r}
ancom_rand_output_sp$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(q_dog_groupmilitary < 0.05) %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  nrow()

ancom_rand_output_sp$res %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  arrange(-lfc_dog_groupmilitary)
```

```{r}
ancom_rand_output_sp$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(lfc_dog_groupmilitary < 0.05) %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  nrow()

ancom_rand_output_sp$res %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  arrange(lfc_dog_groupmilitary)
```
```{r dif_sled_others_sp, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=9, fig.fullwidth=TRUE}
ancombc_rand_table_sp <- ancom_rand_output_sp$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  dplyr::arrange(q_dog_groupmilitary) %>%
  mutate(taxon = str_remove(taxon, "_Unclassified")) %>%
  left_join(., taxonomy[c(3,8)] %>% distinct(species, .keep_all = TRUE), by=join_by("taxon"=="species")) %>%
  left_join(., taxonomy[c(3,7)] %>% distinct(genus, .keep_all = TRUE), by=join_by("taxon"=="genus")) %>%
  left_join(., taxonomy[c(3,6)] %>% distinct(family, .keep_all = TRUE), by=join_by("taxon"=="family")) %>%
  mutate(phylum = coalesce(phylum.x, phylum.y, phylum)) %>%
  select(-phylum.x, -phylum.y)%>%
  dplyr::arrange(lfc_dog_groupmilitary)

genus_matrix <- ancombc_rand_table_sp %>%
      mutate(taxon=factor(taxon,levels=ancombc_rand_table_sp$taxon)) %>%
      filter(!is.na(phylum))

genus_matrix <- genus_matrix %>%
  arrange(desc(-lfc_dog_groupmilitary)) %>%
  slice_head(n = 30) %>%
  bind_rows(genus_matrix %>%
              arrange(-lfc_dog_groupmilitary) %>%
              slice_head(n = 30)) 

tax_table <- as.data.frame(unique(genus_matrix$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by = "phylum") %>%
  dplyr::arrange(phylum) %>%
  dplyr::select(colors) %>%
  pull()

genus_matrix %>%
  ggplot(., aes(x=lfc_dog_groupmilitary, y=forcats::fct_reorder(taxon,lfc_dog_groupmilitary), fill=phylum)) + 
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  geom_text(aes(4.3, 45), label = "Enriched in\n military dogs", color="#666666") +
  geom_text(aes(-4.6, 25), label = "Enriched in\n civilian-owned\n dogs", color="#666666") +
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))      
```

## Functional differences

```{r gift_analysis_func_sled, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts),]
genome_counts_filt <- genome_counts_filt %>%
  select(genome, all_of(sample_metadata$sample)) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))%>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome,]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]

GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_row <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>% 
  column_to_rownames(., "genome") 

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)
```

### Different localities in Greenland

#### MCI

```{r}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  mutate(MCI = str_c(round(MCI, 3), "±", round(sd, 3))) %>% 
  select(-sd) %>% 
  tt()
```

```{r}
MCI_pf <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(residuals(lm(value ~ locality, data = MCI_pf))) 
bartlett.test(value ~ locality, data = MCI_pf)  

kruskal.test(value ~ locality, data = MCI_pf)
pw <- pairwise.wilcox.test(
  x = MCI_pf$value,
  g = MCI_pf$locality,
  p.adjust.method = "BH"
)

wilcox_tbl <- pw$p.value %>%
  as.data.frame() %>%
  rownames_to_column("group1") %>%
  pivot_longer(
    -group1,
    names_to = "group2",
    values_to = "p_adjust"
  ) %>%
  filter(!is.na(p_adjust)) %>%
  mutate(
    comparison = paste(group1, "vs", group2),
    p_adjust = round(p_adjust, 3),
    sig = case_when(
      p_adjust < 0.001 ~ "***",
      p_adjust < 0.01  ~ "**",
      p_adjust < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(comparison, p_adjust, sig)

wilcox_tbl %>%
  rename("Compared countries" = comparison,
         "Adjusted p-value" = p_adjust) %>%
  tt()
```

#### Functional pathways

```{r commun_wilcox_elem_sled, comment="", message=FALSE, warning=FALSE}
element_long <- GIFTs_elements_community %>% 
  as.data.frame() %>%
  rownames_to_column(., "sample") %>% 
  filter(sample %in% sample_metadata$sample) %>% 
  left_join(sample_metadata %>% select(sample,locality), by="sample") %>%
  pivot_longer(-c(sample,locality), names_to = "trait", values_to = "value")

uniqueGIFT_db <- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_traits <- element_long %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ locality)$p.value) %>%
  mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
  filter(p_adjust < 0.05) %>%
  pull(trait)

effect_sizes <- element_long %>%
  group_by(trait) %>%
  kruskal_effsize(value ~ locality)

kw_results <- element_long %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ locality)$p.value) %>%
  mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
  filter(p_adjust < 0.05) %>%
  left_join(effect_sizes, by = "trait")

pairwise_results <- element_long %>%
  filter(trait %in% significant_traits) %>%
  group_by(trait) %>%
  dunn_test(value ~ locality, p.adjust.method = "BH") %>%
  filter(p.adj<0.05) %>%
  mutate(comparison = paste(group1, "vs", group2)) %>%
  left_join(uniqueGIFT_db, by = join_by(trait == Code_element))
```

```{r heatmap_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
trait_summary <- element_long %>%
  group_by(trait, locality) %>%
  summarise(median_value = median(value, na.rm = TRUE), .groups = "drop")
trait_summary_sig <- trait_summary %>%
  filter(trait %in% significant_traits)

trait_order <- kw_results %>%
  arrange(desc(effsize)) %>%
  pull(trait)

trait_summary_sig$trait <- factor(trait_summary_sig$trait, levels = trait_order)

trait_summary_sig <- trait_summary_sig %>%
  group_by(trait) %>%
  mutate(z_value = scale(median_value)[,1]) %>%
  ungroup() %>% 
  left_join(uniqueGIFT_db, by = join_by(trait == Code_element))

ggplot(trait_summary_sig,
       aes(locality, Function, fill = z_value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    name = "Relative capacity"
  ) +
  theme_minimal()
```
```{r heatmap2_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=15, fig.fullwidth=TRUE}
heatmap_matrix <- element_long %>%
  filter(trait %in% significant_traits) %>%
  group_by(trait, locality) %>%
  summarise(median_value = median(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = locality, values_from = median_value) %>% 
  left_join(uniqueGIFT_db %>% select(Code_element,Function), by = join_by(trait == Code_element)) %>% 
  column_to_rownames(., "Function") %>% 
  select(-trait) %>% 
  as.matrix()

heatmap_scaled <- t(scale(t(heatmap_matrix)))

sig_annotation <- kw_results %>% 
  left_join(uniqueGIFT_db %>% select(Code_element,Function), by = join_by(trait == Code_element)) %>% 
  select(-trait) %>% 
  filter(Function %in% rownames(heatmap_matrix)) %>%
  select(Function, effsize)%>%
  column_to_rownames("Function") %>%
  as.data.frame()

pheatmap::pheatmap(
  heatmap_scaled,
  annotation_row = sig_annotation,
  clustering_method = "complete",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  scale = "none",
  fontsize_row = 12,
  fontsize_col = 12,
  border_color = NA
)
```

```{r function_plot_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
ggplot(pairwise_results, aes(x = comparison, y = fct_rev(Function))) +
  geom_tile(aes(fill = p.adj), color = "white") +
  geom_text(aes(label = p.adj.signif), size = 4, color = "black") +
  scale_fill_gradient(low = "#b30000", high = "#fef0d9" , name = "Adjusted p-value") +
  theme_minimal() +
  labs(
    title = "Significant Pairwise Comparisons by Trait",
    x = "Locality Comparison",
    y = "Trait"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

```{r barplot_func_sled_locality, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>%
  rownames_to_column(., "sample") %>% 
  filter(sample %in% sample_metadata$sample) %>% 
  left_join(sample_metadata %>% select(sample,locality), by="sample")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, locality), names_to = "trait", values_to = "value") %>%
  filter(trait %in% pairwise_results$trait)%>% 
  left_join(., uniqueGIFT_db, by=join_by(trait==Code_element)) 

ggplot(plot_data, aes(x = locality, y = value, fill = locality)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~ Function, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 16) +
  theme(strip.text = element_text(size = 25, lineheight = 0.6))+
  labs(
    title = "Trait Distributions Across Localities",
    x = "Locality",
    y = "Trait Value" )
```

### Military vs civilian-owned dogs
```{r}
sample_metadata <- sample_metadata %>% 
  mutate(
    dog_group = case_when(
      locality == "Daneborg" ~ "military",
      TRUE                   ~ "civilian"
    )
  )
```

#### MCI

```{r gitfs_functional_group,results=TRUE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_group) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  tt()
```

```{r}
MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(residuals(lm(value ~ dog_group, data = MCI)))
bartlett.test(value ~ dog_group, data = MCI)

wilcox.test(value ~ dog_group, data = MCI)
```

#### Functional pathways

```{r}
enrichment_results <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  filter(sample %in% sample_metadata$sample) %>%
  left_join(sample_metadata %>% select(sample,dog_group), by="sample")%>%
  pivot_longer(-c(sample,dog_group),
               names_to="trait",
               values_to="value") %>%
  group_by(trait) %>%
  summarise(
    p_value = wilcox.test(value ~ dog_group)$p.value,
    median_military = median(value[dog_group=="military"]),
    median_civilian = median(value[dog_group=="civilian"]),
    log2FC = log2((median_military + 1) /
                  (median_civilian + 1)),
    difference = median_military - median_civilian) %>%
  mutate(p_adjust = p.adjust(p_value,"BH")) %>% 
  filter(p_adjust<0.05) 
```

```{r plot3_sled, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
uniqueGIFT_db<- unique(GIFT_db[c(2,3,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

enrichment_code <- enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by=join_by(trait==Code_element))

unique_codes<-unique(enrichment_code$Code_function)

gift_colors <- read_tsv("resources/gift_colors.tsv") %>% 
  filter(Code_function %in% unique_codes)%>% 
  mutate(legend=str_c(Code_function," - ",Function))
gift_color_values <- setNames(gift_colors$Color, gift_colors$legend)

```

```{r final_plot_season, comment="", message=FALSE, warning=FALSE, fig.height=12, fig.width=12, fig.fullwidth=TRUE}
enrichment_code %>%
  left_join(
    gift_colors %>% select(Code_function, Color, legend),
    by = join_by(Code_function == Code_function)
  ) %>%
  ggplot(aes(x = forcats::fct_reorder(Function, log2FC), y = log2FC, fill = legend)) +
  geom_col() +
  scale_fill_manual(values = gift_color_values) +
  geom_hline(yintercept = 0) +
  coord_flip() +
  geom_text(aes(85, 0.060), label = "Enriched in\n Sirius dogs", color="#666666") +
  geom_text(aes(31, -0.050), label = "Enriched in\n civilian dogs", color="#666666") +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        panel.background = element_blank())+
  xlab("Microbial Functional Capacity") + 
  ylab("log2FC")
```

**Effect size: log2FC**

***Military***
```{r}
enrichment_code %>% 
  left_join(unique(GIFT_db[c(2,4,5,6)]), by=join_by("trait"=="Code_element")) %>% 
  filter(log2FC>0) %>% 
  group_by(Function.y) %>% 
  summarise(count=n()) %>% 
  arrange(-count)
```
```{r}
enrichment_code %>% 
  filter(log2FC>0) %>% 
    arrange(Function) %>% 
  select(Function) %>% 
  pull()
```

***Civilian***
```{r}
enrichment_code %>% 
  left_join(unique(GIFT_db[c(2,4,5,6)]), by=join_by("trait"=="Code_element")) %>% 
  filter(log2FC<0) %>% 
  group_by(Function.y) %>% 
  summarise(count=n())%>% 
  arrange(-count)
```

```{r}
enrichment_code %>% 
  filter(log2FC<0) %>% 
  arrange(Function) %>% 
  select(Function) %>% 
  pull()
```

```{r barplot_func_sled, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>%
  rownames_to_column(., "sample") %>% 
  filter(sample %in% sample_metadata$sample) %>% 
  left_join(sample_metadata %>% select(sample,dog_group), by="sample")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_group), names_to = "trait", values_to = "value") %>%
  filter(trait %in% enrichment_code$trait)%>% 
  left_join(., uniqueGIFT_db, by=join_by(trait==Code_element)) 

ggplot(plot_data, aes(x = dog_group, y = value, fill = dog_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~ Function, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 16) +
  labs(
    title = "Trait Distributions Across Localities",
    x = "Locality",
    y = "Trait Value" )
```