# Lower virulence and antimicrobial resistance of Qimmit microbiomes

```{r load_data_amr, comment="", message=FALSE, warning=FALSE}
load("resources/data.Rdata")
load("output/alpha_all.RData")

type_colors <- c('#41B6C0','#E5D388',"#bb6749")
new_labels <- c(
    "companion" = "Companion dogs",
    "sled_dog" = "Sled dogs",
    "wolf" = "Wolves"
    )

management_colors <- c('#41B6C0','#E5D388',"#f5c564","#bb6749")
management_labels <- c(
    "companion" = "Companion dogs",
    "civilian" = "Civilian dogs",
    "military" = "Sirius dogs",
    "wolf" = "Wolves"
    )

locality_colors <- c("#f5c564","#c49c4b","#9c7c3b", "#614e27")
locality_labels <- c(
    "Daneborg" = "Sirius dogs",
    "Disko" = "Disko Island",
    "Illulissat" = "Illulissat",
    "Ittoqqortoormiit" = "Ittoqqortoormiit"
    )

remove_sled <- read_csv("resources/sample_metadata.csv", show_col_types = FALSE) %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_group = case_when(
      dog_type == "wolf" ~ "wolf",
      locality == "Disko" ~ "civilian",
      locality == "Daneborg" ~ "military",
      locality == "Illulissat" ~ "civilian",
      locality == "Ittoqqortoormiit" ~ "civilian",
      TRUE                   ~ "companion"
    )
  )

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")
genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

```{r load_genome_anno, warning=FALSE, comment="", message=FALSE}
genome_annotations <- read_tsv(
  "resources/gene_annotations.tsv.xz",
  col_types = cols(.default = col_character(), start = col_double(), end = col_double())
) %>%
  mutate(genome = str_extract(gene, "^[^^]+")) %>%
  mutate(ec = if_else(is.na(ec), NA_character_, str_c("[EC:", ec, "]"))) %>% 
  select(genome, everything())
```

## Genome length

```{r size_per_sample, warning=FALSE, comment="", message=FALSE}
sample_genome_size <- genome_counts_filt %>% 
  mutate(across(-genome,~ .x / sum(.x, na.rm = TRUE))) %>% 
  inner_join(genome_metadata %>% select(genome,length), by = "genome") %>%
  pivot_longer(
    cols = -c(genome, length),
    names_to  = "sample",
    values_to = "rel_abundance"
  ) %>%
  group_by(sample) %>%
  summarise(
    size = weighted.mean(length, w = rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(sample_metadata,by="sample")
```

### Canids
```{r}
sample_genome_size %>% 
  group_by(dog_type) %>%
  summarise(
    mean = mean(size, na.rm = T)/1000000,
    sd = sd(size, na.rm = T)/1000000) %>% 
  arrange(-mean) %>% 
  mutate(length=str_c(round(mean,3),"±",round(sd,3))) %>% 
  select(dog_type,length) %>% 
  rename(Canid=1, "Genome length"=2) %>% 
  tt()
```

```{r}
sample_genome_size %>%
  kruskal_test(size ~ dog_type) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

```{r}
sample_genome_size %>%
  wilcox_test(size ~ dog_type, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)%>% 
  tt()
```

### Management groups
```{r}
sample_genome_size %>% 
  group_by(dog_group) %>%
  summarise(
    mean = mean(size, na.rm = T)/1000000,
    sd = sd(size, na.rm = T)/1000000) %>% 
  arrange(-mean) %>% 
  mutate(genome_size = str_c(round(mean, 3), " ± ", round(sd, 3))) %>%
  select(dog_group, genome_size) %>%
  rename(Group = 1, "Genome length" = 2)
```

```{r}
sample_genome_size %>%
  kruskal_test(size ~ dog_group) %>%
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>%
  tt()
```

```{r}
sample_genome_size %>%
  wilcox_test(size ~ dog_group, p.adjust.method = "BH") %>%
  filter(p.adj < 0.05)%>% 
  tt()
```

```{r}
sample_genome_size %>%
  wilcox_test(size ~ dog_group, p.adjust.method = "BH") %>%
  filter(p.adj > 0.05)%>% 
  tt()
```

```{r size_per_sample_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_genome_size %>%
  ggplot(aes(x = dog_type, y = size, color = dog_type, fill = dog_type)) +
  geom_jitter(width = 0.2, alpha = 0.7, show.legend = FALSE, size = 0.5) +
  geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = type_colors) +
  scale_fill_manual(values = type_colors) +
  scale_x_discrete(labels = new_labels) +
  theme_classic()+
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size =12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Genome length")
```
```{r genome_size_per_doggroup_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_genome_size %>% 
  mutate(dog_group=factor(dog_group,levels=c("companion","civilian","military", "wolf"))) %>%
  ggplot(aes(x = dog_group, y = size, color = dog_group, fill = dog_group)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = management_colors) +
  scale_fill_manual(values = management_colors) +
  scale_x_discrete(labels = management_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Genome length")
```
```{r size_per_sled_loc_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_genome_size %>% 
  filter(dog_type=="sled_dog") %>% 
  ggplot(aes(x=locality,y=size,color=locality,fill=locality)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 7),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 9, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Genome length")
```

```{r}
sample_genome_size %>%
  filter(dog_type=="sled_dog") %>% 
  kruskal_test(size ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
sample_genome_size %>%
  filter(dog_type=="sled_dog") %>% 
  wilcox_test(size ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

## Antimicrobial resistance (AMR)

```{r amr_per_genome, warning=FALSE, comment="", message=FALSE}
genome_amr <- genome_annotations %>% 
  group_by(genome) %>%
  summarise(AMR = sum(resistance_type == "AMR", na.rm = TRUE), .groups = "drop")
```

```{r amr_per_sample, warning=FALSE, comment="", message=FALSE}
sample_amr <- genome_counts_filt %>%
  mutate(across(-genome, ~ .x / sum(.x, na.rm = TRUE))) %>%
  inner_join(genome_amr, by = "genome") %>%
  pivot_longer(cols = -c(genome, AMR), names_to  = "sample", values_to = "rel_abundance") %>%
  group_by(sample) %>%
  summarise(AMR = weighted.mean(AMR, w = rel_abundance, na.rm = TRUE), .groups = "drop") %>%
  left_join(sample_metadata, by = "sample")
```

**Top 20 MAGs with highest values**
```{r}
genome_amr %>%
  left_join(genome_metadata, by = "genome") %>%
  arrange(-AMR) %>%
  head(20) %>%
  tt()
```

### Canids
```{r}
sample_amr %>%
  group_by(dog_type) %>%
  summarise(mean = mean(AMR, na.rm = T), sd = sd(AMR, na.rm = T)) %>%
  arrange(-mean) %>%
  mutate(AMR = str_c(round(mean, 3), " ± ", round(sd, 3))) %>%
  select(dog_type, AMR) %>% 
  rename(Canid=1)%>% 
  tt()
```

```{r amr_per_group_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_amr %>% 
  ggplot(aes(x = dog_type, y = AMR, color = dog_type, fill = dog_type)) +
  geom_jitter(width = 0.2, alpha = 0.7, show.legend = FALSE, size = 0.5) +
  geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = type_colors) +
  scale_fill_manual(values = type_colors) +
  scale_x_discrete(labels = new_labels)+
  theme_classic() +
  theme(
    axis.text = element_text(size = 10), 
    strip.background = element_blank(), 
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
  ) +
  labs(y = "AMR")
```

```{r}
sample_amr %>%
  kruskal_test(AMR ~ dog_type) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

```{r}
sample_amr %>%
  wilcox_test(AMR ~ dog_type, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05) %>% 
  tt()
```

### Management groups
```{r}
sample_amr %>% 
  group_by(dog_group) %>%
  summarise(
    mean = mean(AMR, na.rm = T),
    sd = sd(AMR, na.rm = T)) %>% 
  arrange(-mean) %>% 
  mutate(AMR=str_c(round(mean,3)," ± ",round(sd,3))) %>% 
  select(dog_group, AMR) %>%
  rename(Group = 1)
```
```{r}
sample_amr %>% 
  filter(country=="Greenland") %>%
  wilcox_test(AMR ~ dog_group) %>% 
  tt()
```

```{r}
sample_amr %>%
  kruskal_test(AMR ~ dog_group) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
sample_amr %>%
  wilcox_test(AMR ~ dog_group, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05) %>% 
  tt()
```

```{r amr_per_doggroup_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_amr %>% 
  mutate(dog_group=factor(dog_group,levels=c("companion","civilian","military", "wolf"))) %>%
  ggplot(aes(x=dog_group, y=AMR, color=dog_group, fill=dog_group)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = management_colors) +
  scale_fill_manual(values = management_colors) +
  scale_x_discrete(labels = management_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="AMR")
```

```{r amr_per_sled_loc_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_amr %>% 
  filter(dog_type=="sled_dog") %>% 
  ggplot(aes(x=locality,y=AMR,color=locality,fill=locality)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 7),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 9, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="AMR")
```

```{r}
sample_amr %>%
  filter(dog_type=="sled_dog") %>% 
  kruskal_test(AMR ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
sample_amr %>%
  filter(dog_type=="sled_dog") %>% 
  wilcox_test(AMR ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

Non-sig
```{r}
sample_amr %>%
      filter(dog_type=="sled_dog") %>% 
  wilcox_test(AMR ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj > 0.05)
```

Max AMR value in sled dogs
```{r}
sample_amr %>% 
  filter(country=="Greenland") %>% 
  arrange(-AMR)
```

Max AMR value in Sirius dogs
```{r}
sample_amr %>% 
  filter(locality=="Daneborg") %>% 
  arrange(-AMR)
```


## Virulence factors

```{r vf_per_genome, warning=FALSE, comment="", message=FALSE}
genome_vf <- genome_annotations %>% 
  group_by(genome) %>%
  summarise(vf = sum(!is.na(vf_type), na.rm = TRUE), .groups = "drop")

sample_vf <- genome_counts_filt %>% 
  mutate(across(-genome,~ .x / sum(.x, na.rm = TRUE))) %>% 
  inner_join(genome_vf, by = "genome") %>%
  pivot_longer(cols = -c(genome, vf), names_to  = "sample", values_to = "rel_abundance") %>%
  group_by(sample) %>%
  summarise(
    vf = weighted.mean(vf, w = rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(sample_metadata, by="sample")
```

**Top 20 MAGs with highest values**
```{r}
genome_vf %>% 
  left_join(genome_metadata, by="genome") %>% 
  arrange(-vf) %>% 
  head(20) %>% 
  tt()
```


### Canids
```{r}
sample_vf %>% 
  group_by(dog_type) %>%
  summarise(
    mean = mean(vf, na.rm = T),
    sd = sd(vf, na.rm = T)) %>% 
  arrange(-mean) %>% 
  mutate(virulence=str_c(round(mean,3),"±",round(sd,3))) %>% 
  select(dog_type,virulence) %>%
  rename("Canid" = 1, "Virulence factor" = 2)%>% 
  tt()
```

```{r vf_per_group_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
  ggplot(aes(x=dog_type,y=vf,color=dog_type,fill=dog_type)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = type_colors) +
  scale_fill_manual(values = type_colors) +
  scale_x_discrete(labels = new_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factor")
```

```{r}
sample_vf %>%
  kruskal_test(vf ~ dog_type) %>% 
  tt()
```

### Management: Civilian & Military
```{r}
sample_vf %>% 
  group_by(dog_group) %>%
  summarise(
    mean = mean(vf, na.rm = T),
    sd = sd(vf, na.rm = T)) %>% 
  arrange(-mean) %>% 
  mutate(virulence=str_c(round(mean,3),"±",round(sd,3))) %>% 
  select(dog_group,virulence) %>%
  rename("Canid" = 1, "Virulence factor" = 2)%>% 
  tt()
```
```{r vf_per_military_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
#  filter(country=="Greenland") %>% 
  mutate(dog_group=factor(dog_group,levels=c("companion","civilian","military", "wolf"))) %>%
  ggplot(aes(x=dog_group,y=vf,color=dog_group,fill=dog_group)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = management_colors) +
  scale_fill_manual(values = management_colors) +
  scale_x_discrete(labels = management_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factor")
```
```{r}
sample_vf %>%
  kruskal_test(vf ~ dog_group) %>% 
  adjust_pvalue(method = "BH")
  tt()
```

```{r vf_per_sled_loc_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
  filter(dog_type=="sled_dog") %>%
  ggplot(aes(x=locality,y=vf,color=locality,fill=locality)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factors")
```
```{r}
sample_vf %>%
    filter(country=="Greenland") %>%
  kruskal_test(vf ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  tt()
```

Significant
```{r}
sample_vf %>%
  filter(country=="Greenland") %>%
  wilcox_test(vf ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

Non-sig
```{r}
sample_vf %>%
  filter(country=="Greenland") %>%
  wilcox_test(vf ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj > 0.05)
```

With pack as shape
```{r vf_per_sled_outliers_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
  filter(dog_type=="sled_dog") %>%
 # filter(!str_starts(sample, "GL_02")) %>% 
  mutate(Pack = as.factor(Pack)) %>%
  ggplot(aes(x=locality,y=vf,color=locality,fill=locality)) +
  geom_jitter(aes(shape = Pack),width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 7),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 9, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factors")
```

Max VF value in sled dogs
```{r}
sample_vf %>% 
  filter(country=="Greenland") %>% 
  arrange(-vf) %>% 
  tt()
```

```{r}
sample_vf %>% 
  filter(locality=="Daneborg") %>% 
  arrange(-vf)%>% 
  tt()
```

## Size-function correlations

```{r size_amr_correlation_plot, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>%
  ggplot(aes(x = size, y = AMR, color = country)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, show.legend = FALSE) +
  theme_classic()
```

```{r size_amr_correlation_plot_group, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>%
  ggplot(aes(x = size, y = AMR, color = dog_type)) +
  geom_point() +
  scale_color_manual(values = type_colors, breaks = c("companion", "sled_dog", "wolf"),
      labels = c("Companion dogs", "Sled dogs", "Wolves")) +
  geom_smooth(method = "lm", se = TRUE, show.legend = FALSE) +
  theme_classic()+
  labs(y="AMR", x="Genome size", color = "Canids")
```
```{r size_amr_correlation_plot_together, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>% 
  ggplot(., aes(x = size, y = AMR)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_classic()+
  labs(y="AMR", x="Genome size")
```

```{r}
df <- sample_genome_size %>% 
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>% 
  left_join(alpha_div, by = "sample")

cor.test(df$size, df$AMR, method = "spearman")

```
```{r}

model <- lm(AMR ~ size, data = df)
summary(model)

model <- lm(AMR ~ size * country, data = df)
summary(model)

df %>%
  group_by(country) %>%
  do(tidy(lm(AMR ~ size, data = .))) %>% 
  filter(term=="size")%>% 
  filter(p.value<0.05) %>% 
  tt()

df %>%
  group_by(dog_type) %>%
  do(tidy(lm(AMR ~ size, data = .))) %>% 
  filter(term=="size") %>% 
  tt()

model <- lm(AMR ~ size * dog_group, data = df)
summary(model)
```

```{r}
df <- sample_genome_size %>% 
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>% 
  left_join(alpha_div, by = "sample")

cor.test(df$size, df$vf, method = "spearman")

model <- lm(vf ~ size, data = df)
summary(model)

model <- lm(vf ~ size * country, data = df)
summary(model)

df%>%
  group_by(country) %>%
  do(tidy(lm(vf ~ size, data = .))) %>% 
  filter(term=="size")%>% 
  filter(p.value<0.05)

df%>%
  group_by(dog_type) %>%
  do(tidy(lm(vf ~ size, data = .))) %>% 
  filter(term=="size")

model <- lm(vf ~ size * dog_group, data = df)
summary(model)
```
```{r size_vf_correlation_plot, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>%
  ggplot(aes(x = size, y = vf, color = country)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, show.legend = FALSE) +
  theme_classic()
```

```{r size_vf_correlation_plot_together, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>% 
  ggplot(., aes(x = size, y = vf)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_classic()
```

**AMR and VF correlation colored by genome size**
```{r size_vf_amr_correlation_plot_group, warning=FALSE, comment="", message=FALSE, fig.height=10, fig.width=12, fig.fullwidth=TRUE}
sample_amr %>%
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>%
  left_join(sample_genome_size %>% select(sample, size), by = "sample") %>% 
  ggplot(., aes(x = AMR, y = vf,color = size)) +
  geom_point(size=3) +
  geom_smooth(method = "lm", se = TRUE) +
    scale_color_viridis_c() +
  scale_size_continuous(range = c(2, 7)) +
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), size = 6, show.legend = FALSE) +
  theme_classic()+
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  )+
  labs(color="Genome size")
```
