# Seasonal differences in eastern Greenland

```{r load_data_host_filt_season, comment="", message=FALSE, warning=FALSE}
load("resources/data.Rdata")
load("output/beta_season.Rdata")

sample_metadata <- sample_metadata %>% 
  filter(locality == "Ittoqqortoormiit" & day %in% c("day_1", "unknown"))

genome_counts_filt <- genome_counts_filt %>% 
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>% 
  semi_join(., genome_counts_filt, by = "genome") %>% 
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")

genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

```{r}
sample_metadata %>% 
  group_by(season) %>% 
  summarise(n=n()) %>% 
  tt()
```

## Taxonomy
**Number of bacteria phyla**
```{r phyla_season, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

```{r taxonomy_barplot_seasons, fig.height=8, fig.width=15, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
  geom_bar(stat="identity", colour="white", linewidth=0.1) + 
  scale_fill_manual(values=phylum_colors) +
  facet_grid(~ season, scale="free", space="free")+
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 14, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.y = element_text(size = 12),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

### Phylum relative abundances

```{r taxonomy_phylum_season, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,season) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_phylum_summary_season, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(season, phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T)
  ) %>%
  arrange(-total_mean) %>%
  mutate(total = str_c(round(total_mean, 4), "±", round(total_sd, 3))) %>%
  dplyr::select(phylum, total, season) %>%
  pivot_wider(names_from = season, values_from = total) %>% 
  tt()
```

### Differences in phylum (ancombc)

```{r zero_phylo_season, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()
phylo_genome <- genome_counts_filt %>%
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  filter(genome %in% rownames(phylo_genome)) %>% 
  mutate(genome2 = genome) %>% 
  column_to_rownames("genome2") %>%
  dplyr::select(domain, phylum, class, order, family, genus, species, genome) %>% 
  as.matrix() %>%
  tax_table() 

physeq_genome_itto <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
physeq_genome_itto <- prune_taxa(taxa_sums(physeq_genome_itto)>0, physeq_genome_itto)
```

```{r ancom_rand_phylum_season, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_phylum_itto = ancombc2(
  data = physeq_genome_itto,
  assay_name = "counts",
  tax_level = "phylum",
  fix_formula = "season",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "season",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 20,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = NULL
)
```

```{r}
ancom_rand_output_phylum_itto$res %>%
  dplyr::select(taxon, lfc_seasonwinter, q_seasonwinter) %>%
  filter(q_seasonwinter < 0.05) %>%
  dplyr::arrange(q_seasonwinter) %>% 
  tt()
```

## Alpha diversity
```{r alpha_div_season, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  filter(genome %in% genome_metadata_bacteria$genome) %>% 
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample))
```


```{r alpha_plots_seasons, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = value, x = season, group=season, color=season, fill=season)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(. ~factor(metric, labels=c("richness" = "Richness", "neutral" = "Neutral", "phylogenetic" = "Phylogenetic")), scales = "free", ncol = 5) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size =12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```

## Beta diversity
```{r beta_div_season, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  filter(genome %in% genome_metadata_bacteria$genome) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, tree = genome_tree)

save(beta_q0n, 
     beta_q1n, 
     beta_q1p,
     file = "output/beta_season.Rdata")
```

**Richness**
```{r}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]
beta_q0n <- as.dist(beta_q0n_filtered)
```

```{r permanova_q0_season, comment="", message=FALSE, warning=FALSE}
adonis2(beta_q0n ~ season,
  data = sample_metadata %>% arrange(match(sample, labels(beta_q0n))),
  permutations = 999) %>%
  broom::tidy() %>%
  tt()
```

```{r beta_div_rich_season, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q0n %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = season)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Season')
```

**Neutral**
```{r permanova_q1n_filt_seasons, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)
beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]
beta_q1n_dis <- as.dist(beta_q1n_filtered)
sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1n_dis))) %>% 
  column_to_rownames(., "sample")
```
```{r permanova_q1n_season, comment="", message=FALSE, warning=FALSE}
adonis2(beta_q1n_dis ~ season,
  data = sample_metadata_ordered,
  permutations = 999) %>%
  broom::tidy() %>%
  tt()
```
```{r beta_div_neutral_season, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = season)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Season')
```


```{r permanova_q1p_filt_season, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
#Phylogenetic
beta_q1p_mat <- as.matrix(beta_q1p$S)
beta_q1p_filtered <- beta_q1p_mat[rownames(beta_q1p_mat) %in% sample_metadata$sample,
                        colnames(beta_q1p_mat) %in% sample_metadata$sample]
beta_q1p_dis <- as.dist(beta_q1p_filtered)
sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1p_dis))) %>% 
  column_to_rownames(., "sample")

adonis2(beta_q1p_dis ~ season,
  data = sample_metadata_ordered,
  permutations = 999) %>%
  broom::tidy() %>%
  tt()

beta_q1p_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = season)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Season')
```

## Differences in MAG abundances: Ancombc2

```{r ancom_rand_mag_season, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_mag_itto_all = ancombc2(
  data = physeq_genome_itto,
  assay_name = "counts",
  tax_level = NULL,
  fix_formula = "season",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "season",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 50,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
  trend_control = NULL
)
```

#### Structural zeros

***Total number of structural zeros***
```{r}
structural_zeros <- ancom_rand_output_mag_itto_all$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("summer"=1) %>% 
  rename("winter"=2)

nrow(structural_zero_groups)
```

***Structural zeros in summer***
```{r}
structural_zero_groups %>% 
  filter(summer==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(summer==TRUE) %>% 
  rownames_to_column(., "genome") %>% 
  left_join(., genome_metadata, by="genome") %>% 
  count(phylum, name = "sample_count") %>%
  arrange(desc(sample_count)) %>% 
  tt()
```

***Structural zeros in winter***
```{r}
structural_zero_groups %>% 
  filter(winter==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(winter==TRUE) %>% 
  rownames_to_column(., "genome") %>% 
  left_join(., genome_metadata, by="genome") %>% 
  count(phylum, name = "sample_count") %>%
  arrange(desc(sample_count)) %>% 
  tt()
```

```{r ancom_table_season, comment="", message=FALSE, warning=FALSE}
taxonomy <- data.frame(physeq_genome_itto@tax_table) %>%
  rownames_to_column(., "taxon") %>%
  mutate_at(vars(phylum, order, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))

ancombc_rand_table_mag_itto <- ancom_rand_output_mag_itto_all$res %>%
  dplyr::select(taxon, lfc_seasonwinter, q_seasonwinter) %>%
  filter(q_seasonwinter < 0.05) %>%
  dplyr::arrange(q_seasonwinter) %>%
  left_join(taxonomy, by = join_by(taxon == taxon)) %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_seasonwinter)
  
colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80")) %>% 
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag_itto$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```
```{r ancombc_rand_plot_season, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ancombc_rand_table_mag_itto %>%
  mutate(genome=factor(genome,levels=ancombc_rand_table_mag_itto$genome)) %>%
  ggplot(., aes(x=lfc_seasonwinter, y=forcats::fct_reorder(genome,lfc_seasonwinter), fill=phylum)) +
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  geom_text(aes(3.5, 135), label = "Enriched\nin winter", color="#666666") +
  geom_text(aes(-4, 35), label = "Enriched\nin summer", color="#666666")+
  xlab("log2FoldChange") + 
  ylab("Genome")+
  guides(fill=guide_legend(title="Phylum"))
```

Enriched in summer (top 20)
```{r}
ancombc_rand_table_mag_itto %>% 
  filter(lfc_seasonwinter<0) %>% 
  arrange(lfc_seasonwinter) %>% 
  select(lfc_seasonwinter, genus, species) %>% 
  head(20) %>% 
  tt()
```

Enriched in winter (top 20)
```{r}
ancombc_rand_table_mag_itto %>% 
  filter(lfc_seasonwinter>0) %>% 
  arrange(-lfc_seasonwinter) %>% 
  select(lfc_seasonwinter, genus, species)%>% 
  head(20) %>% 
  tt()
```

## Functional differences
```{r gift_analysis_season, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt %>%
  filter(genome_counts_filt$genome %in% rownames(genome_gifts)) %>% 
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))%>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome,]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]

GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_row <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>% 
  column_to_rownames(., "genome") 

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)
```

## Groups MCI
```{r gitfs_functional_season}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  tt()
```

***Shapiro test***
```{r}
MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample))

shapiro.test(MCI$value)
wilcox.test(value ~ season, data=MCI)
```

### Community elements differences:
```{r comunity_elem_season, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select(sample,season), by=("sample"))
```

```{r commun_wilcox_elem_season, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

element_gift %>%
  pivot_longer(-c(sample, season), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = wilcox.test(value ~ season)$p.value,
            .groups = "drop") %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05) %>% 
  left_join(., uniqueGIFT_db, by=join_by("trait"=="Code_element"))
```