# Virulence factor analysis

```{r load_data_virulence}
load("data/data.Rdata")
virulence_summary <- read_tsv("data/virulence_summary_dogs.tsv") %>%
    rename(genome=1)%>%
    arrange(match(genome,genome_metadata$genome))%>% 
  column_to_rownames(., "genome")
```

# V1 value

```{r V1, comment="", message=FALSE, warning=FALSE}
genome_counts_rel <- genome_counts %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  arrange(match(genome,genome_metadata$genome))%>% 
  semi_join(., virulence_summary %>% rownames_to_column("genome"), by = "genome") %>% 
  column_to_rownames(., "genome")

result_V1 <- data.frame(matrix(ncol = ncol(genome_counts_rel), nrow = nrow(genome_counts_rel)))
colnames(result_V1) <- colnames(genome_counts_rel)
rownames(result_V1) <- rownames(genome_counts_rel)

for (i in 1:nrow(virulence_summary)) {
  result_V1[i, ] <- genome_counts_rel[i, ] * virulence_summary[i, 1]
}
```

## Mean V1 values
```{r V1_mean, comment="", message=FALSE, warning=FALSE}
result_V1 %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  group_by(region) %>%
  summarise(average=mean(V1))
```

## Plots
```{r V1_plot2, comment="", message=FALSE, warning=FALSE}
result_V1 %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = V1, x = region, group=region, color=region, fill=region)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(alpha=0.5) +
      scale_color_manual(name="Region",
          values=c("#e5bd5b", "#6b7398")) +
      scale_fill_manual(name="Region",
          values=c("#e5bd5b50", "#6b739850")) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 12, color="black",face="bold"),
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12))+
    guides(fill = guide_legend(override.aes = list(size=3)))
```

### Mixed models 
```{r V1_values, comment="", message=FALSE, warning=FALSE}
V1_values <- result_V1 %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))
```

```{r model_v1_lm, comment="", message=FALSE, warning=FALSE}
Modelq1_V1<- lm(formula = V1 ~ region, data = V1_values) 
anova(Modelq1_V1)
r.squaredGLMM(Modelq1_V1)
```

# V1 + V2 value

```{r V1_2, comment="", message=FALSE, warning=FALSE}
# virulence_summary <- virulence_summary %>% 
#   mutate(sumrow= V1 + V2) %>% 
#   rename(V1_2=sumrow)%>% 
#   arrange(desc(V1_2))

result_V1_2 <- data.frame(matrix(ncol = ncol(genome_counts_rel), nrow = nrow(genome_counts_rel)))
colnames(result_V1_2) <- colnames(genome_counts_rel)
rownames(result_V1_2) <- rownames(genome_counts_rel)

for (i in 1:nrow(virulence_summary)) {
  result_V1_2[i, ] <- genome_counts_rel[i, ] * virulence_summary[i, 8]
}
```

## Mean V12 values
```{r V1_2_mean, comment="", message=FALSE, warning=FALSE}
result_V1_2 %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1_2=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  group_by(region) %>%
  summarise(average=mean(V1_2))
```

## Plots
```{r V12_plot1, comment="", message=FALSE, warning=FALSE}
result_V1_2 %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1_2=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = V1_2, x = region, group=region, color=region, fill=region)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(alpha=0.5) +
  scale_color_manual(name="Region",
          values=c("#e5bd5b", "#6b7398")) +
  scale_fill_manual(name="Region",
          values=c("#e5bd5b50", "#6b739850")) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 12, color="black",face="bold"),
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12))+
    guides(fill = guide_legend(override.aes = list(size=3)))
```

### Mixed models 
```{r v12_values, comment="", message=FALSE, warning=FALSE}
V1_2_values <- result_V1_2 %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1_2=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))
```

```{r model_v12_lm, comment="", message=FALSE, warning=FALSE}
Modelq1_nvi<- lm(formula = V1_2 ~ region, data = V1_2_values) 
anova(Modelq1_nvi)
```

# AVI value

```{r AVI, comment="", message=FALSE, warning=FALSE}
result_AVI <- data.frame(matrix(ncol = ncol(genome_counts_rel), nrow = nrow(genome_counts_rel)))
colnames(result_AVI) <- colnames(genome_counts_rel)
rownames(result_AVI) <- rownames(genome_counts_rel)

for (i in 1:nrow(virulence_summary)) {
  result_AVI[i, ] <- genome_counts_rel[i, ] * virulence_summary[i, 6]
}
```

## Mean AVI values
```{r AVI_mean, comment="", message=FALSE, warning=FALSE}
result_AVI %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(AVI=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  group_by(region) %>%
  summarise(average=mean(AVI))
```

## Plots
```{r AVI_plot1, comment="", message=FALSE, warning=FALSE}
result_AVI %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(AVI=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = AVI, x = region, group=region, color=region, fill=region)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(alpha=0.5) +
      scale_color_manual(name="Region",
          values=c("#e5bd5b", "#6b7398")) +
      scale_fill_manual(name="Region",
          values=c("#e5bd5b50", "#6b739850"))+
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 12, color="black",face="bold"),
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12))+
    guides(fill = guide_legend(override.aes = list(size=3)))
```


### Mixed models 
```{r AVI_values, comment="", message=FALSE, warning=FALSE}
AVI_values <- result_AVI %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(AVI=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))
```

```{r model_lm_avi, comment="", message=FALSE, warning=FALSE}
Modelq1_avi<- lm(formula = AVI ~ region, data = AVI_values) 
anova(Modelq1_avi)
r.squaredGLMM(Modelq1_avi)
```


# NVI value

```{r NVI, comment="", message=FALSE, warning=FALSE}

result_NVI <- data.frame(matrix(ncol = ncol(genome_counts_rel), nrow = nrow(genome_counts_rel)))
colnames(result_NVI) <- colnames(genome_counts_rel)
rownames(result_NVI) <- rownames(genome_counts_rel)

for (i in 1:nrow(virulence_summary)) {
  result_NVI[i, ] <- genome_counts_rel[i, ] * virulence_summary[i, 7]
}
```

## Mean NVI values
```{r NVI_mean, comment="", message=FALSE, warning=FALSE}
result_NVI %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(NVI=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  group_by(region) %>%
  summarise(average=mean(NVI))
```

## Plots
```{r NVI_plot1, comment="", message=FALSE, warning=FALSE}
result_NVI %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(NVI=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = NVI, x = region, group=region, color=region, fill=region)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(alpha=0.5) +
      scale_color_manual(name="Region",
          values=c("#e5bd5b", "#6b7398")) +
      scale_fill_manual(name="Region",
          values=c("#e5bd5b50", "#6b739850")) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 12, color="black",face="bold"),
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12))+
    guides(fill = guide_legend(override.aes = list(size=3)))
```

### Mixed models 
```{r NVI_values, comment="", message=FALSE, warning=FALSE}
NVI_values <- result_NVI %>%
  colSums() %>% 
  as.data.frame() %>% 
  rename(NVI=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))
```

```{r model_NVI_lm, comment="", message=FALSE, warning=FALSE}
Modelq1_nvi<- lm(formula = NVI ~ region, data = NVI_values) 
anova(Modelq1_nvi)
r.squaredGLMM(Modelq1_nvi)
```






# Some other analysis

```{r}
virulence_summary %>% 
  colMeans()
```

## Clostridium virulence

```{r}
virulence_summary %>% 
  rownames_to_column(., "genome") %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  filter(class=="c__Clostridia") %>% 
  column_to_rownames(., "genome") %>% 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
```
```{r}
virulence_summary %>% 
  rownames_to_column(., "genome") %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  filter(class=="c__Fusobacteriia") %>% 
  column_to_rownames(., "genome") %>% 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
```
```{r}
virulence_summary %>% 
  rownames_to_column(., "genome") %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  filter(class=="c__Fusobacteriia") %>% 
  column_to_rownames(., "genome") %>% 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
```
```{r}
top10_V1 <- virulence_summary %>% 
  arrange(desc(V1)) %>%
  slice(1:10)  
genome_metadata %>% 
  filter(genome %in% rownames(top10_V1)) %>%
  left_join(top10_V1 %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
    arrange(match(genome,rownames(top10_V1))) %>% 
  select(phylum, family, species, V1, V2, AVI,NVI, V1_2)
```

## Top 10

```{r}
top10_V1 <- virulence_summary %>% 
  arrange(desc(V1)) %>%
  slice(1:10)
genome_metadata %>% 
  filter(genome %in% rownames(top10_V1)) %>%
  left_join(top10_V1 %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
    arrange(match(genome,rownames(top10_V1))) %>% 
  select(phylum, family, species, V1, V2, AVI,NVI, V1_2)
```

```{r}
top10_V1_2 <- virulence_summary %>% 
  arrange(desc(V1_2)) %>%
  slice(1:10)  
genome_metadata %>% 
  filter(genome %in% rownames(top10_V1_2)) %>%
  left_join(top10_V1_2 %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
    arrange(match(genome,rownames(top10_V1_2))) %>% 
  select(phylum, family, species, V1, V2, AVI,NVI, V1_2)
```
```{r}
top10_AVI <- virulence_summary %>% 
  arrange(desc(AVI)) %>%
  slice(1:10)  
genome_metadata %>% 
  filter(genome %in% rownames(top10_AVI)) %>%
  left_join(top10_AVI %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
    arrange(match(genome,rownames(top10_AVI))) %>% 
  select(phylum, family, species, V1, V2, AVI,NVI, V1_2)
```

```{r}
top10_NVI <- virulence_summary %>% 
  arrange(desc(NVI)) %>%
  slice(1:10)  
genome_metadata %>% 
  filter(genome %in% rownames(top10_NVI)) %>%
  left_join(top10_NVI %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
    arrange(match(genome,rownames(top10_NVI))) %>% 
  select(phylum, family, species, V1, V2, AVI,NVI, V1_2)
```


## Mag with highest virulence factor

```{r}
virulence_summary %>% 
  filter(V1 == max(V1)) 
top10 <- virulence_summary %>% 
  arrange(desc(V1)) %>%
  slice(1:10)  
top10
```

```{r}
top10_result_V1 <- result_V1 %>% 
  rownames_to_column(., "genome") %>% 
  filter(genome %in% rownames(top10))
```
```{r}
genome_metadata %>% 
  filter(genome %in% rownames(top10)) %>%
  left_join(top10 %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
    arrange(match(genome,rownames(top10))) %>% 
  select(species, V1, V2, AVI,NVI)
```

## Mean top10 values based on v1
```{r V1_mean, comment="", message=FALSE, warning=FALSE}
top10_result_V1 %>%
  column_to_rownames(., "genome") %>% 
  colSums()%>% 
  as.data.frame() %>% 
  rename(V1=".") %>% 
  arrange(desc(V1))

top10_result_V1 %>%
  column_to_rownames(., "genome") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  group_by(region) %>%
  summarise(average=mean(V1))
```

```{r V1_plot2, comment="", message=FALSE, warning=FALSE}
top10_result_V1 %>%
  column_to_rownames(., "genome")%>%
  colSums()%>% 
  as.data.frame() %>% 
  rename(V1=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = V1, x = region, group=region, color=region, fill=region)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(alpha=0.5) +
  stat_compare_means()+
      scale_color_manual(name="Region",
          values=c("#e5bd5b", "#6b7398")) +
      scale_fill_manual(name="Region",
          values=c("#e5bd5b50", "#6b739850"))+
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 12, color="black",face="bold"),
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12))+
    guides(fill = guide_legend(override.aes = list(size=3)))
```

## Mag with highest virulence factor (v1 + v2)

```{r}
top10_v12 <- virulence_summary %>% 
  arrange(desc(V1_2)) %>%
  slice(1:10)  
top10_v12
```
```{r}
top10_result_V12 <- result_V1_2 %>% 
  rownames_to_column(., "genome") %>% 
  filter(genome %in% rownames(top10_v12))
```
```{r}
genome_metadata %>% 
  filter(genome %in% rownames(top10_v12)) %>%
  left_join(top10_v12 %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
    arrange(match(genome,rownames(top10_v12))) %>% 
  select(phylum, family, species, V1, V2, AVI,NVI, V1_2)
```

## Mean top10 values based on v1 & v2
```{r V1_mean, comment="", message=FALSE, warning=FALSE}
top10_result_V12 %>%
  column_to_rownames(., "genome") %>% 
  colSums()%>% 
  as.data.frame() %>% 
  rename(V1_2=".") %>% 
  arrange(desc(V1_2))

top10_result_V12 %>%
  column_to_rownames(., "genome") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(V1_2=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample))%>%
  group_by(region) %>%
  summarise(average=mean(V1_2))
```

```{r V12_plot2, comment="", message=FALSE, warning=FALSE}
top10_result_V12 %>%
  column_to_rownames(., "genome")%>%
  colSums()%>% 
  as.data.frame() %>% 
  rename(V1_2=".") %>%
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = V1_2, x = region, group=region, color=region, fill=region)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(alpha=0.5) +
  stat_compare_means()+
  scale_color_manual(name="Region",
          values=c("#e5bd5b", "#6b7398")) +
  scale_fill_manual(name="Region",
          values=c("#e5bd5b50", "#6b739850"))+
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    axis.ticks.x = element_blank(),
        strip.text.x = element_text(size = 12, color="black",face="bold"),
        strip.background = element_blank(),
        panel.grid.minor.x = element_line(size = .1, color = "grey"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        legend.text=element_text(size=10),
        legend.title = element_text(size=12))+
    guides(fill = guide_legend(override.aes = list(size=3)))
```
```{r genus_all, comment="", echo=FALSE}
phylum_summary <- genome_counts %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,region) %>%
  summarise(relabun=sum(count))

genus_arrange <- genus_summary %>%
    group_by(genus) %>%
    summarise(mean=sum(relabun)) %>%
    filter(genus != "g__")%>%
    arrange(-mean) %>%
    dplyr::select(genus) %>%
    mutate(genus= sub("^g__", "", genus)) %>%
    pull()

genus_summary %>%
    group_by(genus) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Daneborg_mean=mean(relabun[region=="Daneborg"]*100, na.rm=T),
              Daneborg_sd=sd(relabun[region=="Daneborg"]*100, na.rm=T),
              Ittoqqortoormiit_mean=mean(relabun[region=="Ittoqqortoormiit"]*100, na.rm=T),
              Ittoqqortoormiit_sd=sd(relabun[region=="Ittoqqortoormiit"]*100, na.rm=T)) %>%
    mutate(total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           Daneborg=str_c(round(Daneborg_mean,2),"±",round(Daneborg_sd,2)),
           Ittoqqortoormiit=str_c(round(Ittoqqortoormiit_mean,2),"±",round(Ittoqqortoormiit_sd,2))) %>% 
    arrange(-total_mean) %>% 
    select(genus,total,Daneborg,Ittoqqortoormiit) %>% 
    tt()
```

```{r virul_phylum, comment="", echo=FALSE}
virulence_summary_meta %>%
    group_by(phylum) %>%
  summarise(V1_mean=mean(V1), V1_sd=sd(V1)) %>% 
  arrange(-V1_mean)
```


```{r virul_family, comment="", echo=FALSE}
virulence_summary_meta %>%
    group_by(family) %>%
  summarise(V1=mean(V1)) %>% 
  arrange(-V1)
```

```{r virul_family, comment="", echo=FALSE}
virulence_summary_meta %>%
    group_by(genus) %>%
  summarise(V1=mean(V1)) %>% 
  arrange(-V1)
```

```{r genus_all, comment="", echo=FALSE}
genus_summary <- genome_counts_rel %>% 
  rownames_to_column(., "genome") %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  select(sample, genome, count, region) %>% 
  group_by(sample,genome,region) %>%
  summarise(relabun=sum(count)) %>% 
    left_join(genome_metadata, by = join_by(genome == genome))

abundant_mags <- genus_summary %>%
    group_by(genome) %>%
    summarise(total_mean=mean(relabun*100, na.rm=T),
              total_sd=sd(relabun*100, na.rm=T),
              Daneborg_mean=mean(relabun[region=="Daneborg"]*100, na.rm=T),
              Daneborg_sd=sd(relabun[region=="Daneborg"]*100, na.rm=T),
              Ittoqqortoormiit_mean=mean(relabun[region=="Ittoqqortoormiit"]*100, na.rm=T),
              Ittoqqortoormiit_sd=sd(relabun[region=="Ittoqqortoormiit"]*100, na.rm=T)) %>%
    mutate(total=str_c(round(total_mean,2),"±",round(total_sd,2)),
           Daneborg=str_c(round(Daneborg_mean,2),"±",round(Daneborg_sd,2)),
           Ittoqqortoormiit=str_c(round(Ittoqqortoormiit_mean,2),"±",round(Ittoqqortoormiit_sd,2))) 

# virulence_summary_meta <- virulence_summary %>% 
#   rownames_to_column("genome") %>% 
#   left_join(genome_metadata, by=join_by(genome==genome))

abundant_mags %>% 
    arrange(-Ittoqqortoormiit_mean) %>% 
    select(genome,total,Daneborg,Ittoqqortoormiit) %>% 
    slice(1:15) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(virulence_summary %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
  arrange(-V1_2)


abundant_mags %>% 
    arrange(-Ittoqqortoormiit_mean) %>% 
    select(genome,total,Daneborg,Ittoqqortoormiit) %>% 
    slice(1:15) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(virulence_summary %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
  summarise(average=mean(V1_2))

```
```{r}
abundant_mags %>% 
    arrange(-Daneborg_mean) %>% 
    select(genome,total,Daneborg,Ittoqqortoormiit) %>% 
    slice(1:15) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(virulence_summary %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
  arrange(-V1_2)

abundant_mags %>% 
    arrange(-Daneborg_mean) %>% 
    select(genome,total,Daneborg,Ittoqqortoormiit) %>% 
    slice(1:15) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(virulence_summary %>% rownames_to_column("genome"), by = join_by(genome == genome)) %>% 
  summarise(average=mean(V1_2))  
```
