---
title: "AlberdiLab | Aizpurua et al. 2026"
subtitle: "Ancestral traits of Greenland’s sled dog gut microbiomes"
author:
  - Ostaizka Aizpurua^[University of Copenhagen, ostaizka.aizpurua@sund.ku.dk]
  - Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/sled_dogs_metagenomics
description: |
  Data analysis code for the study on the study of gut microbiomes of sled dogs and beyond.
link-citations: yes
github-repo: alberdilab/sled_dogs_metagenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of gut microbiomes of Greenlandic sled dogs.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/sled_dogs_metagenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
#library(rairtable)
#library(janitor)
#library(broom)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
 library(ggplot2)
 library(ggrepel)
 library(ggpubr)
 library(ggnewscale)
 library(gridExtra)
 library(ggtreeExtra)
 library(ggtree)
# library(ggh4x)
 library(UpSetR)
# library(viridis)

# For statistics
# library(spaa)
 library(vegan)
# library(Rtsne)
# library(geiger)
# library(hilldiv2)
 library(distillR)
# library(ANCOMBC)
# library(lme4)
# library(nlme)
 library(pairwiseAdonis)
 library(emmeans)
# library(pheatmap)
 library(rstatix)
# library(broom)
```

<!--chapter:end:index.Rmd-->

# Prepare data

Load the original data files outputted by the bioinformatic pipeline.

### Reads
```{r load_reads, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts <- read_tsv("resources/counts.tsv") %>%
    rename(genome=1) %>% 
    rename_with(~ str_remove(.x, "_EKDL.*")) %>% 
    rename_with(~ str_remove(.x, "_M$"))
```

### Sample metadata
```{r load_sample_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
sample_metadata <- read_csv("resources/sample_metadata.csv")
```

### Filter read counts
```{r load_read_counts, warning=FALSE, comments="", message=FALSE, eval=FALSE}
read_counts <- read_counts %>%
  select(one_of(c("genome",sample_metadata$sample))) %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) 
```

### Genome taxonomy
```{r load_genome_metadata, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_taxonomy <- read_tsv("resources/genome_taxonomy.tsv") %>% 
  rename(genome = user_genome) %>% 
  separate(
    classification,
    into  = c("domain", "phylum", "class", "order", "family", "genus","species"),
    sep   = ";",
    fill  = "right",
    remove = TRUE
  ) %>%
  mutate(across(domain:species, ~ str_remove(., "^[a-z]__"))) %>% 
  select(genome,domain,phylum,class,order,family,genus,species) %>% 
  filter(genome %in% read_counts$genome)

genome_features <- read_tsv("resources/quality_report.tsv") %>% 
  rename(genome=1) %>% 
  rename(length=Genome_Size) %>% 
  filter(genome %in% read_counts$genome)
  
genome_metadata <- left_join(genome_taxonomy,genome_features, by="genome")

genome_metadata <- genome_metadata %>% 
  mutate(phylum = case_when(
        phylum == "Bacillota_A" ~ "Bacillota",
        phylum == "Bacillota_C" ~ "Bacillota",
        phylum == "Bacillota_B" ~ "Bacillota",
        phylum == "Bacillota_I" ~ "Bacillota",
        TRUE ~ phylum))
```

### Genome base hits

```{r load_genome_hits, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_coverage <- read_tsv("resources/bases.tsv") %>%
  rename(genome = 1) %>%
  rename_with( ~ str_remove(.x, "_EKDL.*")) %>%
  rename_with( ~ str_remove(.x, "_M$")) %>%
  filter(genome %in% read_counts$genome) %>%
  select(one_of(c("genome",sample_metadata$sample))) 
```


### Genome tree

```{r load_genome_tree, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_tree <- read_tree("resources/bacteria.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "")
genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")
genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

### Genome annotations

```{r load_genome_annotation, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_annotations <- read_tsv("resources/gene_annotations.tsv.xz") %>%
  mutate(genome = str_extract(gene, "^[^^]+")) %>%
  mutate(ec = str_c("[EC:",ec,"]")) %>% 
  select(genome, everything())
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Filter reads by coverage
```{r match, warning=FALSE, comments="", message=FALSE, eval=FALSE}
# Keep only genomes and samples present in both datasets
common_genomes <- intersect(read_counts$genome, genome_coverage$genome)
common_samples <- intersect(colnames(read_counts)[-1], colnames(genome_coverage)[-1])

read_counts <- read_counts %>%
  filter(genome %in% common_genomes) %>%
  select(genome, all_of(common_samples)) %>%
  arrange(match(genome, common_genomes))

genome_coverage <- genome_coverage %>%
  filter(genome %in% common_genomes) %>%
  select(genome, all_of(common_samples)) %>%
  arrange(match(genome, common_genomes))
```

```{r filter_coverage, warning=FALSE, comments="", message=FALSE, eval=FALSE}
min_coverage=0.3
# read_counts_filt <- genome_coverage %>%
#   mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
#   mutate(across(-1, ~ . * read_counts[[cur_column()]]))

read_counts_filt <- read_counts %>%
  select(genome) %>%
  bind_cols(
    genome_coverage %>%
      select(-genome) %>%               # drop genome column
      mutate(across(everything(), ~ as.numeric(. > min_coverage))) %>%  # 0/1 coverage
      map2_df(.x = ., .y = read_counts %>% select(-genome), ~ .x * .y)   # multiply by read counts
  )
```

### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  slice(match(genome_metadata$genome, genome)) %>%  # reorder rows to match genome_metadata
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength)))
```

### Distill annotations into GIFTs


```{r distill_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol=1,annotcol=c(6,7,8,9), verbosity=TRUE)
genome_gifts <- genome_gifts[, !grepl("^S", colnames(genome_gifts))]
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    mutate(phylum = str_remove(phylum, "^p__")) %>% 
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```


## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(
  sample_metadata,
  genome_metadata,
  genome_metadata_bacteria,
  read_counts,
  genome_counts,
  genome_counts_filt,
  genome_tree,
  genome_gifts,
  phylum_colors,
  file = "resources/data.Rdata"
)
```
```{r extra_objects, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(genome_gifts, 
     file = "resources/genome_gifts.Rdata")
```


<!--chapter:end:01_prepare_data.Rmd-->

# Data statistics

```{r load_data_stats}
load("resources/data.Rdata")
```

```{r}
seq <- read_counts %>% 
  column_to_rownames(., "genome") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename(seq=".") %>% 
  arrange(seq)
```

## Sample overview 

```{r}
sample_metadata %>% 
  nrow()
```

```{r}
sample_metadata %>% 
  group_by(country) %>% 
  summarise(count=n())
```

```{r}
sample_metadata %>% 
  group_by(dog_type) %>% 
  summarise(count=n())
```

<!--chapter:end:02_data_statistics.Rmd-->

# MAG catalogue

```{r load_data_mag, message=FALSE, warning=FALSE}
load("resources/data.Rdata")
load("resources/genome_gifts.Rdata")
```

## Genome phylogeny

```{r genome_phylogeny, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
# Generate the phylum color heatmap
phylum_heatmap <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome,phylum) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome")

# Generate new species table
newspecies_table <- genome_metadata %>% 
  mutate(newspecies=ifelse(species=="","Y","N")) %>% 
  select(genome,newspecies) %>% 
  column_to_rownames(var = "genome")

# Generate  basal tree
circular_tree <- force.ultrametric(genome_tree, method="extend") %>% # extend to ultrametric for the sake of visualisation
    ggtree(., layout="fan", open.angle=10, size=0.2)

# Add phylum ring
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=0.05, width=0.05, color=NA, colnames=FALSE) +
        scale_fill_manual(values=phylum_colors, name = "Phylum") +
guides(fill = guide_legend(override.aes = list(alpha = 1)))+
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))

# Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add Completeness ring
circular_tree <- circular_tree +
        new_scale_fill() +
        scale_fill_gradient(low = "#d1f4ba", high = "#f4baba") +
        geom_fruit(
                data=genome_metadata,
                geom=geom_bar,
                mapping = aes(x=Completeness, y=genome, fill=Contamination),
                offset = 0.24,
                pwidth = 0.1,
                orientation="y",
              stat="identity")

# Add genome-size ring
circular_tree <- circular_tree + new_scale_fill()

circular_tree <- gheatmap(circular_tree, newspecies_table, offset=0.15, width=0.05, color=NA, colnames=FALSE) +
        scale_fill_manual(values=c("#f4f4f4","#74C8AE"), name = "New species")+
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))

circular_tree <- circular_tree + new_scale_fill()

circular_tree <-  circular_tree +
  geom_fruit(
    data=genome_metadata,
    geom=geom_bar,
    mapping = aes(x=length, y=genome),
    fill = "#1e6e55",
    offset = 0.05,
    orientation="y",
    stat="identity")

circular_tree <-  circular_tree +
  new_scale_fill() +
  annotate('text', x=1.45, y=0, label='Phylum', family='sans', size=4, hjust = 0.004) +
  annotate('text', x=1.7, y=0, label='Genome quality', family='sans', size=4, hjust = 0.004)+
  annotate('text', x=1.9, y=0, label='Genome size', family='sans', size=4, hjust = 0.004) +
  annotate('text', x=1.55, y=0, label='New species', family='sans', size=4, hjust = 0.01)

  

# circular_tree <-  circular_tree +
#         annotate('text', x=1.4, y=0, label='Phylum',
#            angle = 0, hjust = 0.5, family='arial', size=3.5) +
#         annotate('text', x=1.7, y=0, label='Genome quality',
#            angle = 0, hjust = 0.5, family='arial', size=3.5) +
#         annotate('text', x=1.9, y=0, label='Genome size',
#            angle = 0, hjust = 0.5, family='arial', size=3.5) +
#         annotate('text', x=1.55, y=0, label='New species',
#            angle = 0, hjust = 0.5, family='arial', size=3.5)

#Plot circular tree
circular_tree %>% open_tree(30) %>% rotate_tree(90)
```

## Taxonomy overview
Mag number
```{r}
nrow(genome_metadata)
```

```{r}
genome_metadata %>% 
  distinct(phylum) %>% 
  tt()
```



```{r genome_taxonomy_bacteria, message=FALSE, warning=FALSE}
tax_mag <-genome_metadata %>% 
  group_by(phylum) %>%
  summarise(mag_n=n()) 

tax_mag %>%
  mutate(percetage_mag=round(mag_n*100/sum(mag_n), 2)) %>% 
  arrange(-percetage_mag) %>%
  tt()
```


```{r}
genome_metadata %>% 
  distinct(phylum, family) %>% 
  tt()
```





## Genome quality

```{r}
genome_metadata %>% 
  summarise(Contamination_mean=mean(Contamination),sd_cont=sd(Contamination),Completeness_mean=mean(Completeness),sd_comp=sd(Completeness)) %>%
  mutate(Contamination=str_c(round(Contamination_mean,2),"±",round(sd_cont,2)),
         Completeness=str_c(round(Completeness_mean,2),"±",round(sd_comp,2))) %>% 
  select(Completeness, Contamination) %>% 
  tt()
```

```{r genome_quality_plot, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}

#Generate quality biplot
genome_biplot <- genome_metadata %>%
  select(c(genome,domain,phylum,Completeness,Contamination,length)) %>%
  arrange(match(genome, rev(genome_tree$tip.label))) %>% #sort MAGs according to phylogenetic tree
  ggplot(aes(x=Completeness,y=Contamination,size=length,color=phylum)) +
              geom_point(alpha=0.7) +
                    ylim(c(10,0)) +
                    scale_color_manual(values=phylum_colors) +
                    labs(y= "Contamination", x = "Completeness") +
                    theme_classic() +
                    theme(legend.position = "none")

#Generate Contamination boxplot
genome_Contamination <- genome_metadata %>%
            ggplot(aes(y=Contamination)) +
                    ylim(c(10,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_void() +
                    theme(legend.position = "none",
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        axis.text.y=element_blank(),
                        axis.ticks.y=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate Completeness boxplot
genome_Completeness <- genome_metadata %>%
        ggplot(aes(x=Completeness)) +
                xlim(c(50,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_void() +
                theme(legend.position = "none",
                    axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks.y=element_blank(),
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
grid.arrange(grobs = list(genome_Completeness,genome_biplot,genome_Contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))

```

## Functional overview

```{r function_heatmap, message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}

# Aggregate basal GIFT into elements
function_table <- genome_gifts %>%
    to.elements(., GIFT_db)

# Generate  basal tree
function_tree <- force.ultrametric(genome_tree, method="extend") %>%
                ggtree(., size = 0.3) 

#Add phylum colors next to the tree tips
function_tree <- gheatmap(function_tree, phylum_heatmap, offset=0, width=0.1, color=NA, colnames=FALSE) +
            scale_fill_manual(values=phylum_colors) +
            labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

#Add functions heatmap
function_tree <- gheatmap(function_tree, function_table, offset=0.5, width=3.5, color=NA, colnames=FALSE) +
            vexpand(.08) +
            coord_cartesian(clip = "off") +
            scale_fill_gradient(low = "#f4f4f4", high = "steelblue", na.value="white") +
            labs(fill="GIFT")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

# Add Completeness barplots
function_tree <- function_tree +
            geom_fruit(data=genome_metadata,
            geom=geom_bar,
            grid.params=list(axis="x", text.size=2, nbreak = 1),
            axis.params=list(vline=TRUE),
            mapping = aes(x=length, y=genome, fill=Completeness),
                 offset = 3.8,
                 orientation="y",
                 stat="identity") +
            scale_fill_gradient(low = "#cf8888", high = "#a2cc87") +
            labs(fill="Genome\nCompleteness")

function_tree
```

## Functional ordination

PCoA functional ordination with PCA loadings.

```{r function_ordination_pcoa, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
gift_pcoa <- genome_gifts %>%
    to.elements(., GIFT_db) %>%
    as.data.frame() %>%
    vegdist(method="euclidean") %>%
    pcoa()

gift_pcoa_rel_eigen <- gift_pcoa$values$Relative_eig[1:10]


# Get genome positions
gift_pcoa_vectors <- gift_pcoa$vectors %>% #extract vectors
  as.data.frame() %>% 
  select(Axis.1,Axis.2) # keep the first 2 axes

gift_pcoa_eigenvalues <- gift_pcoa$values$Eigenvalues[c(1,2)]

gift_pcoa_gifts <- cov(genome_gifts, scale(gift_pcoa_vectors)) %*% diag((gift_pcoa_eigenvalues/(nrow(genome_gifts)-1))^(-0.5)) %>%
  as.data.frame() %>% 
  rename(Axis.1=1,Axis.2=2) %>% 
  rownames_to_column(var="label") %>% 
  #get function summary vectors
  mutate(func=substr(label,1,3)) %>% 
  group_by(func) %>% 
  summarise(Axis.1=mean(Axis.1),
            Axis.2=mean(Axis.2)) %>% 
  rename(label=func) %>% 
  filter(!label %in% c("S01","S02","S03"))

```


```{r function_ordination_pcoa_plot, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
scale <- 15
gift_pcoa_vectors %>% 
  rownames_to_column(var="genome") %>% 
  left_join(genome_metadata, by="genome") %>%
  ggplot() +
      #genome positions
      scale_color_manual(values=phylum_colors)+
      geom_point(aes(x=Axis.1,y=Axis.2, color=phylum, size=length), 
                 alpha=0.9, shape=16) +
      #scale_color_manual(values=phylum_colors) +
      scale_size_continuous(range = c(0.1,5)) +
      #loading positions
      geom_segment(data=gift_pcoa_gifts, 
                   aes(x=0, y=0, xend=Axis.1 * scale, yend=Axis.2 * scale),
                    arrow = arrow(length = unit(0.3, "cm"), 
                    type = "open", 
                    angle = 25),
                    linewidth = 0.5, 
                    color = "black") +
     #Primary and secondary scale adjustments
     scale_x_continuous(name = paste0("PCoA1 (",round(gift_pcoa_rel_eigen[1]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA1")
            ) +
     scale_y_continuous(name = paste0("PCoA2 (",round(gift_pcoa_rel_eigen[2]*100, digits = 2), " %)"),
                      sec.axis = sec_axis(~ . / scale, name = "Loadings on PCoA2")
            ) +
    geom_label_repel(data = gift_pcoa_gifts,
                     aes(label = label, x = Axis.1 * scale, y = Axis.2 * scale),
                     segment.color = 'transparent') +
    theme_minimal() + 
    theme(legend.position = "none")
```

```{r function_ordination_pcoa_plot_treatment_time_mci, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  filter(!is.na(dog_type)) %>% 
  group_by(genome,dog_type) %>% 
  summarise(count=mean(count)) %>% 
  left_join(gift_pcoa_vectors %>% rownames_to_column(var="genome"), by="genome") %>%
  left_join(genome_metadata, by="genome") %>%
  ggplot(aes(x=Axis.1, y=count, group=dog_type, color=dog_type)) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3), se = FALSE) +
  theme_minimal() + 
  theme(legend.position = "none")
```



<!--chapter:end:03_mag_catalogue.Rmd-->

# Differences in the gut microbiota community and funciton of canids

## Filter data

```{r load_data_community, comment="", message=FALSE, warning=FALSE}
load("resources/data.Rdata")
load("output/alpha_all.RData")
```

## Sample overview
```{r total, comment="", message=FALSE, warning=FALSE}
sample_metadata %>% 
  summarise(total_samples=n()) %>% 
  pull() %>% 
  cat()
```

```{r summary_sam, comment="", message=FALSE, warning=FALSE}
sample_metadata %>% 
  group_by(dog_type, season) %>% 
  summarise(n=n()) %>% 
  rename(Canid=1, Season=2, N=3) %>% 
  tt()
```

## Sled dogs: number of individuals ineach season and day

```{r summary_greenland, comment="", message=FALSE, warning=FALSE}
sample_metadata  %>% 
  filter(country=="Greenland") %>% 
  group_by(season, day) %>% 
  summarise(n=n()) %>% 
  rename(Season=1, Day=2, N=3)%>% 
  tt()
```

```{r}
remove_sled <- sample_metadata %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample)

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")
genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)

phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()
phylo_genome <- genome_counts_filt %>%
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  filter(genome %in% rownames(phylo_genome)) %>%
  mutate(genome2 = genome) %>% 
  column_to_rownames("genome2") %>%
  dplyr::select(domain, phylum, class, order, family, genus, species, genome) %>% 
  as.matrix() %>%
  tax_table()

physeq_genome <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
```

## Taxonomy overview 

### Stacked barplot
```{r taxonomy_barplot_canids, fig.height=10, fig.width=20, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
    geom_bar(stat="identity", colour="white", linewidth=0.1) + 
    scale_fill_manual(values=phylum_colors) +
  facet_grid(~ country, scale="free", space="free")+
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 8, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```
```{r taxonomy_barplot, fig.height=10, fig.width=15, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(., genome_metadata, by = join_by(genome == genome)) %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
  geom_bar(stat="identity", colour="white", linewidth=0.1) + 
  scale_fill_manual(values = phylum_colors) +
  facet_grid(~factor(dog_type, labels=c("companion" = "Companion dogs", "sled_dog" = "Sled dogs", "wolf" = "Wolves")), scale="free", space="free")+
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

**Number of bacteria phyla**

```{r phyla, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

**Number of Archaea phyla**

```{r arch, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

### Genus and species annotation

**Number of MAGs without species-level annotation**
```{r nonspe, comment="", message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "") %>%
  summarize(Mag_nospecies = n())%>%
  select(Mag_nospecies) %>% 
  pull() %>% 
  cat()
```
```{r nonspe_phylum, comment="", message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage_nosp=100*count_nospecies/count_total) %>% 
  arrange(percentage_nosp) %>% 
  tt()
```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", message=FALSE, warning=FALSE}
nmags <- nrow(genome_metadata)
nonspecies <- genome_metadata %>%
    filter(species == "") %>%
    nrow()
perct <- nonspecies*100/nmags
perct %>% 
  cat()
```

**Number of MAGs without genera-level annotation**
```{r nongenera, comment="", message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
    filter(genus == "") %>%
    nrow()
cat(nongenera)
```

**Percentage of MAGs without genera-level annotation**
```{r gen_percet, comment="", message=FALSE, warning=FALSE}
perct_g <- nongenera*100/nmags
perct_g %>% 
  cat()
```

### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,dog_type) %>%
  summarise(relabun=sum(count))
```

#### Origin: Companion vs Sled dogs vs Wolves
```{r taxonomy_phylum_summary_origin, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T),
    Dog_mean = mean(relabun[dog_type == "companion"] * 100, na.rm = T),
    Dog_sd = sd(relabun[dog_type == "companion"] * 100, na.rm = T),
    Sled_dog_mean = mean(relabun[dog_type == "sled_dog"] * 100, na.rm = T),
    Sled_dog_sd = sd(relabun[dog_type == "sled_dog"] * 100, na.rm = T),
    Wolf_mean = mean(relabun[dog_type == "wolf"] * 100, na.rm = T),
    Wolf_sd = sd(relabun[dog_type == "wolf"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    "Companion dogs" = str_c(round(Dog_mean, 3), "±", round(Dog_sd, 3)),
    "Sled dogs" = str_c(round(Sled_dog_mean, 3), "±", round(Sled_dog_sd, 3)),
    Wolves = str_c(round(Wolf_mean, 3), "±", round(Wolf_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(phylum, Total, "Companion dogs", "Sled dogs", Wolves) %>%
  rename(Phylum=phylum) %>% 
  tt()
```

**Using CLR transformation**

```{r}
set.seed(0206)
phylo_counts <- genome_counts_filt %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_geno_meta <- genome_metadata %>%
  column_to_rownames(var="genome")%>%
  as.matrix()%>%
  tax_table()

physeq_genome_filt <- phyloseq(phylo_counts,phylo_geno_meta)
physeq_phylum <- tax_glom(physeq_genome_filt, taxrank = "phylum")
physeq_phylum_clr <- microbiome::transform(physeq_phylum, "clr")

df <- psmelt(physeq_phylum_clr)

results <- df %>%
  left_join(., sample_metadata, by=join_by("Sample"=="sample")) %>% 
  group_by(phylum) %>%
  do({
    model <- lm(Abundance ~ dog_type, data = .)
    as.data.frame(pairs(emmeans(model, ~ dog_type), adjust = "BH"))
  }) %>%
  ungroup()%>% 
  filter(p.value < 0.05) 

```
```{r}
results %>% 
  filter(contrast=="companion - sled_dog")%>% 
  tt()
```

```{r}
results %>% 
  filter(contrast=="companion - wolf")%>% 
  tt()
```

```{r}
results %>% 
  filter(contrast=="sled_dog - wolf")%>% 
  tt()
```

## Alpha diversity
```{r alpha_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  filter(genome %in% genome_metadata_bacteria$genome) %>% 
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample))
```

```{r save_alpha, eval=FALSE}
save(alpha_div, file="output/alpha_all.RData")
```

```{r alpha_div_diets_summary_all, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
     total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Dog_mean = mean(value[dog_type == "companion"], na.rm = T),
    Dog_sd = sd(value[dog_type == "companion"], na.rm = T),
    Sled_dog_mean = mean(value[dog_type == "sled_dog"], na.rm = T),
    Sled_dog_sd = sd(value[dog_type == "sled_dog"], na.rm = T),
    Wolf_mean = mean(value[dog_type == "wolf"], na.rm = T),
    Wolf_sd = sd(value[dog_type == "wolf"], na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    "Companion dogs" = str_c(round(Dog_mean, 3), "±", round(Dog_sd, 3)),
    "Sled dogs" = str_c(round(Sled_dog_mean, 3), "±", round(Sled_dog_sd, 3)),
    Wolves = str_c(round(Wolf_mean, 3), "±", round(Wolf_sd, 3))
  ) %>%
  arrange(-Sled_dog_mean) %>%
  dplyr::select(alpha, Total, "Companion dogs", "Sled dogs", Wolves) %>%
  rename(Alpha=alpha) %>% 
  tt()
```

```{r alpha_plots, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
type_colors <- c('#41B6C0','#E5D388',"#bb6749")
new_labels <- c(
    "companion" = "Companion dogs",
    "sled_dog" = "Sled dogs",
    "wolf" = "Wolves"
    )

alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(
    y = value,
    x = dog_type,
    group = dog_type,
    color = dog_type,
    fill = dog_type
  )) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_color_manual(
    name = "Canids",
    values = type_colors,
    labels = new_labels
  ) +
  scale_fill_manual(
    name = "Canids",
    values = type_colors,
    labels = new_labels
  ) +
facet_wrap(
  ~ factor(
    metric,
    levels = c("richness","neutral","phylogenetic"),
    labels = c("Richness","Neutral","Phylogenetic")
  ),
  scales = "free",
  ncol = 5
) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  kruskal_test(value ~ dog_type) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))%>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  wilcox_test(value ~ dog_type, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)%>% 
  tt()
```

## Gut microbiota composition based on beta diversity

```{r}
type_colors <- c(
  companion = "#41B6C0",
  sled_dog = "#E5D388",
  wolf = "#bb6749"
)
new_labels <- c(
    "companion" = "Companion dogs",
    "sled_dog" = "Sled dogs",
    "wolf" = "Wolves"
    )

load("output/beta.RData")
```

```{r hillpair, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
hillpair <- function(data,q=c(0,1,2),metric=c("S","C","U","V"),tree,dist,tau,out="dist"){

  ###
  # Type of analysis detection
  ###

  #Hill numbers type definition
  if(missing(tree) && missing(dist)){hilltype="neutral"}
  if(!missing(tree) && missing(dist)){hilltype="phylogenetic"}
  if(missing(tree) && !missing(dist)){hilltype="functional"}
  if(!missing(tree) && !missing(dist)) stop("Phylogenetic and functional trait information cannot be added at once. Use either phylogenetic (tree) or functional (fun) information.")

  ###
  # Create pairwise tables
  ###

  # List of pairwise combinations
  pairs <- list()
  for (i in 1:(ncol(data)-1)) {
    for (j in (i + 1):ncol(data)) {
      pair <- data[, c(i, j), drop = FALSE]
      pairs <- append(pairs, list(pair))
    }
  }

  # List of pairwise sample names
  pairs_names <- list()
  for (i in 1:(ncol(data)-1)) {
    for (j in (i + 1):ncol(data)) {
      pair_names <- colnames(data[, c(i, j), drop = FALSE])
      pairs_names <- append(pairs_names, list(pair_names))
    }
  }

  ###
  # Estimate processing time
  ###
  pairs_length <- length(pairs_names)
  start.time <- Sys.time()
  if(hilltype == "neutral"){suppressMessages(pairs_dist <- lapply(pairs[1:10], hilldiss, q=q, metric=metric))}
  if(hilltype == "phylogenetic"){suppressMessages(pairs_dist <- lapply(pairs[1:10], hilldiss, q=q, metric=metric, tree=tree))}
  if(hilltype == "functional"){suppressMessages(pairs_dist <- lapply(pairs[1:10], hilldiss, q=q, metric=metric, dist=dist, tau=tau))}
  end.time <- Sys.time()
  time_estimate <- round((end.time - start.time)/10*pairs_length)
  options(warn=1)
  message(paste0("The estimated time for calculating the ",pairs_length," pairwise combinations is ",time_estimate+2, " seconds."))

  ###
  # Compute all pairwise distances
  ###

   if(hilltype == "neutral"){suppressMessages(pairs_dist <- lapply(pairs, hilldiss, q=q, metric=metric))}
   if(hilltype == "phylogenetic"){suppressMessages(pairs_dist <- lapply(pairs, hilldiss, q=q, metric=metric, tree=tree))}
   if(hilltype == "functional"){suppressMessages(pairs_dist <- lapply(pairs, hilldiss, q=q, metric=metric, dist=dist, tau=tau))}

  ###
  # Arrange output
  ###

  #Flatten list into matrix
  flatten_matrix <- function(matrix){
      matrix %>%
      as.data.frame() %>%
      pivot_longer(cols = everything(), names_to = "Variable") %>%
      select(value) %>%
      pull()
      }

    pairs_dist_matrix <- pairs_dist %>%
      lapply(flatten_matrix) %>%
      do.call(rbind,.) %>%
      as.data.frame()

    #List metric combinations
    if(length(q)>1){
      metric_combinations <- expand.grid(colnames(pairs_dist[[1]]),rownames(pairs_dist[[1]])) %>%
          mutate(name = paste(Var2, Var1, sep = "") ) %>%
          select(name)  %>%
          pull()
    }else{
      metric_combinations <- metric
    }

    #Assign column names to different metrics
    colnames(pairs_dist_matrix) <- metric_combinations

    #Add sample names
    pairs_dist_matrix <- pairs_dist_matrix %>%
      mutate(first=unlist(lapply(pairs_names, function(x) x[1]))) %>%
      relocate(first, .before = 1) %>%
      mutate(second=unlist(lapply(pairs_names, function(x) x[2]))) %>%
      relocate(second, .after = 1)

    #Convert table into distance object
    if (out == "dist"){
      create_distance_matrix <- function(column_name) {
        sample_names <- colnames(data)
        matrix_df <- matrix(
          0,
          nrow = length(sample_names),
          ncol = length(sample_names),
          dimnames = list(sample_names, sample_names)
        )
        pair_index <- cbind(
          match(pairs_dist_matrix$first, sample_names),
          match(pairs_dist_matrix$second, sample_names)
        )
        matrix_df[pair_index] <- pairs_dist_matrix[[column_name]]
        matrix_df[pair_index[, 2:1, drop = FALSE]] <- pairs_dist_matrix[[column_name]]
        matrix_df <- as.dist(matrix_df)
        return(matrix_df)
      }

      pairs_dist_matrix <- lapply(colnames(pairs_dist_matrix[3:ncol(pairs_dist_matrix)]), create_distance_matrix)
      names(pairs_dist_matrix) <- metric_combinations

      if(length(pairs_dist_matrix) == 1){
        pairs_dist_matrix <- pairs_dist_matrix[[1]]
      }
    }

  return(pairs_dist_matrix)

}
```

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

# beta_q1p <- genome_counts_filt %>%
#   filter(genome %in% genome_metadata_bacteria$genome) %>% 
#   column_to_rownames(., "genome") %>%
#   filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
#   select_if(~!all(. == 0)) %>%
#   hillpair(., q = 1, tree = genome_tree)

save(beta_q0n, 
     beta_q1n, 
     #beta_q1p, 
     file = "output/beta.Rdata")
```

```{r q1, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
# table <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
#   select(where(~ !all(. == 0)))
# labs <- colnames(table)
# 
# beta_q1n <- hillpair(table, q = 1)
# attr(beta_q1n$S, "Labels") <- labs
# beta_q1n <- beta_q1n$S
# save(beta_q1n,file="output/beta_q1n.RData")
```

```{r clr_q1, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
### CLR
# beta_q1n <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(.) > 0) %>%
#   select(where(~ sum(.) > 0)) %>%
#   t() %>%
#   (\(x) x + 1)() %>%
#   clr() %>%
#   dist()
```

```{r bray_q1, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
### Bray curtis
# beta_q1n_bray <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(.) > 0) %>%
#   select(where(~ sum(.) > 0)) %>%
#   t() %>%
#   vegdist(method = "bray")
```

```{r helliq1, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
## Hellinger
# mat <- genome_counts_filt %>%
#   column_to_rownames("genome") %>%
#   filter(rowSums(.) > 0) %>%
#   select(where(~ sum(.) > 0)) %>%
#   t()
# 
# mat_hel <- decostand(mat, method = "hellinger")
# 
# beta_hel <- dist(mat_hel, method = "euclidean")
```

### Richness
```{r permanova__q0, comment="", message=FALSE, warning=FALSE}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]

beta_q0n_dis <- as.dist(beta_q0n_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q0n_dis)) %>%
  arrange(match(sample, labels(beta_q0n_dis))) %>%
  column_to_rownames(., "sample")

betadisper(beta_q0n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q0n_dis, sample_metadata_ordered$dog_type) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q0n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q0n_dis ~ dog_type,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```


Significant
```{r}
pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$dog_type, perm = 999) %>%
  filter(p.adjusted < 0.05) %>% 
  tt()
```

No significant
```{r}
pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$dog_type, perm = 999) %>% 
  filter(p.adjusted>0.05) %>% 
  tt()
```

Centroids
```{r}
betadisper_obj <- betadisper(beta_q0n_dis, sample_metadata_ordered$dog_type)
dist_centroids <- dist(betadisper_obj$centroids)
```

### Neutral
```{r permanova__q1, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)
beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]

beta_q1n_dis <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q1n_dis)) %>%
  arrange(match(sample, labels(beta_q1n_dis))) %>%
  column_to_rownames(., "sample")

all(rownames(sample_metadata_ordered) == labels(beta_q1n_dis))

#betadisper(beta_q1n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q1n_dis, sample_metadata_ordered$dog_type)%>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n_dis ~ dog_type + country + locality,
        data = sample_metadata_ordered,
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q1n_dis ~ dog_type + country + locality,
        data = sample_metadata_ordered,
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q1n_dis ~ dog_type,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

***Pairwise***
Significant
```{r}
pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$dog_type, perm = 999) %>%
  filter(p.adjusted < 0.05)
```

Centroids
```{r}
betadisper_obj <- betadisper(beta_q1n_dis, sample_metadata_ordered$dog_type)
dist_centroids <- dist(betadisper_obj$centroids)
dist_centroids
```

***NMDS***
```{r generate_nmds, comment="", message=FALSE, warning=FALSE, eval=FALSE}
nmds_beta <- beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup()
save(nmds_beta,file="output/nmds_beta.RData")

nmds_beta_country <- beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type, country) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup()
save(nmds_beta_country,file="output/nmds_beta_country.RData")
```

```{r nmds_plot_type, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
load("output/nmds_beta.RData")
nmds_beta %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = 'Canids') 
```

```{r nmds_plot_country, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
load("output/nmds_beta_country.RData")

nmds_beta_country %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels)+
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = 'Canids')
```


#### PCoa country
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100
```

```{r pcoa_neutral_loc, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
centroids <- pcoa_df %>%
  group_by(dog_type, country) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )
centroids_loc <- centroids %>%
  mutate(group = paste(dog_type, country, sep = "_"))

pcoa_df1 <- pcoa_df %>%
  mutate(group = paste(dog_type, country, sep = "_"))

ggplot(pcoa_df1, aes(Axis.1, Axis.2, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(
    data = centroids_loc,
    aes(Axis.1, Axis.2, fill = group),
    color = "black",
    shape = 21,
    size = 6,
    show.legend = FALSE
  ) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Group"
  )  +
  theme_minimal() +
  theme(legend.text = element_text(size = 8)) +
  guides(color = guide_legend(ncol = 1))
```

***Centroid distances***

```{r centroid_distances_type_locality, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=12, fig.fullwidth=TRUE}
centroids_loc <- pcoa_df %>%
  group_by(dog_type, country) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )

centroids_loc <- centroids_loc %>%
  mutate(group = paste(dog_type, country, sep = "_"))


coords <- centroids_loc[, c("Axis.1", "Axis.2")] %>%
  as.data.frame()
rownames(coords) <- centroids_loc$group

dist_loc <- dist(coords)

dist_loc_df <- as.data.frame(as.table(as.matrix(dist_loc))) %>%
  rename(group1 = Var1,
         group2 = Var2,
         distance = Freq) %>%
  filter(group1 != group2)

ggplot(dist_loc_df, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank()) +
  labs(fill = "Centroids distance")
```
```{r}
 dist_loc_df %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(group1, group2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) %>% 
  arrange(distance) %>% 
  tt()
```

#### PCoa dog type
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100

centroids <- pcoa_df %>%
  group_by(dog_type) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )
```
```{r pcoa_neutral_dogtype, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
ggplot(pcoa_df, aes(Axis.1, Axis.2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(data = centroids,
             aes(Axis.1, Axis.2, fill = dog_type),
             color = "black",
             shape = 21,
             size = 6,
             show.legend = FALSE) +
    scale_color_manual(
  name = "Canid",
  values = type_colors,
  labels = new_labels)+
scale_fill_manual(
  name = "Canid",        
  values = type_colors,
  labels = new_labels) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Canid"
  ) +
  theme_classic(base_size = 14)
```

Show samples closest to companion dogs
```{r pcoa_neutral_type_sample, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
domestic_centroid <- centroids %>% filter(dog_type == "companion")
sled_centroid <- centroids %>% filter(dog_type == "sled_dog")

pcoa_df <- pcoa_df %>%
  mutate(dist_to_domestic = sqrt((Axis.1 - domestic_centroid$Axis.1)^2 +
                                 (Axis.2 - domestic_centroid$Axis.2)^2),
    dist_to_sled = sqrt((Axis.1 - sled_centroid$Axis.1)^2 +
                        (Axis.2 - sled_centroid$Axis.2)^2)
  )


closest_sled_to_domestic <- pcoa_df %>%
  filter(dog_type == "sled_dog") %>%
  arrange(dist_to_domestic) %>%
  slice_head(n = 5)

closest_samples <- closest_sled_to_domestic$sample

ggplot(pcoa_df, aes(Axis.1, Axis.2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(aes(color = dog_type), level = 0.95, linewidth = 1.2) +
  geom_point(
    data = centroids,
    aes(x = Axis.1, y = Axis.2, fill = dog_type),
    color = "black",
    shape = 21,
    size = 6,
    show.legend = FALSE
  ) +
  geom_point(
    data = pcoa_df %>% filter(sample %in% closest_samples),
    aes(x = Axis.1, y = Axis.2),
    shape = 21,
    fill = NA,
    color = "black",
    size = 5,
    stroke = 1.5
  ) +
  scale_color_manual(name = "Canid",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canid",
                    values = type_colors,
                    labels = new_labels) +
  geom_text_repel(
    data = closest_sled_to_domestic,
    aes(x = Axis.1, y = Axis.2, label = locality),
    size = 4,
    box.padding = 0.5,
    point.padding = 0.5,
    segment.color = "grey50"
  ) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Dog type"
  ) +
  theme_classic(base_size = 14)
```

**Centroid distances**
```{r cent, comment="", message=FALSE, warning=FALSE}
centroid_coords <- centroids[, c("Axis.1", "Axis.2")]
rownames(centroid_coords) <- centroids$dog_type

centroid_dist <- dist(centroid_coords)
centroid_dist_mat <- as.matrix(centroid_dist)

dist_table <- as.data.frame(as.table(centroid_dist_mat)) %>%
  rename(group1 = Var1, group2 = Var2, distance = Freq) %>%
  filter(group1 != group2) %>%
  arrange(distance)
```

```{r centroid_distances_type_figure, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.fullwidth=TRUE}
ggplot(dist_table, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_blank()) +
  labs(fill="Centroids distance")
```

#### Distance comparison test

This tests whether samples of the same dog type are more similar than different ones
```{r}
dist_mat <- as.matrix(beta_q1n_dis)

dist_df <- as.data.frame(dist_mat) %>%
  mutate(sample1 = rownames(.)) %>%
  pivot_longer(-sample1, names_to = "sample2", values_to = "distance") %>%
  filter(sample1 < sample2)

meta <- sample_metadata_ordered %>%
  mutate(sample = rownames(.))

dist_df <- dist_df %>%
  left_join(meta, by = c("sample1" = "sample")) %>%
  rename(dog1 = dog_type, loc1 = locality) %>%
  left_join(meta, by = c("sample2" = "sample")) %>%
  rename(dog2 = dog_type, loc2 = locality)

dist_df <- dist_df %>%
  mutate(
    same_dog = dog1 == dog2,
    same_locality = loc1 == loc2
  )
```

```{r}
dist_cross_loc <- dist_df %>%
  filter(!same_locality)

wilcox.test(distance ~ same_dog, data = dist_cross_loc)

ggplot(dist_cross_loc, aes(x = same_dog, y = distance)) +
  geom_boxplot() +
  labs(
    x = "Same dog type",
    y = "Beta diversity distance",
    title = "Distance comparison (across localities only)"
  )
```
If same_dog == TRUE has lower distances: suggests similarity within dog types across locations


## Differential abundance analysis
```{r}
load("output/ancombc_phylum_sled_companion_wolf.Rdata")
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
```

### Phylum level
```{r pair_all_phylum, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_types <- lapply(region_pairs_types, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

   fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome),
    fix_formula = "dog_type",
    tax_level = "phylum", 
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )
   
  res_table <- fit$res

  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})
names(results_types) <- sapply(region_pairs_types, paste, collapse = "_vs_")

save(results_types,
     file = "output/ancombc_phylum_sled_companion_wolf.Rdata")
```

#### Companion vs sled_dog

**Enriched in companion dogs**
```{r}
companion <- results_types$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
companion %>%
  nrow() %>% 
  cat()
```

```{r}
companion %>% 
  tt()
```

**Enriched in sled dogs**

```{r}
sled <- results_types$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)

sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled %>% 
  tt()
```


#### Sled dogs vs wolves

**Enriched in sled dogs**
```{r}
sled <- results_types$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)
```

```{r}
sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled %>% 
  tt()
```

**Enriched in wolves**

```{r}
wolf <- results_types$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)

wolf %>%
  nrow() %>%
  cat()
```

```{r}
wolf %>% 
  tt()
```

#### All plots
```{r generate_all, comment="", message=FALSE,warning=FALSE, eval=FALSE}
taxa <- as.data.frame(physeq_genome@tax_table)  %>% 
  distinct(phylum)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxa, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (name in names(results_types)) {

  parts <- strsplit(name, "_vs_")[[1]]
  
  name1 <- parts[1]
  name2 <- parts[2]
  
  title <- paste0("Enriched in ", name1, "  ( − LFC )   vs   ", "Enriched in ", name2, "  ( + LFC )")
  
  tab <- results_types[[name]]

  coef <- grep("^lfc_dog_type", colnames(tab), value = TRUE)[1]
  qcol <- sub("^lfc_", "q_", coef)
  
  plot_data <- tab %>%
    filter(.data[[qcol]] < 0.05) %>%
    mutate(logFC = .data[[coef]])
  
  tax_color <- plot_data %>% 
    rename(phylum=1) %>% 
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()

  plot <- ggplot(plot_data, aes(x = reorder(taxon, logFC), y = logFC, fill = taxon)) +
    geom_col() +
    scale_fill_manual(values = tax_color) +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    labs(
      title = title,
      x = "Phylum",
      y = "Log-Fold Change",
      fill = "Phylum"
    )

  ggsave(paste0("output/figures/Fig_phylum_", name, ".png"),
         plot = plot,
         width = 6,
         height = 8)
}
```

### MAG level
```{r load_all, comment="", message=FALSE, warning=FALSE}
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
load("output/ancombc_pairwise_mag_sled_companion_sepa.Rdata")
```

```{r pair_all_mag, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_mags <- lapply(region_pairs_types, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

  fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome),
    fix_formula = "dog_type",
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )

  res_table <- fit$res

  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})

names(results_mags) <- sapply(region_pairs_types, paste, collapse = "_vs_")
save(results_mags,
     file = "output/ancombc_pairwise_mag_sled_companion_sepa.Rdata")
```

#### Companion vs Sled dogs

**Total number of enriched MAGs**
```{r}
results_mags$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  nrow
```

**Enriched in companion dogs**
```{r}
companion <- results_mags$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)

all <- companion %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

```{r}
perc <- 100-(noanno*100/all)
perc %>% 
  cat()
```

```{r}
companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, genus, family, everything()) %>% 
  arrange(lfc_dog_typesled_dog) %>% 
  head(10) %>% 
  tt()
```

```{r wolf_sum, comment="", message=FALSE, warning=FALSE}
companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, genus, family, phylum, everything()) %>% 
  filter(species=="") %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count)%>% 
  head(10) %>% 
  tt()
```

```{r comp_sumq1, comment="", message=FALSE, warning=FALSE}
companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags) %>% 
  head(10) %>% 
  tt()
```

**Enriched in sled dogs**

```{r}
sled <- results_mags$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)

all <- sled %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

Percentage of bacteria annotated to species level
```{r}
perc <- 100-(noanno*100/all)
perc %>% 
  cat()
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, genus, family, everything()) %>% 
  head(10) %>% 
  tt()
```
```{r sle_phf_sumq1, comment="", message=FALSE, warning=FALSE}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, genus, family, phylum, everything()) %>% 
  filter(species=="") %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count) %>% 
  head(10) %>% 
  tt()
```
```{r slef_mag_sumq1, comment="", message=FALSE, warning=FALSE}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)%>% 
  head(10) %>% 
  tt()

sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)%>% 
  head(10) %>% 
  tt()
```

#### Sled dogs vs wolves

**Total number of enriched MAGs**
```{r}
results_mags$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  nrow
```

**Enriched in sled dogs**
```{r}
sled <- results_mags$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)

all <- sled %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```
```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, family, everything()) %>% 
  head(10) %>% 
  tt()
```

```{r sled_sumq1, comment="", message=FALSE, warning=FALSE}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags) %>% 
  head(10) %>% 
  tt()
```

**Enriched in wolves**

```{r}
wolf <- results_mags$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)

all <- wolf %>%
  nrow()
all %>% 
  cat()
```

how many of the enriched bacteria were not annotated to species
```{r}
noanno <- wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, family, everything()) %>% 
  head(10) %>% 
  tt()
```

```{r wolf_sumq1, comment="", message=FALSE, warning=FALSE}
wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)%>% 
  head(10) %>% 
  tt()

wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags) %>% 
  head(10) %>% 
  tt()
```

### All plots
```{r plots_diff, comment="", message=FALSE, warning=FALSE, eval=FALSE}
taxonomy <- data.frame(physeq_genome@tax_table) %>%
  distinct(phylum) %>% 
  rownames_to_column(., "taxon")

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (name in names(results_mags)) {

  parts <- strsplit(name, "_vs_")[[1]]
  
  name1 <- parts[1]
  name2 <- parts[2]
  
  title <- paste0("Enriched in ", name1, "  ( − LFC )   vs   ", "Enriched in ", name2, "  ( + LFC )")
  
  tab <- results_mags[[name]]

  coef <- grep("^lfc_dog_type", colnames(tab), value = TRUE)[1]
  qcol <- sub("^lfc_", "q_", coef)
  
  plot_data <- tab %>%
    filter(.data[[qcol]] < 0.05) %>%
    mutate(logFC = .data[[coef]]) %>% 
    left_join(genome_metadata, by=join_by("taxon"=="genome"))
  
  tax_color <- plot_data %>% 
    distinct(phylum) %>% 
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()

  plot <- ggplot(plot_data, aes(x = reorder(taxon, logFC), y = logFC, fill = phylum)) +
    geom_col() +
    scale_fill_manual(values = tax_color) +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 6),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    labs(
      title = title,
      x = "MAGs",
      y = "Log-Fold Change",
      fill = "Phylum"
    )

  ggsave(paste0("output/figures/Fig_mag_", name, ".png"),
         plot = plot,
         width = 10,
         height = 25)
}
```

### Species level
```{r}
genome_tax <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(genus=="",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus),
    species  = if_else(species=="",
                     if_else(str_detect(genus, "Unclassified"), genus, paste0(genus, "_Unclassified")),
                     species)
  )
```

```{r phy_sp, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  left_join(genome_tax[c(1,3,4,6,7,8)], by = "genome") %>% 
  select(-genome) %>%
  group_by(phylum, class, family, genus, species) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class, family, genus))%>%
  column_to_rownames("species") %>%
  otu_table(., taxa_are_rows = TRUE)

physeq_genome_sp <- phyloseq(phylo_genome, phylo_samples)
region_pairs_types <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
load("output/ancombc_species_sled_companion_wolf.Rdata")
```

```{r pair_all_dogtype, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_species <- lapply(region_pairs_types, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

   fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome_sp),
    fix_formula = "dog_type",
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )
   # Extract main results table
  res_table <- fit$res

  # Add comparison label
  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})
names(results_species) <- sapply(region_pairs_types, paste, collapse = "_vs_")
 save(results_species,
     file = "output/ancombc_species_sled_companion_wolf.Rdata")
```

#### Companion vs sled dogs

**Enriched in companion dogs**
```{r}
companion <- results_species$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)

all <- companion %>%
  nrow()
all %>% 
  cat()
```

How many of the enriched bacteria were not annotated to species?
```{r}
noanno <- companion %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```
```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
companion %>% 
  head(10) %>% 
  tt()
```

**Enriched in sled dogs**

```{r}
sled <- results_species$companion_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)

all <- sled %>%
  nrow()
all %>% 
  cat()
```

```{r}
sled %>% 
  head(10) %>% 
  tt()
```


#### Sled dogs vs wolves

**Enriched in sled dogs**
```{r}
sled <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)

sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled
```

**Enriched in wolves**

```{r}
wolf <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)

wolf %>%
  nrow() %>%
  cat()
```

```{r}
wolf %>% 
  head(10) %>% 
  tt()
```


## Functional

```{r gift_analysis_func_all, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts), ]
genome_counts_filt <- genome_counts_filt %>%
  select(genome, all_of(sample_metadata$sample)) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if( ~ !all(. == 0)) %>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome, ]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]

#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts, GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome, ]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>%
  select_if( ~ !is.numeric(.) || sum(.) != 0)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered, GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions, GIFT_db)

#Get community-weighed average GIFTs per sample
genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome),  ~ . / sum(.)) %>%
  column_to_rownames(., "genome")

GIFTs_elements_community <- to.community(GIFTs_elements_filtered, genome_counts_row, GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions, genome_counts_row, GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains, genome_counts_row, GIFT_db)
```


### MCI
```{r gitfs_functional_type}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type) %>%
  summarise(mean = round(mean(value), 3), sd = round(sd(value), 3)) %>%
  mutate(MCI = str_c(round(mean, 3), "±", round(sd, 3))) %>%
  select(dog_type, MCI) %>%
  rename("Canid" = "dog_type") %>%
  tt()
```

```{r}
MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample))

shapiro.test(residuals(lm(value ~ dog_type, data = MCI)))
bartlett.test(value ~ dog_type, data = MCI)

kruskal.test(value ~ dog_type, data = MCI)
pw <- pairwise.wilcox.test(x = MCI$value,
                           g = MCI$dog_type,
                           p.adjust.method = "BH")

wilcox_tbl <- pw$p.value %>%
  as.data.frame() %>%
  rownames_to_column("group1") %>%
  pivot_longer(-group1, names_to = "group2", values_to = "p_adjust") %>%
  filter(!is.na(p_adjust)) %>%
  mutate(
    comparison = paste(group1, "vs", group2),
    p_adjust = round(p_adjust, 3),
    sig = case_when(
      p_adjust < 0.001 ~ "***",
      p_adjust < 0.01  ~ "**",
      p_adjust < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(comparison, p_adjust, sig)

wilcox_tbl %>%
  rename("Compared groups" = comparison,
         "Adjusted p-value" = p_adjust) %>%
  tt()
```

### Community elements differences:

```{r}
function_matrix <- GIFTs_functions_community %>%
  as.data.frame() %>%
  as.matrix()

metadata <- sample_metadata %>%
  select(sample, dog_type) %>%
  column_to_rownames("sample")

metadata <- metadata[rownames(function_matrix), , drop=FALSE]
stopifnot(all(rownames(function_matrix) %in% rownames(metadata)))
stopifnot(all(rownames(function_matrix) == rownames(metadata)))
meta_row <- rownames_to_column(metadata, "sample")

adonis2(function_matrix ~ dog_type,
        data = metadata,
        method = "euclidean",
        permutations = 999)
```

```{r commun_wilcox_elem_type, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample")

uniqueGIFT_db <- unique(GIFT_db[c(2, 4, 5, 6)]) %>%
  unite("Function", Function:Element, sep = "_", remove = FALSE)

significant_elements <- element_gift %>%
  pivot_longer(-c(sample, dog_type),
               names_to = "trait",
               values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ dog_type)$p.value,
            .groups = "drop") %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05)

selected_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type),
               names_to = "trait",
               values_to = "value") %>%
  filter(trait %in% significant_elements$trait)

pairwise_results <- selected_data %>%
  group_by(trait) %>%
  dunn_test(value ~ dog_type, p.adjust.method = "BH") %>%
  ungroup() %>%
  filter(p.adj < 0.05) %>%
  mutate(comparison = paste(group1, "vs", group2)) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element))
```

```{r ele_plot_type, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
plot_elem <- ggplot(pairwise_results, aes(x = comparison, y = fct_rev(Function))) +
  geom_tile(aes(fill = p.adj), color = "white") +
  geom_text(aes(label = p.adj.signif), size = 4, color = "black") +
  scale_fill_gradient(low = "#b30000", high = "#fef0d9" , name = "Adjusted p-value") +
  theme_minimal() +
  labs(title = "Significant Pairwise Comparisons by Trait", x = "Locality Comparison", y = "Trait") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

#ggsave("Figures/plot_element_dog_type.png", plot = plot_elem, width = 10, height = 25, dpi = 300)

plot_elem
```

```{r barplot_ele_type2, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>%  select(sample, dog_type) %>% filter(), by = "sample")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% pairwise_results$trait) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element))

plot_data %>%
  ggplot(., aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal() +
  theme(strip.text = element_text(size = 18))
```

**Starch**
```{r starch_plot, fig.height=4, fig.width=6, fig.fullwidth=TRUE}
plot_data %>%
  filter(Element == "Starch") %>%
#  left_join(sample_metadata_origin %>% select(sample, origin), by = "sample") %>% 
  ggplot(., aes(x = dog_type, y = value, fill = dog_type, color=dog_type)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  theme_minimal()  +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  #geom_text_repel(aes(label = sample), size = 3, max.overlaps = Inf, show.legend = FALSE) +
  theme_classic()+
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.text.x = element_blank()
  ) 
```
```{r}
selected_data %>%
  group_by(trait) %>%
  dunn_test(value ~ dog_type, p.adjust.method = "BH") %>% 
  ungroup()%>% 
  filter(trait=="D0203") %>% 
  tt()
```

Differences between companion dogs and sled dogs
```{r}
pairwise_results %>%
  filter(comparison == "companion vs sled_dog") %>%
  nrow %>% 
  cat()
```

```{r barplot_ele_type, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
# element_gift <- GIFTs_elements_community %>% 
#   as.data.frame() %>% 
#   rownames_to_column(., "sample") %>% 
#   left_join(sample_metadata %>%  select(sample,dog_type), by="sample")

dog_sled <- pairwise_results %>%
  filter(comparison == "companion vs sled_dog")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sled$trait) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element)) 

plot_data %>%
  filter(dog_type != "wolf") %>%
  ggplot(., aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal() +
  theme(strip.text = element_text(size = 18))
```

```{r}
enrichment_results <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  filter(sample %in% sample_metadata$sample) %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample") %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sled$trait) %>%
  filter(dog_type != "wolf") %>%
  group_by(trait) %>%
  summarise(
    median_sled = median(value[dog_type == "sled_dog"]),
    median_domestic = median(value[dog_type == "companion"]),
    log2FC = log2((median_sled + 1) / (median_domestic + 1)),
    difference = median_sled - median_domestic
  )

# enrichment_results %>%
#   left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
#   select(log2FC, Function) %>%
#   arrange(desc(Function)) %>%
#   tt()
```

```{r plot3_sled_dom, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,3,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

enrichment_code <- enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by=join_by(trait==Code_element))

unique_codes <- unique(enrichment_code$Code_function)

gift_colors <- read_tsv("resources/gift_colors.tsv") %>%
  filter(Code_function %in% unique_codes) %>%
  mutate(legend = str_c(Code_function, " - ", Function))

gift_color_values <- setNames(gift_colors$Color, gift_colors$legend)
```

```{r final_plot_sled_dom, comment="", message=FALSE, warning=FALSE, fig.height=18, fig.width=12, fig.fullwidth=TRUE}
enrichment_code %>%
  left_join(
    gift_colors %>% select(Code_function, Color, legend),
    by = join_by(Code_function == Code_function)
  ) %>%
  ggplot(aes(x = forcats::fct_reorder(Function, log2FC), y = log2FC, fill = legend)) +
  geom_col() +
  scale_fill_manual(values = gift_color_values) +
  geom_hline(yintercept = 0) +
  coord_flip() +
  geom_text(aes(75, 0.040), label = "Enriched in\n sled dogs", color = "#666666") +
  geom_text(aes(11, -0.09), label = "Enriched in\n companion dogs", color = "#666666") +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_blank(),
    panel.background = element_blank()
  ) +
  xlab("Microbial Functional Capacity") +
  ylab("log2FC")
```


Number of enriched functions in each group
```{r}
sled_enr <- enrichment_code %>%
  left_join(., unique(GIFT_db[c(2, 5)]), by = join_by(trait == Code_element)) %>%
  filter(log2FC > 0) %>%
  group_by(Function.y) %>%
  summarise(Sled = n())

enrichment_code %>%
  left_join(., unique(GIFT_db[c(2, 5)]), by = join_by(trait == Code_element)) %>%
  filter(log2FC < 0) %>%
  group_by(Function.y) %>%
  summarise(Companion = n()) %>%
  full_join(sled_enr, by = "Function.y") %>%
  arrange(-Sled) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), 0, .))) %>%
  mutate(difference = Companion - Sled) %>% 
  tt()
```

Differences between sled dogs and wolves
```{r barplot_ele_type2_wolf, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>%  select(sample, dog_type) %>% filter(), by =
              "sample")

dog_sledwolf <- pairwise_results %>%
  filter(comparison == "sled_dog vs wolf")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sledwolf$trait) %>%
  left_join(., uniqueGIFT_db, by = join_by(trait == Code_element))

plot_data %>%
  filter(dog_type != "companion") %>%
  ggplot(., aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal() +
  theme(strip.text = element_text(size = 18))
```

```{r}
enrichment_results <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  filter(sample %in% sample_metadata$sample) %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample") %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% dog_sledwolf$trait) %>%
  filter(dog_type != "companion") %>%
  group_by(trait) %>%
  summarise(
    median_sled = median(value[dog_type == "sled_dog"]),
    median_wolves = median(value[dog_type == "wolf"]),
    log2FC = log2((median_sled + 1) /
                    (median_wolves + 1)),
    difference = median_sled - median_wolves
  )

enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  nrow()

# enrichment_results %>%
#   left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
#   select(log2FC, Function) %>%
#   tt()
```

```{r plot3_sled_domest, comment="", message=FALSE, warning=FALSE}

uniqueGIFT_db<- unique(GIFT_db[c(2,3,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

enrichment_code <- enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element))

unique_codes <- unique(enrichment_code$Code_function)

gift_colors <- read_tsv("resources/gift_colors.tsv") %>%
  filter(Code_function %in% unique_codes) %>%
  mutate(legend = str_c(Code_function, " - ", Function))

gift_color_values <- setNames(gift_colors$Color, gift_colors$legend)
```

```{r final_plot_sled_domest, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=12, fig.fullwidth=TRUE}
enrichment_code %>%
  left_join(
    gift_colors %>% select(Code_function, Color, legend),
    by = join_by(Code_function == Code_function)
  ) %>%
  ggplot(aes(x = forcats::fct_reorder(Function, log2FC), y = log2FC, fill = legend)) +
  geom_col() +
  scale_fill_manual(values = gift_color_values) +
  geom_hline(yintercept = 0) +
  coord_flip() +
  geom_text(aes(30, 0.040), label = "Enriched in\n sled dogs", color = "#666666") +
  geom_text(aes(9, -0.05), label = "Enriched in\n wolves ", color = "#666666") +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.position = "right",
    panel.background = element_blank()
  ) +
  xlab("Microbial Functional Capacity") +
  ylab("log2FC")
```

Higher in wolves (top 10)
```{r}
enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  select(log2FC, Function) %>%
  filter(log2FC < 0) %>%
  head(10) %>% 
  tt()
```

Higher in sled dogs (top 10)
```{r}
enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by = join_by(trait == Code_element)) %>%
  select(log2FC, Function) %>%
  filter(log2FC > 0) %>%
  head(10) %>% 
  tt()
```

### Community function differences:

```{r commun_wilcox_func_type, comment="", message=FALSE, warning=FALSE}
function_gift <- GIFTs_functions_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample")

uniqueFuncGIFT_db <- unique(GIFT_db[c(3, 4, 5)])

significant_functions <- function_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ dog_type)$p.value) %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05) %>%
  rownames_to_column(., "functions")  %>%
  select(-functions)

selected_data <- function_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% significant_functions$trait)

pairwise_results_func <- selected_data %>%
  group_by(trait) %>%
  dunn_test(value ~ dog_type, p.adjust.method = "BH") %>%
  filter(p.adj < 0.05) %>%
  mutate(comparison = paste(group1, "vs", group2)) %>%
  left_join(., uniqueFuncGIFT_db, by = join_by(trait == Code_function))
```

```{r pairwise_plot_func_type, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
plot_funct <- ggplot(pairwise_results_func, aes(x = comparison, y = fct_rev(Function))) +
  geom_tile(aes(fill = p.adj), color = "white") +
  geom_text(aes(label = p.adj.signif), size = 4, color = "black") +
  scale_fill_gradient(low = "#b30000", high = "#fef0d9" , name = "Adjusted p-value") +
  theme_minimal() +
  labs(title = "Significant Pairwise Comparisons by Trait", x = "Locality Comparison", y = "Trait") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

#ggsave("output/figures/plot_function_dog_type.png", plot = plot_funct, width = 8, height = 8, dpi = 300)

plot_funct
```

```{r barplot_func_dog_type, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=30, fig.fullwidth=TRUE}
function_gift <- GIFTs_functions_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  left_join(sample_metadata %>% select(sample, dog_type), by = "sample")

plot_data <- function_gift %>%
  pivot_longer(-c(sample, dog_type), names_to = "trait", values_to = "value") %>%
  filter(trait %in% pairwise_results_func$trait) %>%
  left_join(., uniqueFuncGIFT_db, by = join_by(trait == Code_function))

ggplot(plot_data, aes(x = dog_type, y = value, fill = dog_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  facet_wrap( ~ Function, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_blank())

plot_data %>%
  group_by(dog_type, Function) %>%
  summarise(mean = mean(value)) %>%
  pivot_wider(names_from = dog_type, values_from = mean) %>% 
  tt()
```

<!--chapter:end:04_canids.Rmd-->

# Management-associated ecology drives functional restructuring of the gut microbiome in Qimmit

```{r load_data_beta_sled_dogs, comment="", message=FALSE, warning=FALSE}
set.seed(1234)
load("resources/data.Rdata")
load("output/beta.Rdata")
load("output/alpha_all.RData")


sample_metadata <- sample_metadata %>% 
  filter(country=="Greenland") %>%
  filter(season != "winter") %>% 
  mutate(
    dog_group = case_when(
      locality == "Daneborg" ~ "military",
      TRUE                   ~ "civilian"
    )
  )

alpha_div <- alpha_div %>% 
  filter(sample %in% sample_metadata$sample)

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata_bacteria %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```
```{r}
locality_cols <- c(
  Daneborg = "#41B6C0",
  Ittoqqortoormiit = "#E5D388",
  Illulissat = "#834630",
  Disko = "#76b183"
)
```

## Taxonomy
**Number of bacteria phyla**
```{r phyla_sled, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() %>% 
  cat()
```

```{r taxonomy_G_barplot_sled, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(., genome_metadata, by = join_by(genome == genome)) %>%
  left_join(., sample_metadata, by = join_by(sample == sample)) %>%
  filter(count > 0) %>%
  ggplot(., aes(x = sample, y = count, fill = phylum, group = phylum)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.1) +
  scale_fill_manual(values = phylum_colors) +
  facet_grid(~ locality, scale = "free", space = "free") +
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 8, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

### Phylum relative abundances

```{r taxonomy_phylum_summary_sled, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample, phylum, locality) %>%
  summarise(relabun = sum(count))
```

```{r taxonomy_phylum_summary_origin_sled, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T),
    Daneborg_mean = mean(relabun[locality == "Daneborg"] * 100, na.rm = T),
    Daneborg_sd = sd(relabun[locality == "Daneborg"] * 100, na.rm = T),
    Disko_mean = mean(relabun[locality == "Disko"] * 100, na.rm = T),
    Disko_sd = sd(relabun[locality == "Disko"] * 100, na.rm = T),
    Illulissat_mean = mean(relabun[locality == "Illulissat"] * 100, na.rm = T),
    Illulissat_sd = sd(relabun[locality == "Illulissat"] * 100, na.rm = T),
    Ittoqqortoormiit_mean = mean(relabun[locality == "Ittoqqortoormiit"] * 100, na.rm = T),
    Ittoqqortoormiit_sd = sd(relabun[locality == "Ittoqqortoormiit"] * 100, na.rm = T)
  ) %>%
  mutate(
    Total = str_c(round(total_mean, 4), "±", round(total_sd, 3)),
    Daneborg = str_c(round(Daneborg_mean, 4), "±", round(Daneborg_sd, 3)),
    Disko = str_c(round(Disko_mean, 4), "±", round(Disko_sd, 3)),
    Illulissat = str_c(round(Illulissat_mean, 4), "±", round(Illulissat_sd, 3)),
    Ittoqqortoormiit = str_c(round(Ittoqqortoormiit_mean, 4), "±", round(Ittoqqortoormiit_sd, 3))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(phylum, Total, Daneborg, Disko, Illulissat, Ittoqqortoormiit) %>%
  rename(Phylum = phylum) %>%
  tt()
```

### Upset: shared MAGs
```{r chart1_sled, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
genome_row <- genome_counts_filt %>%
  column_to_rownames(., "genome")

sample_metadata_row <- sample_metadata %>%
  column_to_rownames(., "sample")

all(colnames(genome_row) == sample_metadata$sample)
common_samples <- intersect(colnames(genome_row), sample_metadata$sample)
genome_row_aligned <- genome_row[, common_samples, drop = FALSE] %>%
  rownames_to_column(., "genome")
sample_metadata_aligned <- sample_metadata_row[common_samples, , drop = FALSE]

genome_counts_rel <- genome_row_aligned %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  column_to_rownames(., "genome")

genome_counts_rel_pa = 1 * (genome_counts_rel > 0)
table_upset_analysis_cont = t(aggregate(
  t(genome_counts_rel_pa),
  by = list(sample_metadata_aligned$locality),
  FUN = sum
)[, -1])
colnames(table_upset_analysis_cont) = levels(as.factor(sample_metadata_aligned$locality))
table_upset_analysis = (table_upset_analysis_cont > 0) * 1
table_upset_analysis = data.frame(table_upset_analysis)
table_upset_analysis = apply(table_upset_analysis, 2, as.integer)
rownames(table_upset_analysis) <- rownames(genome_counts_rel_pa)


upset(
  as.data.frame(table_upset_analysis),
  keep.order = T,
  sets = rev(c("Daneborg", "Ittoqqortoormiit", "Illulissat", "Disko")),
  sets.bar.color = rev(locality_cols),
  mb.ratio = c(0.55, 0.45),
  order.by = "freq"
)
```

## Alpha diversity

```{r alpha_plots_sled, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
#  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))  %>%
#  mutate(locality=factor(locality,levels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island")))%>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>% 
  ggplot(aes(y = value, x = locality, group=locality, color=locality, fill=locality)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_color_manual(name="Locality",
          breaks=names(locality_cols),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=locality_cols)+
  scale_fill_manual(name="Locality",
          breaks=names(locality_cols),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=locality_cols) +
  facet_wrap(. ~factor(metric, labels=c("richness" = "Richness", "neutral" = "Neutral", "phylogenetic" = "Phylogenetic")), scales = "free", ncol = 5) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```

```{r alpha_values_season_sled, comment="", message=FALSE, warning=FALSE}
alpha_div %>%
  pivot_longer(-sample, names_to = "alpha", values_to = "value") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(alpha) %>%
  summarise(
    total_mean = mean(value, na.rm = T),
    total_sd = sd(value, na.rm = T),
    Daneborg_mean = mean(value[locality == "Daneborg"], na.rm =T),
    Daneborg_sd = sd(value[locality == "Daneborg"], na.rm =T),
    Disko_mean = mean(value[locality == "Disko"], na.rm = T),
    Disko_sd = sd(value[locality == "Disko"], na.rm =T),
    Illulissat_mean = mean(value[locality == "Illulissat"], na.rm = T),
    Illulissat_sd = sd(value[locality == "Illulissat"], na.rm = T),
    Ittoqqortoormiit_mean = mean(value[locality == "Ittoqqortoormiit"], na.rm = T),
    Ittoqqortoormiit_sd = sd(value[locality == "Ittoqqortoormiit"], na.rm = T)) %>%
  mutate(
    Total = str_c(round(total_mean, 2), "±", round(total_sd, 2)),
    "Daneborg-Sirius" = str_c(round(Daneborg_mean, 2), "±", round(Daneborg_sd, 2)),
    "Disko Island" = str_c(round(Disko_mean, 2), "±", round(Disko_sd, 2)),
    Illulissat = str_c(round(Illulissat_mean, 2), "±", round(Illulissat_sd, 2)),
    Ittoqqortoormiit = str_c(round(Ittoqqortoormiit_mean, 2), "±", round(Ittoqqortoormiit_sd, 2))
  ) %>%
  arrange(-total_mean) %>%
  dplyr::select(alpha,Total, "Daneborg-Sirius",Ittoqqortoormiit, Illulissat, "Disko Island") %>% 
  rename(Alpha=alpha) %>% 
  tt()
```

```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  kruskal_test(value ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))%>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  wilcox_test(value ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

Non-significant
```{r}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic")))%>%
  left_join(sample_metadata, by = "sample") %>%
  group_by(metric) %>%
  wilcox_test(value ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj > 0.05)
```


## Beta diversity

### Richness

**PERMANOVA**
```{r permanova_greenland_q0_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]
beta_q0n_dis <- as.dist(beta_q0n_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q0n_dis)) %>%
  arrange(match(sample, labels(beta_q0n_dis))) %>%
  column_to_rownames(., "sample")
            
all(rownames(sample_metadata_ordered) == labels(beta_q0n_dis))

betadisper(beta_q0n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n_dis ~ locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$locality, perm = 999)
```

**NMDS**
```{r beta_div_nmds_neutral_plot_G_sled, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
beta_q0n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = locality)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +labs(color='Locality')
```

### Neutral

**PERMANOVA**

```{r permanova_greenland_q1_sled, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)

beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]

beta_q1n_dis <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q1n_dis)) %>%
  arrange(match(sample, labels(beta_q1n_dis))) %>%
  column_to_rownames(., "sample")
            
all(rownames(sample_metadata_ordered) == labels(beta_q1n_dis))

betadisper(beta_q1n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n_dis ~ locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$locality, perm = 999)
```

**NMDS**
```{r beta_div_neutral_G_sled, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=12, fig.fullwidth=TRUE}
beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = locality)) +
    geom_point(size = 3) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    scale_color_manual(name="Locality",
          breaks=names(locality_cols),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=locality_cols)+
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Locality')
```

***Centroid distance based on the NMDS***
```{r}
betadisper_obj <- betadisper(beta_q1n_dis, sample_metadata_ordered$locality)
dist(betadisper_obj$centroids)
```

**PCoA**
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100
```

```{r pcoa_neutral, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
centroids <- pcoa_df %>%
  group_by(locality) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )

ggplot(pcoa_df, aes(Axis.1, Axis.2, color = locality)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(data = centroids,
             aes(Axis.1, Axis.2, fill = locality),
             color = "black",
             shape = 21,
             size = 6,
             show.legend = FALSE) +
    scale_color_manual(name="Locality",
          breaks=names(locality_cols),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=locality_cols)+
  scale_fill_manual(name="Locality",
          breaks=names(locality_cols),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=locality_cols) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Dog type"
  ) +
  theme_classic(base_size = 14)
```

***Centroid distances based on the PCoA***
```{r}
centroid_coords <- centroids %>%
  select(locality, Axis.1, Axis.2)

dist_mat <- dist(centroid_coords[, c("Axis.1", "Axis.2")]) %>% as.matrix()

dist_table <- as.data.frame(as.table(as.matrix(dist_mat))) %>%
  mutate(
    group1 = centroids$locality[Var1],
    group2 = centroids$locality[Var2],
    distance = Freq
  ) %>% 
  select(-Var1, -Var2, -Freq)
```

```{r centroid_distances_locality, comment="", message=FALSE, warning=FALSE, fig.height=6, fig.width=8, fig.fullwidth=TRUE}
ggplot(dist_table, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_blank()) +
  labs(fill="Centroids distance")
```


```{r permanova_greenland_qP_sled, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
### Phylogenetic


beta_q1p_mat <- as.matrix(beta_q1p$S)

beta_q1p_filtered <- beta_q1p_mat[rownames(beta_q1p_mat) %in% sample_metadata$sample,
                        colnames(beta_q1p_mat) %in% sample_metadata$sample]

beta_q1p_dis <- as.dist(beta_q1p_filtered)

sample_metadata_ordered <- sample_metadata %>%
  filter(sample %in% labels(beta_q1p_dis)) %>%
  arrange(match(sample, labels(beta_q1p_dis))) %>%
  column_to_rownames(., "sample")

betadisper(beta_q1p_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1p_dis ~ locality,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

pairwise.adonis(beta_q1p_dis, sample_metadata_ordered$locality, perm = 999)
```


```{r beta_div_nmds_phylo1_plot_G, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE, echo=FALSE, eval=FALSE}
beta_q1p_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  mutate(x_cen = median(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = median(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = locality)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    scale_color_manual(name="Locality",
          breaks=names(locality_cols),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=locality_cols)+
  scale_fill_manual(name="Locality",
          breaks=names(locality_cols),
          labels=c("Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island"),
          values=locality_cols) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 20, face = "bold"),
      axis.text = element_text(face = "bold", size = 18),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 16),
      legend.title = element_text(size = 18),
      legend.position = "right", legend.box = "vertical"
    ) +
    labs(color='Locality')
```

### Is there any location closer to domestic dogs?
```{r load_data_beta_dan, comment="", message=FALSE, warning=FALSE}
set.seed(1234)
load("resources/data.Rdata")
load("output/beta.Rdata")

remove_sled <- sample_metadata %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_class = case_when(
      breed == "Wolf" ~ "wolf",
      locality == "Disko" ~ "Disko",
      locality == "Daneborg" ~ "Daneborg",
      locality == "Illulissat" ~ "Illulissat",
      locality == "Ittoqqortoormiit" ~ "Ittoqqortoormiit",
      TRUE ~ "Companion"
    )
  ) %>%
  mutate(dog_group = case_when(locality == "Daneborg" ~ "military", TRUE ~ "civilian"))

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata_bacteria %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

**Neutral**
```{r permanova_greenland_q1_dan, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)

beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]

beta_q1n_distance <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1n_distance))) %>% 
  column_to_rownames(., "sample")
            
all(rownames(sample_metadata_ordered) == labels(beta_q1n_distance))

betadisper(beta_q1n_distance, sample_metadata_ordered$dog_class) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n_distance ~ dog_class, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

pairwise.adonis(beta_q1n_distance, sample_metadata_ordered$dog_class, perm = 999)
```

```{r}
betadisper_obj <- betadisper(beta_q1n_distance, sample_metadata_ordered$dog_class)
dist(betadisper_obj$centroids)
```

```{r beta_div_neutral_dan, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=14, fig.fullwidth=TRUE}
beta_q1n_distance %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_class) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_class)) +
    geom_point(size = 3) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    scale_color_manual(name="Dog group",
          breaks=c("Companion","Daneborg","Ittoqqortoormiit","Illulissat","Disko", "wolf"),
          labels=c("Companion","Daneborg-Sirius","Ittoqqortoormiit","Illulissat","Disko Island", "Wolf"),
          values=c('#41B6C0',"#f5c564","#c49c4b","#9c7c3b", "#614e27","#bb6749"))+
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    )
```


## Differential abundance
```{r phylo_sled, comment="", message=FALSE, warning=FALSE}
sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample) %>%
  filter(country == "Greenland")

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

phylo_samples <- sample_metadata %>%
  filter(country=="Greenland") %>% 
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>%
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)

phylo_taxonomy <- genome_metadata %>%
  filter(genome %in% rownames(phylo_genome)) %>%
  mutate(genome2 = genome) %>% 
  column_to_rownames("genome2") %>%
  dplyr::select(domain, phylum, class, order, family, genus, species, genome) %>% 
  as.matrix() %>%
  tax_table()

physeq_genome_green <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
```

### Differences across localities

#### MAG level
```{r load_all_sled, comment="", message=FALSE, warning=FALSE}
load("output/ancombc_sled_greenland.Rdata")
region_pairs_sled <- combn(unique(sample_metadata$locality), 2, simplify = FALSE)
```

```{r pair_all_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
results_sled_mag <- lapply(region_pairs_sled, function(pair) {
  subset_idx <- sample_metadata$locality %in% pair
  
  sled <-  ancombc2(
    data = prune_samples(subset_idx, physeq_genome_green),
    fix_formula = "locality",
    group = "locality",
    p_adj_method = "BH",
    pseudo_sens = TRUE,
    prv_cut = 0.05,
    lib_cut = 1000,
    s0_perc = 0.05,
    struc_zero = TRUE,
    neg_lb = FALSE,
    alpha = 0.05,
    n_cl = 2,
    verbose = TRUE,
    global = FALSE,
    pairwise = FALSE,
    dunnet = FALSE,
    trend = FALSE,
    iter_control = list(
      tol = 1e-5,
      max_iter = 50,
      verbose = FALSE
    ),
    em_control = list(tol = 1e-5, max_iter = 100),
    mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
    trend_control = NULL
  )
  sled_table <- sled$res
  
  sled_table$comparison <- paste(pair, collapse = "_vs_")
  
  return(sled_table)
})

names(results_sled_mag) <- sapply(region_pairs_sled, paste, collapse = "_vs_")

save(results_sled_mag, file = "output/ancombc_sled_greenland.Rdata")
```

```{r extract_sled, comment="", message=FALSE, warning=FALSE}
extract_pair <- function(res, pair) {

  tab <- res
  if (is.null(tab) || nrow(tab) == 0) return(NULL)

  # Identify country coefficient column
  coef_cols <- grep("^lfc_locality", colnames(tab), value = TRUE)
  coef_cols <- coef_cols[!grepl("\\(Intercept\\)", coef_cols)]
  if (length(coef_cols) == 0) return(NULL)

  coef <- coef_cols[1]  # only one since pairwise subset
  country_nonref <- sub("^lfc_locality", "", coef)
  country_ref <- setdiff(pair, country_nonref)

  tibble(
    taxon       = tab$taxon,
    model_term  = coef,
    non_reference = country_nonref,
    reference     = country_ref,
    logFC       = tab[[coef]],
    q_value     = tab[[sub("^lfc_", "q_", coef)]],
    diff_abn    = tab[[sub("^lfc_", "diff_", coef)]],
    enriched_in = ifelse(tab[[coef]] > 0, country_nonref, country_ref),
    depleted_in = ifelse(tab[[coef]] > 0, country_ref, country_nonref),
    comparison  = paste(pair, collapse = "_vs_")
  )
}
pairwise_tables_greenland <- map2(
  results_sled_mag,
  region_pairs_sled,
  extract_pair
)
pairwise_tables_greenland <- discard(pairwise_tables_greenland, is.null)

final_table_tables_greenland <- bind_rows(pairwise_tables_greenland)
nrow(final_table_tables_greenland)
table(final_table_tables_greenland$comparison)

sig_taxa_tables_greenland <- final_table_tables_greenland %>%
  filter(q_value < 0.05) %>% 
  left_join(genome_metadata %>% select(genome, phylum), by=join_by("taxon"=="genome"))
```

```{r loop_plots_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
taxonomy <- data.frame(physeq_genome_green@tax_table) %>%
  distinct(phylum) %>% 
  rownames_to_column(., "taxon")

unique_contrasts <- unique(sig_taxa_tables_greenland$comparison)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (c in unique_contrasts) {

  subset_data <- sig_taxa_tables_greenland %>%
    filter(comparison == c) %>%
    arrange(enriched_in, logFC) %>%
    mutate(taxon_ordered = factor(taxon, levels = unique(taxon)))
  
  non_ref <- unique(subset_data$non_reference)
  ref <- unique(subset_data$reference)
  title <- paste0("Enriched in ", ref, " (− LFC) vs  ", "Enriched in ", non_ref, " (+ LFC)")

  tax_color <- as.data.frame(unique(subset_data$phylum)) %>%
    rename(phylum = 1) %>%
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()
  
  p <- ggplot(subset_data, aes(x = taxon_ordered, y = logFC, fill = phylum)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 6),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    scale_fill_manual(values = tax_color) +
    labs(
      title = title,
      x = "MAGs",
      y = "Log-Fold Change",
      fill = "Phylum"
    )
  
  ggsave(paste0("output/figures/Fig_", gsub(" ", "_", c), ".png"), plot = p, width = 10, height = 25, dpi = 300)
}
```

**Daneborg-Sirius vs Disko Island**

***Enriched in Daneborg-Sirius***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC < 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC < 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

***Enriched in Disko Island***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC > 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Disko") %>%
  filter(logFC > 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

**Daneborg-Sirius vs Illulissat**

***Enriched in Daneborg-Sirius***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC < 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC < 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

***Enriched in Illulissat***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC > 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Illulissat") %>%
  filter(logFC > 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

**Daneborg-Sirius vs Ittoqqortoormiit**

***Enriched in Daneborg-Sirius***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC < 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC < 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

***Enriched in Ittoqqortoormiit***
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC > 0) %>%
  nrow()
```
```{r}
sig_taxa_tables_greenland %>%
  left_join(genome_metadata %>% select(-phylum), by = join_by("taxon" == "genome")) %>%
  filter(comparison == "Daneborg_vs_Ittoqqortoormiit") %>%
  filter(logFC > 0) %>%
  group_by(phylum, family, genus, species) %>%
  summarise(Mags = n(), .groups = "drop") %>%
  arrange(-Mags)
```

### Differences between military and civilian-owned dogs

#### MAG level
```{r ancom_phylum_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
ancom_rand_output = ancombc2(
  data = physeq_genome_green,
  assay_name = "counts",
  fix_formula = "dog_group",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "dog_group",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 50,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
  trend_control = NULL
)

save(ancom_rand_output, 
     file = "output/ancombc_daneborg_others.Rdata")
```

```{r ancom_rand_sled, comment="", message=FALSE, warning=FALSE}
load("output/ancombc_daneborg_others.Rdata")
taxonomy <- data.frame(physeq_genome_green@tax_table) %>%
  rownames_to_column(., "taxon")

ancombc_rand_table_mag <- ancom_rand_output$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  dplyr::arrange(q_dog_groupmilitary) %>%
  merge(., taxonomy, by="taxon") %>%
  dplyr::arrange(lfc_dog_groupmilitary)
  
colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>%
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```

```{r ancombc_rand_plot_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=8, fig.fullwidth=TRUE}
ancombc_rand_table_mag %>%
  mutate(genome=factor(taxon,levels=ancombc_rand_table_mag$taxon)) %>%
  ggplot(., aes(x=lfc_dog_groupmilitary, y=forcats::fct_reorder(genome,lfc_dog_groupmilitary), fill=phylum)) + #forcats::fct_rev()
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  geom_text(aes(3, 445), label = "Enriched in\n military dogs", color="#666666") +
  geom_text(aes(-5, 145), label = "Enriched in\n civilian-owned\n dogs", color="#666666") +
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Genome")+
  guides(fill=guide_legend(title="Phylum"))
```

```{r dif_sled_others_genus, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
genus_matrix <- ancombc_rand_table_mag%>%
      mutate(taxon=factor(taxon,levels=ancombc_rand_table_mag$taxon)) %>%
      filter(!is.na(phylum))

genus_matrix <- genus_matrix %>%
  arrange(desc(-lfc_dog_groupmilitary)) %>%
  slice_head(n = 30) %>%
  bind_rows(genus_matrix %>%
              arrange(-lfc_dog_groupmilitary) %>%
              slice_head(n = 30)) 

tax_table <- as.data.frame(unique(genus_matrix$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by = "phylum") %>%
  dplyr::arrange(phylum) %>%
  dplyr::select(colors) %>%
  pull()

genus_matrix %>%
  ggplot(., aes(x=lfc_dog_groupmilitary, y=forcats::fct_reorder(genome,lfc_dog_groupmilitary), fill=phylum)) + 
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  geom_text(aes(4.3, 35), label = "Enriched in\n military dogs", color="#666666") +
  geom_text(aes(-5.6, 20), label = "Enriched in\n civilian-owned\n dogs", color="#666666") +
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Genome") +
  guides(fill=guide_legend(title="Phylum"))      
```

***Military dogs***
```{r mil, comment="", message=FALSE, warning=FALSE}
ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  arrange(desc(species)) %>% 
  select(species) %>% 
  nrow %>% 
  cat()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  group_by(phylum) %>% 
  summarise(count=n()) %>% 
  arrange(-count) %>% 
  tt()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary>0) %>% 
  group_by(phylum, family, genus, species) %>% 
  summarise(count=n()) %>% 
  arrange(-count) %>% 
  head(10) %>% 
  tt()
```

***Civilian-owned dogs***
```{r civ, comment="", message=FALSE, warning=FALSE}
ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  arrange(desc(species)) %>% 
  select(species) %>% 
  nrow()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  group_by(phylum) %>% 
  summarise(count=n()) %>% 
  arrange(-count)%>% 
  tt()

ancombc_rand_table_mag %>% 
  filter(lfc_dog_groupmilitary<0) %>% 
  group_by(phylum, family, genus, species) %>% 
  summarise(count=n()) %>% 
  arrange(-count)%>% 
  head(10) %>% 
  tt()
```

#### Species level
```{r}
genome_tax <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order), if_else(
      str_detect(class, "Unclassified"),
      class,
      paste0(class, "_Unclassified")
    ), order),
    family = if_else(is.na(family), if_else(
      str_detect(order, "Unclassified"),
      order,
      paste0(order, "_Unclassified")
    ), family),
    genus  = if_else(genus == "", if_else(
      str_detect(family, "Unclassified"),
      family,
      paste0(family, "_Unclassified")
    ), genus),
    species  = if_else(species == "", if_else(
      str_detect(genus, "Unclassified"),
      genus,
      paste0(genus, "_Unclassified")
    ), species)
  )
```

```{r}
phylo_samples <- sample_metadata %>%
  filter(country=="Greenland") %>% 
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  left_join(genome_tax[c(1,3,4,6,8)], by = "genome") %>% 
  select(-genome) %>%
  group_by(phylum, class, family, species) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class, family))%>%
  column_to_rownames("species") %>%
  otu_table(., taxa_are_rows = TRUE)

physeq_genome_sp <- phyloseq(phylo_genome, phylo_samples)

load("output/ancombc_danenborg_greenland_sp.Rdata")
```

```{r ancom_sp_sled, comment="", message=FALSE, warning=FALSE, eval=FALSE}
ancom_rand_output_sp = ancombc2(
  data = physeq_genome_sp,
  assay_name = "counts",
  fix_formula = "dog_group",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "dog_group",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 50,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
  trend_control = NULL
)

save(ancom_rand_output_sp, 
     file = "output/ancombc_danenborg_greenland_sp.Rdata")
```

```{r}
ancom_rand_output_sp$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  filter(lfc_dog_groupmilitary > 0) %>%
  nrow() %>% 
  cat()

ancom_rand_output_sp$res %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  filter(lfc_dog_groupmilitary > 0) %>%
  arrange(-lfc_dog_groupmilitary) %>% 
  head(10) %>% 
  tt()
```

```{r}
ancom_rand_output_sp$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  filter(lfc_dog_groupmilitary < 0) %>%
  nrow() %>% 
  cat()

ancom_rand_output_sp$res %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  filter(lfc_dog_groupmilitary < 0) %>%
  arrange(lfc_dog_groupmilitary)%>% 
  head(10) %>% 
  tt()
```
```{r dif_sled_others_sp, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=9, fig.fullwidth=TRUE}
ancombc_rand_table_sp <- ancom_rand_output_sp$res %>%
  dplyr::select(taxon, lfc_dog_groupmilitary, q_dog_groupmilitary) %>%
  filter(q_dog_groupmilitary < 0.05) %>%
  dplyr::arrange(q_dog_groupmilitary) %>%
  mutate(taxon = str_remove(taxon, "_Unclassified")) %>%
  left_join(., taxonomy[c(3,8)] %>% distinct(species, .keep_all = TRUE), by=join_by("taxon"=="species")) %>%
  left_join(., taxonomy[c(3,7)] %>% distinct(genus, .keep_all = TRUE), by=join_by("taxon"=="genus")) %>%
  left_join(., taxonomy[c(3,6)] %>% distinct(family, .keep_all = TRUE), by=join_by("taxon"=="family")) %>%
  mutate(phylum = coalesce(phylum.x, phylum.y, phylum)) %>%
  select(-phylum.x, -phylum.y)%>%
  dplyr::arrange(lfc_dog_groupmilitary)

genus_matrix <- ancombc_rand_table_sp %>%
      mutate(taxon=factor(taxon,levels=ancombc_rand_table_sp$taxon)) %>%
      filter(!is.na(phylum))

genus_matrix <- genus_matrix %>%
  arrange(desc(-lfc_dog_groupmilitary)) %>%
  slice_head(n = 30) %>%
  bind_rows(genus_matrix %>%
              arrange(-lfc_dog_groupmilitary) %>%
              slice_head(n = 30)) 

tax_table <- as.data.frame(unique(genus_matrix$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by = "phylum") %>%
  dplyr::arrange(phylum) %>%
  dplyr::select(colors) %>%
  pull()

genus_matrix %>%
  ggplot(., aes(x=lfc_dog_groupmilitary, y=forcats::fct_reorder(taxon,lfc_dog_groupmilitary), fill=phylum)) + 
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  geom_text(aes(4.3, 45), label = "Enriched in\n military dogs", color="#666666") +
  geom_text(aes(-4.6, 25), label = "Enriched in\n civilian-owned\n dogs", color="#666666") +
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  xlab("log2FoldChange") + 
  ylab("Species")+
  guides(fill=guide_legend(title="Phylum"))      
```

## Functional differences

```{r gift_analysis_func_sled, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts),]
genome_counts_filt <- genome_counts_filt %>%
  select(genome, all_of(sample_metadata$sample)) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))%>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome,]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]

GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_row <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>% 
  column_to_rownames(., "genome") 

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)
```

### Different localities in Greenland

#### MCI

```{r}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(locality) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  mutate(MCI = str_c(round(MCI, 3), "±", round(sd, 3))) %>% 
  select(-sd) %>% 
  tt()
```

```{r}
MCI_pf <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(residuals(lm(value ~ locality, data = MCI_pf))) 
bartlett.test(value ~ locality, data = MCI_pf)  

kruskal.test(value ~ locality, data = MCI_pf)
pw <- pairwise.wilcox.test(
  x = MCI_pf$value,
  g = MCI_pf$locality,
  p.adjust.method = "BH"
)

wilcox_tbl <- pw$p.value %>%
  as.data.frame() %>%
  rownames_to_column("group1") %>%
  pivot_longer(
    -group1,
    names_to = "group2",
    values_to = "p_adjust"
  ) %>%
  filter(!is.na(p_adjust)) %>%
  mutate(
    comparison = paste(group1, "vs", group2),
    p_adjust = round(p_adjust, 3),
    sig = case_when(
      p_adjust < 0.001 ~ "***",
      p_adjust < 0.01  ~ "**",
      p_adjust < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(comparison, p_adjust, sig)

wilcox_tbl %>%
  rename("Compared countries" = comparison,
         "Adjusted p-value" = p_adjust) %>%
  tt()
```

#### Functional pathways

```{r commun_wilcox_elem_sled, comment="", message=FALSE, warning=FALSE}
element_long <- GIFTs_elements_community %>% 
  as.data.frame() %>%
  rownames_to_column(., "sample") %>% 
  filter(sample %in% sample_metadata$sample) %>% 
  left_join(sample_metadata %>% select(sample,locality), by="sample") %>%
  pivot_longer(-c(sample,locality), names_to = "trait", values_to = "value")

uniqueGIFT_db <- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

significant_traits <- element_long %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ locality)$p.value) %>%
  mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
  filter(p_adjust < 0.05) %>%
  pull(trait)

effect_sizes <- element_long %>%
  group_by(trait) %>%
  kruskal_effsize(value ~ locality)

kw_results <- element_long %>%
  group_by(trait) %>%
  summarise(p_value = kruskal.test(value ~ locality)$p.value) %>%
  mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
  filter(p_adjust < 0.05) %>%
  left_join(effect_sizes, by = "trait")

pairwise_results <- element_long %>%
  filter(trait %in% significant_traits) %>%
  group_by(trait) %>%
  dunn_test(value ~ locality, p.adjust.method = "BH") %>%
  filter(p.adj<0.05) %>%
  mutate(comparison = paste(group1, "vs", group2)) %>%
  left_join(uniqueGIFT_db, by = join_by(trait == Code_element))
```

```{r heatmap_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
trait_summary <- element_long %>%
  group_by(trait, locality) %>%
  summarise(median_value = median(value, na.rm = TRUE), .groups = "drop")
trait_summary_sig <- trait_summary %>%
  filter(trait %in% significant_traits)

trait_order <- kw_results %>%
  arrange(desc(effsize)) %>%
  pull(trait)

trait_summary_sig$trait <- factor(trait_summary_sig$trait, levels = trait_order)

trait_summary_sig <- trait_summary_sig %>%
  group_by(trait) %>%
  mutate(z_value = scale(median_value)[,1]) %>%
  ungroup() %>% 
  left_join(uniqueGIFT_db, by = join_by(trait == Code_element))

ggplot(trait_summary_sig,
       aes(locality, Function, fill = z_value)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    name = "Relative capacity"
  ) +
  theme_minimal()
```
```{r heatmap2_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=15, fig.fullwidth=TRUE}
heatmap_matrix <- element_long %>%
  filter(trait %in% significant_traits) %>%
  group_by(trait, locality) %>%
  summarise(median_value = median(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = locality, values_from = median_value) %>% 
  left_join(uniqueGIFT_db %>% select(Code_element,Function), by = join_by(trait == Code_element)) %>% 
  column_to_rownames(., "Function") %>% 
  select(-trait) %>% 
  as.matrix()

heatmap_scaled <- t(scale(t(heatmap_matrix)))

sig_annotation <- kw_results %>% 
  left_join(uniqueGIFT_db %>% select(Code_element,Function), by = join_by(trait == Code_element)) %>% 
  select(-trait) %>% 
  filter(Function %in% rownames(heatmap_matrix)) %>%
  select(Function, effsize)%>%
  column_to_rownames("Function") %>%
  as.data.frame()

pheatmap::pheatmap(
  heatmap_scaled,
  annotation_row = sig_annotation,
  clustering_method = "complete",
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  scale = "none",
  fontsize_row = 12,
  fontsize_col = 12,
  border_color = NA
)
```

```{r function_plot_sled, comment="", message=FALSE, warning=FALSE, fig.height=20, fig.width=10, fig.fullwidth=TRUE}
ggplot(pairwise_results, aes(x = comparison, y = fct_rev(Function))) +
  geom_tile(aes(fill = p.adj), color = "white") +
  geom_text(aes(label = p.adj.signif), size = 4, color = "black") +
  scale_fill_gradient(low = "#b30000", high = "#fef0d9" , name = "Adjusted p-value") +
  theme_minimal() +
  labs(
    title = "Significant Pairwise Comparisons by Trait",
    x = "Locality Comparison",
    y = "Trait"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

```{r barplot_func_sled_locality, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>%
  rownames_to_column(., "sample") %>% 
  filter(sample %in% sample_metadata$sample) %>% 
  left_join(sample_metadata %>% select(sample,locality), by="sample")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, locality), names_to = "trait", values_to = "value") %>%
  filter(trait %in% pairwise_results$trait)%>% 
  left_join(., uniqueGIFT_db, by=join_by(trait==Code_element)) 

ggplot(plot_data, aes(x = locality, y = value, fill = locality)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~ Function, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 16) +
  theme(strip.text = element_text(size = 25, lineheight = 0.6))+
  labs(
    title = "Trait Distributions Across Localities",
    x = "Locality",
    y = "Trait Value" )
```

### Military vs civilian-owned dogs
```{r}
sample_metadata <- sample_metadata %>% 
  mutate(
    dog_group = case_when(
      locality == "Daneborg" ~ "military",
      TRUE                   ~ "civilian"
    )
  )
```

#### MCI

```{r gitfs_functional_group,results=TRUE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_group) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  tt()
```

```{r}
MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) 

shapiro.test(residuals(lm(value ~ dog_group, data = MCI)))
bartlett.test(value ~ dog_group, data = MCI)

wilcox.test(value ~ dog_group, data = MCI)
```

#### Functional pathways

```{r}
enrichment_results <- GIFTs_elements_community %>%
  as.data.frame() %>%
  rownames_to_column(., "sample") %>%
  filter(sample %in% sample_metadata$sample) %>%
  left_join(sample_metadata %>% select(sample,dog_group), by="sample")%>%
  pivot_longer(-c(sample,dog_group),
               names_to="trait",
               values_to="value") %>%
  group_by(trait) %>%
  summarise(
    p_value = wilcox.test(value ~ dog_group)$p.value,
    median_military = median(value[dog_group=="military"]),
    median_civilian = median(value[dog_group=="civilian"]),
    log2FC = log2((median_military + 1) /
                  (median_civilian + 1)),
    difference = median_military - median_civilian) %>%
  mutate(p_adjust = p.adjust(p_value,"BH")) %>% 
  filter(p_adjust<0.05) 
```

```{r plot3_sled, comment="", message=FALSE, warning=FALSE, fig.height=5, fig.width=10, fig.fullwidth=TRUE}
uniqueGIFT_db<- unique(GIFT_db[c(2,3,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

enrichment_code <- enrichment_results %>%
  left_join(uniqueGIFT_db[c(1:4)], by=join_by(trait==Code_element))

unique_codes<-unique(enrichment_code$Code_function)

gift_colors <- read_tsv("resources/gift_colors.tsv") %>% 
  filter(Code_function %in% unique_codes)%>% 
  mutate(legend=str_c(Code_function," - ",Function))
gift_color_values <- setNames(gift_colors$Color, gift_colors$legend)

```

```{r final_plot_season, comment="", message=FALSE, warning=FALSE, fig.height=12, fig.width=12, fig.fullwidth=TRUE}
enrichment_code %>%
  left_join(
    gift_colors %>% select(Code_function, Color, legend),
    by = join_by(Code_function == Code_function)
  ) %>%
  ggplot(aes(x = forcats::fct_reorder(Function, log2FC), y = log2FC, fill = legend)) +
  geom_col() +
  scale_fill_manual(values = gift_color_values) +
  geom_hline(yintercept = 0) +
  coord_flip() +
  geom_text(aes(85, 0.060), label = "Enriched in\n Sirius dogs", color="#666666") +
  geom_text(aes(31, -0.050), label = "Enriched in\n civilian dogs", color="#666666") +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 12),
        legend.position = "right", 
        panel.background = element_blank())+
  xlab("Microbial Functional Capacity") + 
  ylab("log2FC")
```

**Effect size: log2FC**

***Military***
```{r}
enrichment_code %>% 
  left_join(unique(GIFT_db[c(2,4,5,6)]), by=join_by("trait"=="Code_element")) %>% 
  filter(log2FC>0) %>% 
  group_by(Function.y) %>% 
  summarise(count=n()) %>% 
  arrange(-count)
```
```{r}
enrichment_code %>% 
  filter(log2FC>0) %>% 
    arrange(Function) %>% 
  select(Function) %>% 
  pull()
```

***Civilian***
```{r}
enrichment_code %>% 
  left_join(unique(GIFT_db[c(2,4,5,6)]), by=join_by("trait"=="Code_element")) %>% 
  filter(log2FC<0) %>% 
  group_by(Function.y) %>% 
  summarise(count=n())%>% 
  arrange(-count)
```

```{r}
enrichment_code %>% 
  filter(log2FC<0) %>% 
  arrange(Function) %>% 
  select(Function) %>% 
  pull()
```

```{r barplot_func_sled, comment="", message=FALSE, warning=FALSE, fig.height=60, fig.width=30, fig.fullwidth=TRUE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>%
  rownames_to_column(., "sample") %>% 
  filter(sample %in% sample_metadata$sample) %>% 
  left_join(sample_metadata %>% select(sample,dog_group), by="sample")

plot_data <- element_gift %>%
  pivot_longer(-c(sample, dog_group), names_to = "trait", values_to = "value") %>%
  filter(trait %in% enrichment_code$trait)%>% 
  left_join(., uniqueGIFT_db, by=join_by(trait==Code_element)) 

ggplot(plot_data, aes(x = dog_group, y = value, fill = dog_group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  facet_wrap(~ Function, scales = "free_y", ncol = 4) +
  theme_minimal(base_size = 16) +
  labs(
    title = "Trait Distributions Across Localities",
    x = "Locality",
    y = "Trait Value" )
```

<!--chapter:end:05_sled_localities.Rmd-->

# Lower virulence and antimicrobial resistance of Qimmit microbiomes

```{r load_data_amr, comment="", message=FALSE, warning=FALSE}
load("resources/data.Rdata")
load("output/alpha_all.RData")

type_colors <- c('#41B6C0','#E5D388',"#bb6749")
new_labels <- c(
    "companion" = "Companion dogs",
    "sled_dog" = "Sled dogs",
    "wolf" = "Wolves"
    )

management_colors <- c('#41B6C0','#E5D388',"#f5c564","#bb6749")
management_labels <- c(
    "companion" = "Companion dogs",
    "civilian" = "Civilian dogs",
    "military" = "Sirius dogs",
    "wolf" = "Wolves"
    )

locality_colors <- c("#f5c564","#c49c4b","#9c7c3b", "#614e27")
locality_labels <- c(
    "Daneborg" = "Sirius dogs",
    "Disko" = "Disko Island",
    "Illulissat" = "Illulissat",
    "Ittoqqortoormiit" = "Ittoqqortoormiit"
    )

remove_sled <- read_csv("resources/sample_metadata.csv", show_col_types = FALSE) %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_group = case_when(
      dog_type == "wolf" ~ "wolf",
      locality == "Disko" ~ "civilian",
      locality == "Daneborg" ~ "military",
      locality == "Illulissat" ~ "civilian",
      locality == "Ittoqqortoormiit" ~ "civilian",
      TRUE                   ~ "companion"
    )
  )

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")
genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

```{r load_genome_anno, warning=FALSE, comment="", message=FALSE}
genome_annotations <- read_tsv(
  "resources/gene_annotations.tsv.xz",
  col_types = cols(.default = col_character(), start = col_double(), end = col_double())
) %>%
  mutate(genome = str_extract(gene, "^[^^]+")) %>%
  mutate(ec = if_else(is.na(ec), NA_character_, str_c("[EC:", ec, "]"))) %>% 
  select(genome, everything())
```

## Genome length

```{r size_per_sample, warning=FALSE, comment="", message=FALSE}
sample_genome_size <- genome_counts_filt %>% 
  mutate(across(-genome,~ .x / sum(.x, na.rm = TRUE))) %>% 
  inner_join(genome_metadata %>% select(genome,length), by = "genome") %>%
  pivot_longer(
    cols = -c(genome, length),
    names_to  = "sample",
    values_to = "rel_abundance"
  ) %>%
  group_by(sample) %>%
  summarise(
    size = weighted.mean(length, w = rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(sample_metadata,by="sample")
```

### Canids
```{r}
sample_genome_size %>% 
  group_by(dog_type) %>%
  summarise(
    mean = mean(size, na.rm = T)/1000000,
    sd = sd(size, na.rm = T)/1000000) %>% 
  arrange(-mean) %>% 
  mutate(length=str_c(round(mean,3),"±",round(sd,3))) %>% 
  select(dog_type,length) %>% 
  rename(Canid=1, "Genome length"=2) %>% 
  tt()
```

```{r}
sample_genome_size %>%
  kruskal_test(size ~ dog_type) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

```{r}
sample_genome_size %>%
  wilcox_test(size ~ dog_type, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)%>% 
  tt()
```

### Management groups
```{r}
sample_genome_size %>% 
  group_by(dog_group) %>%
  summarise(
    mean = mean(size, na.rm = T)/1000000,
    sd = sd(size, na.rm = T)/1000000) %>% 
  arrange(-mean) %>% 
  mutate(genome_size = str_c(round(mean, 3), " ± ", round(sd, 3))) %>%
  select(dog_group, genome_size) %>%
  rename(Group = 1, "Genome length" = 2)
```

```{r}
sample_genome_size %>%
  kruskal_test(size ~ dog_group) %>%
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>%
  tt()
```

```{r}
sample_genome_size %>%
  wilcox_test(size ~ dog_group, p.adjust.method = "BH") %>%
  filter(p.adj < 0.05)%>% 
  tt()
```

```{r}
sample_genome_size %>%
  wilcox_test(size ~ dog_group, p.adjust.method = "BH") %>%
  filter(p.adj > 0.05)%>% 
  tt()
```

```{r size_per_sample_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_genome_size %>%
  ggplot(aes(x = dog_type, y = size, color = dog_type, fill = dog_type)) +
  geom_jitter(width = 0.2, alpha = 0.7, show.legend = FALSE, size = 0.5) +
  geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = type_colors) +
  scale_fill_manual(values = type_colors) +
  scale_x_discrete(labels = new_labels) +
  theme_classic()+
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size =12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Genome length")
```
```{r genome_size_per_doggroup_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_genome_size %>% 
  mutate(dog_group=factor(dog_group,levels=c("companion","civilian","military", "wolf"))) %>%
  ggplot(aes(x = dog_group, y = size, color = dog_group, fill = dog_group)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = management_colors) +
  scale_fill_manual(values = management_colors) +
  scale_x_discrete(labels = management_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Genome length")
```
```{r size_per_sled_loc_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_genome_size %>% 
  filter(dog_type=="sled_dog") %>% 
  ggplot(aes(x=locality,y=size,color=locality,fill=locality)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 7),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 9, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Genome length")
```

```{r}
sample_genome_size %>%
  filter(dog_type=="sled_dog") %>% 
  kruskal_test(size ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
sample_genome_size %>%
  filter(dog_type=="sled_dog") %>% 
  wilcox_test(size ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

## Antimicrobial resistance (AMR)

```{r amr_per_genome, warning=FALSE, comment="", message=FALSE}
genome_amr <- genome_annotations %>% 
  group_by(genome) %>%
  summarise(AMR = sum(resistance_type == "AMR", na.rm = TRUE), .groups = "drop")
```

```{r amr_per_sample, warning=FALSE, comment="", message=FALSE}
sample_amr <- genome_counts_filt %>%
  mutate(across(-genome, ~ .x / sum(.x, na.rm = TRUE))) %>%
  inner_join(genome_amr, by = "genome") %>%
  pivot_longer(cols = -c(genome, AMR), names_to  = "sample", values_to = "rel_abundance") %>%
  group_by(sample) %>%
  summarise(AMR = weighted.mean(AMR, w = rel_abundance, na.rm = TRUE), .groups = "drop") %>%
  left_join(sample_metadata, by = "sample")
```

**Top 20 MAGs with highest values**
```{r}
genome_amr %>%
  left_join(genome_metadata, by = "genome") %>%
  arrange(-AMR) %>%
  head(20) %>%
  tt()
```

### Canids
```{r}
sample_amr %>%
  group_by(dog_type) %>%
  summarise(mean = mean(AMR, na.rm = T), sd = sd(AMR, na.rm = T)) %>%
  arrange(-mean) %>%
  mutate(AMR = str_c(round(mean, 3), " ± ", round(sd, 3))) %>%
  select(dog_type, AMR) %>% 
  rename(Canid=1)%>% 
  tt()
```

```{r amr_per_group_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_amr %>% 
  ggplot(aes(x = dog_type, y = AMR, color = dog_type, fill = dog_type)) +
  geom_jitter(width = 0.2, alpha = 0.7, show.legend = FALSE, size = 0.5) +
  geom_boxplot(width = 0.5, alpha = 0.5, outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = type_colors) +
  scale_fill_manual(values = type_colors) +
  scale_x_discrete(labels = new_labels)+
  theme_classic() +
  theme(
    axis.text = element_text(size = 10), 
    strip.background = element_blank(), 
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
  ) +
  labs(y = "AMR")
```

```{r}
sample_amr %>%
  kruskal_test(AMR ~ dog_type) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

```{r}
sample_amr %>%
  wilcox_test(AMR ~ dog_type, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05) %>% 
  tt()
```

### Management groups
```{r}
sample_amr %>% 
  group_by(dog_group) %>%
  summarise(
    mean = mean(AMR, na.rm = T),
    sd = sd(AMR, na.rm = T)) %>% 
  arrange(-mean) %>% 
  mutate(AMR=str_c(round(mean,3)," ± ",round(sd,3))) %>% 
  select(dog_group, AMR) %>%
  rename(Group = 1)
```
```{r}
sample_amr %>% 
  filter(country=="Greenland") %>%
  wilcox_test(AMR ~ dog_group) %>% 
  tt()
```

```{r}
sample_amr %>%
  kruskal_test(AMR ~ dog_group) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
sample_amr %>%
  wilcox_test(AMR ~ dog_group, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05) %>% 
  tt()
```

```{r amr_per_doggroup_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_amr %>% 
  mutate(dog_group=factor(dog_group,levels=c("companion","civilian","military", "wolf"))) %>%
  ggplot(aes(x=dog_group, y=AMR, color=dog_group, fill=dog_group)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = management_colors) +
  scale_fill_manual(values = management_colors) +
  scale_x_discrete(labels = management_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="AMR")
```

```{r amr_per_sled_loc_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_amr %>% 
  filter(dog_type=="sled_dog") %>% 
  ggplot(aes(x=locality,y=AMR,color=locality,fill=locality)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 7),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 9, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="AMR")
```

```{r}
sample_amr %>%
  filter(dog_type=="sled_dog") %>% 
  kruskal_test(AMR ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  filter(p.adj < 0.05) %>% 
  tt()
```

Significant
```{r}
sample_amr %>%
  filter(dog_type=="sled_dog") %>% 
  wilcox_test(AMR ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

Non-sig
```{r}
sample_amr %>%
      filter(dog_type=="sled_dog") %>% 
  wilcox_test(AMR ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj > 0.05)
```

Max AMR value in sled dogs
```{r}
sample_amr %>% 
  filter(country=="Greenland") %>% 
  arrange(-AMR)
```

Max AMR value in Sirius dogs
```{r}
sample_amr %>% 
  filter(locality=="Daneborg") %>% 
  arrange(-AMR)
```


## Virulence factors

```{r vf_per_genome, warning=FALSE, comment="", message=FALSE}
genome_vf <- genome_annotations %>% 
  group_by(genome) %>%
  summarise(vf = sum(!is.na(vf_type), na.rm = TRUE), .groups = "drop")

sample_vf <- genome_counts_filt %>% 
  mutate(across(-genome,~ .x / sum(.x, na.rm = TRUE))) %>% 
  inner_join(genome_vf, by = "genome") %>%
  pivot_longer(cols = -c(genome, vf), names_to  = "sample", values_to = "rel_abundance") %>%
  group_by(sample) %>%
  summarise(
    vf = weighted.mean(vf, w = rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(sample_metadata, by="sample")
```

**Top 20 MAGs with highest values**
```{r}
genome_vf %>% 
  left_join(genome_metadata, by="genome") %>% 
  arrange(-vf) %>% 
  head(20) %>% 
  tt()
```


### Canids
```{r}
sample_vf %>% 
  group_by(dog_type) %>%
  summarise(
    mean = mean(vf, na.rm = T),
    sd = sd(vf, na.rm = T)) %>% 
  arrange(-mean) %>% 
  mutate(virulence=str_c(round(mean,3),"±",round(sd,3))) %>% 
  select(dog_type,virulence) %>%
  rename("Canid" = 1, "Virulence factor" = 2)%>% 
  tt()
```

```{r vf_per_group_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
  ggplot(aes(x=dog_type,y=vf,color=dog_type,fill=dog_type)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = type_colors) +
  scale_fill_manual(values = type_colors) +
  scale_x_discrete(labels = new_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factor")
```

```{r}
sample_vf %>%
  kruskal_test(vf ~ dog_type) %>% 
  tt()
```

### Management: Civilian & Military
```{r}
sample_vf %>% 
  group_by(dog_group) %>%
  summarise(
    mean = mean(vf, na.rm = T),
    sd = sd(vf, na.rm = T)) %>% 
  arrange(-mean) %>% 
  mutate(virulence=str_c(round(mean,3),"±",round(sd,3))) %>% 
  select(dog_group,virulence) %>%
  rename("Canid" = 1, "Virulence factor" = 2)%>% 
  tt()
```
```{r vf_per_military_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
#  filter(country=="Greenland") %>% 
  mutate(dog_group=factor(dog_group,levels=c("companion","civilian","military", "wolf"))) %>%
  ggplot(aes(x=dog_group,y=vf,color=dog_group,fill=dog_group)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = management_colors) +
  scale_fill_manual(values = management_colors) +
  scale_x_discrete(labels = management_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factor")
```
```{r}
sample_vf %>%
  kruskal_test(vf ~ dog_group) %>% 
  adjust_pvalue(method = "BH")
  tt()
```

```{r vf_per_sled_loc_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
  filter(dog_type=="sled_dog") %>%
  ggplot(aes(x=locality,y=vf,color=locality,fill=locality)) +
  geom_jitter(width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 10),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 12, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factors")
```
```{r}
sample_vf %>%
    filter(country=="Greenland") %>%
  kruskal_test(vf ~ locality) %>% 
  adjust_pvalue(method = "BH") %>%
  tt()
```

Significant
```{r}
sample_vf %>%
  filter(country=="Greenland") %>%
  wilcox_test(vf ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj < 0.05)
```

Non-sig
```{r}
sample_vf %>%
  filter(country=="Greenland") %>%
  wilcox_test(vf ~ locality, p.adjust.method = "BH")%>% 
  filter(p.adj > 0.05)
```

With pack as shape
```{r vf_per_sled_outliers_plot, warning=FALSE, comment="", message=FALSE, fig.height=5, fig.width=8, fig.fullwidth=TRUE}
sample_vf %>% 
  filter(dog_type=="sled_dog") %>%
 # filter(!str_starts(sample, "GL_02")) %>% 
  mutate(Pack = as.factor(Pack)) %>%
  ggplot(aes(x=locality,y=vf,color=locality,fill=locality)) +
  geom_jitter(aes(shape = Pack),width = 0.2, alpha=0.7, show.legend = FALSE, size=0.5) +
  geom_boxplot(width = 0.5, alpha=0.5,outlier.shape = NA, show.legend = FALSE) +
  scale_color_manual(values = locality_colors) +
  scale_fill_manual(values = locality_colors) +
  scale_x_discrete(labels = locality_labels) +
  theme_classic() +
  theme(
    axis.text = element_text(size = 7),
    strip.background = element_blank(),
    axis.title.y = element_text(size = 9, margin = margin(t = 0, r = 20, b = 0, l = 0)),
    axis.title.x = element_blank()
    )+
  labs(y="Virulence factors")
```

Max VF value in sled dogs
```{r}
sample_vf %>% 
  filter(country=="Greenland") %>% 
  arrange(-vf) %>% 
  tt()
```

```{r}
sample_vf %>% 
  filter(locality=="Daneborg") %>% 
  arrange(-vf)%>% 
  tt()
```

## Size-function correlations

```{r size_amr_correlation_plot, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>%
  ggplot(aes(x = size, y = AMR, color = country)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, show.legend = FALSE) +
  theme_classic()
```

```{r size_amr_correlation_plot_group, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>%
  ggplot(aes(x = size, y = AMR, color = dog_type)) +
  geom_point() +
  scale_color_manual(values = type_colors, breaks = c("companion", "sled_dog", "wolf"),
      labels = c("Companion dogs", "Sled dogs", "Wolves")) +
  geom_smooth(method = "lm", se = TRUE, show.legend = FALSE) +
  theme_classic()+
  labs(y="AMR", x="Genome size", color = "Canids")
```
```{r size_amr_correlation_plot_together, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>% 
  ggplot(., aes(x = size, y = AMR)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_classic()+
  labs(y="AMR", x="Genome size")
```

```{r}
df <- sample_genome_size %>% 
  left_join(sample_amr %>% select(sample, AMR), by = "sample") %>% 
  left_join(alpha_div, by = "sample")

cor.test(df$size, df$AMR, method = "spearman")

```
```{r}

model <- lm(AMR ~ size, data = df)
summary(model)

model <- lm(AMR ~ size * country, data = df)
summary(model)

df %>%
  group_by(country) %>%
  do(tidy(lm(AMR ~ size, data = .))) %>% 
  filter(term=="size")%>% 
  filter(p.value<0.05) %>% 
  tt()

df %>%
  group_by(dog_type) %>%
  do(tidy(lm(AMR ~ size, data = .))) %>% 
  filter(term=="size") %>% 
  tt()

model <- lm(AMR ~ size * dog_group, data = df)
summary(model)
```

```{r}
df <- sample_genome_size %>% 
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>% 
  left_join(alpha_div, by = "sample")

cor.test(df$size, df$vf, method = "spearman")

model <- lm(vf ~ size, data = df)
summary(model)

model <- lm(vf ~ size * country, data = df)
summary(model)

df%>%
  group_by(country) %>%
  do(tidy(lm(vf ~ size, data = .))) %>% 
  filter(term=="size")%>% 
  filter(p.value<0.05)

df%>%
  group_by(dog_type) %>%
  do(tidy(lm(vf ~ size, data = .))) %>% 
  filter(term=="size")

model <- lm(vf ~ size * dog_group, data = df)
summary(model)
```
```{r size_vf_correlation_plot, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>%
  ggplot(aes(x = size, y = vf, color = country)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, show.legend = FALSE) +
  theme_classic()
```

```{r size_vf_correlation_plot_together, warning=FALSE, comment="", message=FALSE}
sample_genome_size %>%
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>%
  left_join(alpha_div, by = "sample") %>% 
  ggplot(., aes(x = size, y = vf)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"))) +
  theme_classic()
```

**AMR and VF correlation colored by genome size**
```{r size_vf_amr_correlation_plot_group, warning=FALSE, comment="", message=FALSE, fig.height=10, fig.width=12, fig.fullwidth=TRUE}
sample_amr %>%
  left_join(sample_vf %>% select(sample, vf), by = "sample") %>%
  left_join(sample_genome_size %>% select(sample, size), by = "sample") %>% 
  ggplot(., aes(x = AMR, y = vf,color = size)) +
  geom_point(size=3) +
  geom_smooth(method = "lm", se = TRUE) +
    scale_color_viridis_c() +
  scale_size_continuous(range = c(2, 7)) +
  stat_cor(aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~")), size = 6, show.legend = FALSE) +
  theme_classic()+
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 18, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  )+
  labs(color="Genome size")
```

<!--chapter:end:06_AMR_VF.Rmd-->

# Seasonal differences in eastern Greenland

```{r load_data_host_filt_season, comment="", message=FALSE, warning=FALSE}
load("resources/data.Rdata")
load("output/beta_season.Rdata")

sample_metadata <- sample_metadata %>% 
  filter(locality == "Ittoqqortoormiit" & day %in% c("day_1", "unknown"))

genome_counts_filt <- genome_counts_filt %>% 
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>% 
  semi_join(., genome_counts_filt, by = "genome") %>% 
  arrange(match(genome,genome_counts_filt$genome))

genome_metadata_bacteria <- genome_metadata %>% 
  filter(domain=="Bacteria")

genome_tree <- keep.tip(genome_tree, tip=genome_metadata_bacteria$genome)
```

```{r}
sample_metadata %>% 
  group_by(season) %>% 
  summarise(n=n()) %>% 
  tt()
```

## Taxonomy
**Number of bacteria phyla**
```{r phyla_season, comment="", message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

```{r taxonomy_barplot_seasons, fig.height=8, fig.width=15, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% 
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% 
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% 
  filter(count > 0) %>% 
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + 
  geom_bar(stat="identity", colour="white", linewidth=0.1) + 
  scale_fill_manual(values=phylum_colors) +
  facet_grid(~ season, scale="free", space="free")+
  guides(fill = guide_legend(ncol = 1)) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 14, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.y = element_text(size = 12),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line = element_line(
      linewidth = 0.5,
      linetype = "solid",
      colour = "black"
    )
  ) +
  labs(fill = "Phylum", y = "Relative abundance", x = "Samples")
```

### Phylum relative abundances

```{r taxonomy_phylum_season, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>%
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,season) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_phylum_summary_season, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(season, phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T)
  ) %>%
  arrange(-total_mean) %>%
  mutate(total = str_c(round(total_mean, 4), "±", round(total_sd, 3))) %>%
  dplyr::select(phylum, total, season) %>%
  pivot_wider(names_from = season, values_from = total) %>% 
  tt()
```

### Differences in phylum (ancombc)

```{r zero_phylo_season, comment="", message=FALSE, warning=FALSE}
phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()
phylo_genome <- genome_counts_filt %>%
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  filter(genome %in% rownames(phylo_genome)) %>% 
  mutate(genome2 = genome) %>% 
  column_to_rownames("genome2") %>%
  dplyr::select(domain, phylum, class, order, family, genus, species, genome) %>% 
  as.matrix() %>%
  tax_table() 

physeq_genome_itto <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
physeq_genome_itto <- prune_taxa(taxa_sums(physeq_genome_itto)>0, physeq_genome_itto)
```

```{r ancom_rand_phylum_season, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_phylum_itto = ancombc2(
  data = physeq_genome_itto,
  assay_name = "counts",
  tax_level = "phylum",
  fix_formula = "season",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "season",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = TRUE,
  pairwise = TRUE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 20,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100),
  trend_control = NULL
)
```

```{r}
ancom_rand_output_phylum_itto$res %>%
  dplyr::select(taxon, lfc_seasonwinter, q_seasonwinter) %>%
  filter(q_seasonwinter < 0.05) %>%
  dplyr::arrange(q_seasonwinter) %>% 
  tt()
```

## Alpha diversity
```{r alpha_div_season, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  filter(genome %in% genome_metadata_bacteria$genome) %>% 
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample))
```


```{r alpha_plots_seasons, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = value, x = season, group=season, color=season, fill=season)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(. ~factor(metric, labels=c("richness" = "Richness", "neutral" = "Neutral", "phylogenetic" = "Phylogenetic")), scales = "free", ncol = 5) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size =12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```

## Beta diversity
```{r beta_div_season, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  filter(genome %in% genome_metadata_bacteria$genome) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, tree = genome_tree)

save(beta_q0n, 
     beta_q1n, 
     beta_q1p,
     file = "output/beta_season.Rdata")
```

**Richness**
```{r}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]
beta_q0n <- as.dist(beta_q0n_filtered)
```

```{r permanova_q0_season, comment="", message=FALSE, warning=FALSE}
adonis2(beta_q0n ~ season,
  data = sample_metadata %>% arrange(match(sample, labels(beta_q0n))),
  permutations = 999) %>%
  broom::tidy() %>%
  tt()
```

```{r beta_div_rich_season, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q0n %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = season)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Season')
```

**Neutral**
```{r permanova_q1n_filt_seasons, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)
beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]
beta_q1n_dis <- as.dist(beta_q1n_filtered)
sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1n_dis))) %>% 
  column_to_rownames(., "sample")
```
```{r permanova_q1n_season, comment="", message=FALSE, warning=FALSE}
adonis2(beta_q1n_dis ~ season,
  data = sample_metadata_ordered,
  permutations = 999) %>%
  broom::tidy() %>%
  tt()
```
```{r beta_div_neutral_season, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = season)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Season')
```


```{r permanova_q1p_filt_season, comment="", message=FALSE, warning=FALSE, echo=FALSE, eval=FALSE}
#Phylogenetic
beta_q1p_mat <- as.matrix(beta_q1p$S)
beta_q1p_filtered <- beta_q1p_mat[rownames(beta_q1p_mat) %in% sample_metadata$sample,
                        colnames(beta_q1p_mat) %in% sample_metadata$sample]
beta_q1p_dis <- as.dist(beta_q1p_filtered)
sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1p_dis))) %>% 
  column_to_rownames(., "sample")

adonis2(beta_q1p_dis ~ season,
  data = sample_metadata_ordered,
  permutations = 999) %>%
  broom::tidy() %>%
  tt()

beta_q1p_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(aes(x = NMDS1, y = NMDS2, color = season)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Season')
```

## Differences in MAG abundances: Ancombc2

```{r ancom_rand_mag_season, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_mag_itto_all = ancombc2(
  data = physeq_genome_itto,
  assay_name = "counts",
  tax_level = NULL,
  fix_formula = "season",
  p_adj_method = "BH",
  pseudo_sens = TRUE,
  prv_cut = 0,
  lib_cut = 0,
  s0_perc = 0.05,
  group = "season",
  struc_zero = TRUE,
  neg_lb = FALSE,
  alpha = 0.05,
  n_cl = 2,
  verbose = TRUE,
  global = FALSE,
  pairwise = FALSE,
  dunnet = FALSE,
  trend = FALSE,
  iter_control = list(
    tol = 1e-5,
    max_iter = 50,
    verbose = FALSE
  ),
  em_control = list(tol = 1e-5, max_iter = 100),
  mdfdr_control = list(fwer_ctrl_method = "BH", B = 100),
  trend_control = NULL
)
```

#### Structural zeros

***Total number of structural zeros***
```{r}
structural_zeros <- ancom_rand_output_mag_itto_all$zero_ind

structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

structural_zero_taxa <- rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]

structural_zero_groups <- structural_zeros_df[structural_zero_taxa, , drop = FALSE] %>% 
  rename("summer"=1) %>% 
  rename("winter"=2)

nrow(structural_zero_groups)
```

***Structural zeros in summer***
```{r}
structural_zero_groups %>% 
  filter(summer==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(summer==TRUE) %>% 
  rownames_to_column(., "genome") %>% 
  left_join(., genome_metadata, by="genome") %>% 
  count(phylum, name = "sample_count") %>%
  arrange(desc(sample_count)) %>% 
  tt()
```

***Structural zeros in winter***
```{r}
structural_zero_groups %>% 
  filter(winter==TRUE) %>% 
  nrow()

structural_zero_groups %>% 
  filter(winter==TRUE) %>% 
  rownames_to_column(., "genome") %>% 
  left_join(., genome_metadata, by="genome") %>% 
  count(phylum, name = "sample_count") %>%
  arrange(desc(sample_count)) %>% 
  tt()
```

```{r ancom_table_season, comment="", message=FALSE, warning=FALSE}
taxonomy <- data.frame(physeq_genome_itto@tax_table) %>%
  rownames_to_column(., "taxon") %>%
  mutate_at(vars(phylum, order, family, genus, species), ~ str_replace(., "[dpcofgs]__", ""))

ancombc_rand_table_mag_itto <- ancom_rand_output_mag_itto_all$res %>%
  dplyr::select(taxon, lfc_seasonwinter, q_seasonwinter) %>%
  filter(q_seasonwinter < 0.05) %>%
  dplyr::arrange(q_seasonwinter) %>%
  left_join(taxonomy, by = join_by(taxon == taxon)) %>%
  mutate_at(vars(phylum, species), ~ str_replace(., "[dpcofgs]__", ""))%>%
  dplyr::arrange(lfc_seasonwinter)
  
colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by=join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80")) %>% 
	unique() %>%
	dplyr::arrange(phylum)

tax_table <- as.data.frame(unique(ancombc_rand_table_mag_itto$phylum))
  
colnames(tax_table)[1] <- "phylum"
tax_color <- merge(tax_table, colors_alphabetic, by="phylum")%>%
	dplyr::arrange(phylum) %>%
	dplyr::select(colors) %>%
	pull()
```
```{r ancombc_rand_plot_season, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=8, fig.fullwidth=TRUE}
ancombc_rand_table_mag_itto %>%
  mutate(genome=factor(genome,levels=ancombc_rand_table_mag_itto$genome)) %>%
  ggplot(., aes(x=lfc_seasonwinter, y=forcats::fct_reorder(genome,lfc_seasonwinter), fill=phylum)) +
  geom_col() + 
  scale_fill_manual(values=tax_color) + 
  geom_hline(yintercept=0) + 
  theme(panel.background = element_blank(),
        axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_blank(),
        axis.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"),
        legend.position = "right", legend.box = "vertical")+
  geom_text(aes(3.5, 135), label = "Enriched\nin winter", color="#666666") +
  geom_text(aes(-4, 35), label = "Enriched\nin summer", color="#666666")+
  xlab("log2FoldChange") + 
  ylab("Genome")+
  guides(fill=guide_legend(title="Phylum"))
```

Enriched in summer (top 20)
```{r}
ancombc_rand_table_mag_itto %>% 
  filter(lfc_seasonwinter<0) %>% 
  arrange(lfc_seasonwinter) %>% 
  select(lfc_seasonwinter, genus, species) %>% 
  head(20) %>% 
  tt()
```

Enriched in winter (top 20)
```{r}
ancombc_rand_table_mag_itto %>% 
  filter(lfc_seasonwinter>0) %>% 
  arrange(-lfc_seasonwinter) %>% 
  select(lfc_seasonwinter, genus, species)%>% 
  head(20) %>% 
  tt()
```

## Functional differences
```{r gift_analysis_season, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt %>%
  filter(genome_counts_filt$genome %in% rownames(genome_gifts)) %>% 
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))%>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome,]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]

GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

genome_counts_row <- genome_counts_filt %>%
  mutate(across(-genome, ~ . / sum(.))) %>% 
  column_to_rownames(., "genome") 

GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)
```

## Groups MCI
```{r gitfs_functional_season}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(season) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  tt()
```

***Shapiro test***
```{r}
MCI <- GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample))

shapiro.test(MCI$value)
wilcox.test(value ~ season, data=MCI)
```

### Community elements differences:
```{r comunity_elem_season, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(., sample_metadata %>% select(sample,season), by=("sample"))
```

```{r commun_wilcox_elem_season, comment="", message=FALSE, warning=FALSE}
uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

element_gift %>%
  pivot_longer(-c(sample, season), names_to = "trait", values_to = "value") %>%
  group_by(trait) %>%
  summarise(p_value = wilcox.test(value ~ season)$p.value,
            .groups = "drop") %>%
  mutate(p_adjust = p.adjust(p_value, method = "BH")) %>%
  filter(p_adjust < 0.05) %>% 
  left_join(., uniqueGIFT_db, by=join_by("trait"=="Code_element"))
```

<!--chapter:end:07_sled_seasons.Rmd-->

