# Beta diversity

```{r load_data_beta1, comment="", message=FALSE, warning=FALSE}
set.seed(1234)
load("data/data.Rdata")
load("data/beta.Rdata")

genome_metadata <- genome_metadata %>% 
  mutate(phylum = case_when(
        phylum == "Bacillota_A" ~ "Bacillota",
        phylum == "Bacillota_C" ~ "Bacillota",
        phylum == "Bacillota_B" ~ "Bacillota",
        phylum == "Bacillota_I" ~ "Bacillota",
        TRUE ~ phylum))

remove_sled <- read_csv("data/sample_metadata.csv") %>% 
  filter(season=="winter")

sample_metadata <- sample_metadata %>%
  mutate(country = if_else(country == "Shangai domestic dog", "Shangai", country)) %>%
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_type = case_when(
      breed == "Wolf" ~ "wolf",
      country == "Greenland" ~ "sled_dog",
      TRUE  ~ "domestic"
    )
  )

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)
```

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

genome_counts <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))%>%
  rownames_to_column(., "genome")

genome_tree <- keep.tip(genome_tree, tip=genome_counts_filt$genome)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, tree = genome_tree)

save(beta_q0n, 
     beta_q1n, 
     beta_q1p, 
     file = "data/beta.Rdata")
```

### Permanova analysis
```{r}
type_colors <- c('#41B6C0','#E5D388',"#bb6749")
new_labels <- c(
    "domestic" = "Domestic dogs",
    "sled_dog" = "Sled dogs",
    "wolf" = "Wolves"
    )
```

**Richness**
```{r permanova__q0, comment="", message=FALSE, warning=FALSE}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]
beta_q0n_dis <- as.dist(beta_q0n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q0n_dis))) %>% 
  column_to_rownames(., "sample")

betadisper(beta_q0n_dis, sample_metadata_ordered$country) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q0n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q0n_dis ~ dog_type,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

***Locality***

Significant
```{r}
pair_results <- pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$locality, perm = 999)
pair_results %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pair_results %>% 
  filter(p.adjusted>0.05)
```

***Dog group***

Significant
```{r}
pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$dog_type, perm = 999) %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pairwise.adonis(beta_q0n_dis, sample_metadata_ordered$dog_type, perm = 999) %>% 
  filter(p.adjusted>0.05)
```

Centroids
```{r}
betadisper_obj <- betadisper(beta_q0n_dis, sample_metadata_ordered$dog_type)
dist_centroids <- dist(betadisper_obj$centroids)
```

**Neutral**
```{r permanova__q1, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)
beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]
beta_q1n_dis <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1n_dis))) %>% 
  column_to_rownames(., "sample")

all(sample_metadata_ordered$sample == labels(beta_q1n_dis))

betadisper(beta_q1n_dis, sample_metadata_ordered$locality) %>% permutest(., pairwise = TRUE) 
betadisper(beta_q1n_dis, sample_metadata_ordered$dog_type)%>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q1n_dis ~ dog_type + country + locality, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()


adonis2(beta_q1n_dis ~ dog_type,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

***Country***

Significant
```{r}
pair_results <- pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$country, perm = 999)
pair_results %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pair_results %>% 
  filter(p.adjusted>0.05)
```

***Dog group***

Significant
```{r}
pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$dog_type, perm = 999) %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pairwise.adonis(beta_q1n_dis, sample_metadata_ordered$dog_type, perm = 999) %>% 
  filter(p.adjusted>0.05)
```

Centroids
```{r}
betadisper_obj <- betadisper(beta_q1n_dis, sample_metadata_ordered$dog_type)
dist_centroids <- dist(betadisper_obj$centroids)
```

**Phylogenetic**
```{r permanova__qP, comment="", message=FALSE, warning=FALSE}
beta_q1p_mat <- as.matrix(beta_q1p$S)
beta_q1p_filtered <- beta_q1n_mat[rownames(beta_q1p_mat) %in% sample_metadata$sample,
                        colnames(beta_q1p_mat) %in% sample_metadata$sample]
beta_q1p_dis <- as.dist(beta_q1p_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1p_dis))) %>% 
  column_to_rownames(., "sample")

adonis2(beta_q1p_dis ~ dog_type + country, 
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q1p_dis ~ dog_type + country, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()

adonis2(beta_q1p_dis ~ dog_type,
        data = sample_metadata_ordered, 
        permutations = 999) %>%
        broom::tidy() %>%
        tt()
```

***Country***

Significant
```{r}
pair_results <- pairwise.adonis(beta_q1p_dis, sample_metadata_ordered$country, perm = 999)
pair_results %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pair_results %>% 
  filter(p.adjusted>0.05)
```

***Dog group***

Significant
```{r}
pairwise.adonis(beta_q1p_dis, sample_metadata_ordered$dog_type, perm = 999) %>%
  filter(p.adjusted < 0.05)
```

No significant
```{r}
pairwise.adonis(beta_q1p_dis, sample_metadata_ordered$dog_type, perm = 999) %>% 
  filter(p.adjusted>0.05)
```

Centroids
```{r}
betadisper_obj <- betadisper(beta_q1p_dis, sample_metadata_ordered$dog_type)
dist_centroids <- dist(betadisper_obj$centroids)
```

### Plots

#### Richness
***Country***
```{r beta_div_nmds_richness_country, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(country) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = country)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Country')
```

***Dog type***
```{r beta_div_nmds_richness_type, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_type)) +
    geom_point(size = 4) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.9, show.legend = FALSE) +
    scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Canids')
```

#### Neutral
***Country***
```{r beta_div_nmds_neutral_country, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(country) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = country)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = 'Country')
```

***Country***
```{r beta_div_nmds_neutral_country_dog, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(country, dog_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = country)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  # scale_color_manual(name = "Canids",
  #                    values = type_colors,
  #                    labels = new_labels) +
  # scale_fill_manual(name = "Canids",
  #                   values = type_colors,
  #                   labels = new_labels)+
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  )
```

***Dog type***
```{r beta_div_nmds_neutral_type, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q1n_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = 'Canids')
```

##### PCoa country
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100
```

```{r pcoa_neutral_loc, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
centroids <- pcoa_df %>%
  group_by(dog_type, country) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )
centroids_loc <- centroids %>%
  mutate(group = paste(dog_type, country, sep = "_"))

pcoa_df1 <- pcoa_df %>%
  mutate(group = paste(dog_type, country, sep = "_"))

ggplot(pcoa_df1, aes(Axis.1, Axis.2, color = group)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(
    data = centroids_loc,
    aes(Axis.1, Axis.2, fill = group),
    color = "black",
    shape = 21,
    size = 6,
    show.legend = FALSE
  ) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Group"
  )  +
  theme_minimal() +
  theme(legend.text = element_text(size = 8)) +
  guides(color = guide_legend(ncol = 1))
```

***Centroid distances***

```{r centroid_distances_type_locality, comment="", message=FALSE, warning=FALSE, fig.height=10, fig.width=12, fig.fullwidth=TRUE}
centroids_loc <- pcoa_df %>%
  group_by(dog_type, country) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )

centroids_loc <- centroids_loc %>%
  mutate(group = paste(dog_type, country, sep = "_"))


coords <- centroids_loc[, c("Axis.1", "Axis.2")] %>%
  as.data.frame()
rownames(coords) <- centroids_loc$group

dist_loc <- dist(coords)

dist_loc_df <- as.data.frame(as.table(as.matrix(dist_loc))) %>%
  rename(group1 = Var1,
         group2 = Var2,
         distance = Freq) %>%
  filter(group1 != group2)

ggplot(dist_loc_df, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank()) +
  labs(fill = "Centroids distance")
```
```{r}
 dist_loc_df %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(group1, group2)), collapse = "_")) %>%
  ungroup() %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) %>% 
  arrange(distance)
```

##### PCoa dog type
```{r}
pcoa_res <- ape::pcoa(beta_q1n_dis)

pcoa_df <- data.frame(pcoa_res$vectors) %>%
  rownames_to_column("sample") %>%
  left_join(sample_metadata, by = "sample")

var_exp <- pcoa_res$values$Relative_eig * 100

centroids <- pcoa_df %>%
  group_by(dog_type) %>%
  summarise(
    Axis.1 = mean(Axis.1),
    Axis.2 = mean(Axis.2),
    .groups = "drop"
  )
```
```{r pcoa_neutral_dogtype, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
ggplot(pcoa_df, aes(Axis.1, Axis.2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(level = 0.95, linewidth = 1) +
  geom_point(data = centroids,
             aes(Axis.1, Axis.2, fill = dog_type),
             color = "black",
             shape = 21,
             size = 6,
             show.legend = FALSE) +
    scale_color_manual(
  name = "Canid",
  values = type_colors,
  labels = new_labels)+
scale_fill_manual(
  name = "Canid",        
  values = type_colors,
  labels = new_labels) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Canid"
  ) +
  theme_classic(base_size = 14)
```

Show samples closest to domestic dogs
```{r pcoa_neutral_type_sample, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=10, fig.fullwidth=TRUE}
domestic_centroid <- centroids %>% filter(dog_type == "domestic")
sled_centroid <- centroids %>% filter(dog_type == "sled_dog")

pcoa_df <- pcoa_df %>%
  mutate(dist_to_domestic = sqrt((Axis.1 - domestic_centroid$Axis.1)^2 +
                                 (Axis.2 - domestic_centroid$Axis.2)^2),
    dist_to_sled = sqrt((Axis.1 - sled_centroid$Axis.1)^2 +
                        (Axis.2 - sled_centroid$Axis.2)^2)
  )


closest_sled_to_domestic <- pcoa_df %>%
  filter(dog_type == "sled_dog") %>%
  arrange(dist_to_domestic) %>%
  slice_head(n = 5)

closest_samples <- closest_sled_to_domestic$sample

ggplot(pcoa_df, aes(Axis.1, Axis.2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(aes(color = dog_type), level = 0.95, linewidth = 1.2) +
  geom_point(
    data = centroids,
    aes(x = Axis.1, y = Axis.2, fill = dog_type),
    color = "black",
    shape = 21,
    size = 6,
    show.legend = FALSE
  ) +
  geom_point(
    data = pcoa_df %>% filter(sample %in% closest_samples),
    aes(x = Axis.1, y = Axis.2),
    shape = 21,
    fill = NA,
    color = "black",
    size = 5,
    stroke = 1.5
  ) +
  scale_color_manual(name = "Canid",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canid",
                    values = type_colors,
                    labels = new_labels) +
  geom_text_repel(
    data = closest_sled_to_domestic,
    aes(x = Axis.1, y = Axis.2, label = locality),
    size = 4,
    box.padding = 0.5,
    point.padding = 0.5,
    segment.color = "grey50"
  ) +
  labs(
    x = paste0("PCoA1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PCoA2 (", round(var_exp[2], 1), "%)"),
    color = "Dog type"
  ) +
  theme_classic(base_size = 14)
```

**Centroid distances**
```{r}
centroid_coords <- centroids[, c("Axis.1", "Axis.2")]
rownames(centroid_coords) <- centroids$dog_type

centroid_dist <- dist(centroid_coords)
centroid_dist_mat <- as.matrix(centroid_dist)

dist_table <- as.data.frame(as.table(centroid_dist_mat)) %>%
  rename(group1 = Var1, group2 = Var2, distance = Freq) %>%
  filter(group1 != group2) %>%
  arrange(distance)
```

```{r centroid_distances_type_figure, comment="", message=FALSE, warning=FALSE, fig.height=3, fig.width=6, fig.fullwidth=TRUE}
ggplot(dist_table, aes(x = group1, y = group2, fill = distance)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=45, hjust=1),
        axis.title = element_blank()) +
  labs(fill="Centroids distance")
```

##### Distance comparison test

This tests whether samples of the same dog type are more similar than different ones
```{r}
dist_mat <- as.matrix(beta_q1n_dis)

dist_df <- as.data.frame(dist_mat) %>%
  mutate(sample1 = rownames(.)) %>%
  pivot_longer(-sample1, names_to = "sample2", values_to = "distance") %>%
  filter(sample1 < sample2)

meta <- sample_metadata_ordered %>%
  mutate(sample = rownames(.))

dist_df <- dist_df %>%
  left_join(meta, by = c("sample1" = "sample")) %>%
  rename(dog1 = dog_type, loc1 = locality) %>%
  left_join(meta, by = c("sample2" = "sample")) %>%
  rename(dog2 = dog_type, loc2 = locality)

dist_df <- dist_df %>%
  mutate(
    same_dog = dog1 == dog2,
    same_locality = loc1 == loc2
  )
```

```{r}
dist_cross_loc <- dist_df %>%
  filter(!same_locality)

wilcox.test(distance ~ same_dog, data = dist_cross_loc)

ggplot(dist_cross_loc, aes(x = same_dog, y = distance)) +
  geom_boxplot() +
  labs(
    x = "Same dog type",
    y = "Beta diversity distance",
    title = "Distance comparison (across localities only)"
  )
```
If same_dog == TRUE has lower distances: suggests similarity within dog types across locations

#### Phylogenetic
***Country***
```{r beta_div_nmds_phylo_country, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q1p_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(country) %>%
  mutate(x_cen = median(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = median(NMDS2, na.rm = TRUE)) %>%
  ungroup()  %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = country)) +
    geom_point(size = 3, alpha = 0.5) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Country')
```

***Dog type***
```{r .beta_div_nmds_phylo_type, comment="", message=FALSE, warning=FALSE, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q1p_dis %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(dog_type) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = dog_type)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
  scale_color_manual(name = "Canids",
                     values = type_colors,
                     labels = new_labels) +
  scale_fill_manual(name = "Canids",
                    values = type_colors,
                    labels = new_labels) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(face = "bold", size = 12),
    panel.background = element_blank(),
    axis.line = element_line(
      size = 0.5,
      linetype = "solid",
      colour = "black"
    ),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.position = "right",
    legend.box = "vertical"
  ) +
  labs(color = 'Canids')
```