```{r load_all_dif, comment="", message=FALSE,warning=FALSE}
load("data/data.Rdata")
genome_metadata <- genome_metadata %>% 
  mutate(phylum = case_when(
        phylum == "Bacillota_A" ~ "Bacillota",
        phylum == "Bacillota_C" ~ "Bacillota",
        phylum == "Bacillota_B" ~ "Bacillota",
        phylum == "Bacillota_I" ~ "Bacillota",
        TRUE ~ phylum))

remove_sled <- read_csv("data/sample_metadata.csv") %>% 
  filter(season=="winter")#63

sample_metadata <- sample_metadata %>%
  mutate(country = if_else(country == "Shangai domestic dog", "Shangai", country)) %>% 
  filter(!sample %in% remove_sled$sample) %>%
  mutate(
    dog_type = case_when(
      breed == "Wolf"        ~ "wolf",
      country == "Greenland" ~ "sled_dog",
      TRUE                   ~ "domestic"
    )
  )

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)
```

```{r zero_phylo1, comment="",  message=FALSE, warning=FALSE}
#load("data/ancombc_all.Rdata")
phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()
phylo_genome <- genome_counts_filt %>%
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  column_to_rownames("genome") %>%
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  filter(genome %in% rownames(phylo_genome)) %>%
  mutate(genome2 = genome) %>% 
  column_to_rownames("genome2") %>%
  dplyr::select(domain, phylum, class, order, family, genus, species, genome) %>% 
  as.matrix() %>%
  tax_table()

physeq_genome <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples)
region_pairs_country <- combn(unique(sample_metadata$country), 2, simplify = FALSE)
```

## Phylum
```{r}
load(file = "data/ancombc_phylum_sled_dogs_wolf.Rdata")
```

```{r pair_all_phylum, comment="", message=FALSE,warning=FALSE, eval=FALSE}
results_species <- lapply(region_pairs_species, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

   fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome),
    fix_formula = "dog_type",
    tax_level = "phylum", 
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )
   # Extract main results table
  res_table <- fit$res

  # Add comparison label
  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})
names(results_species) <- sapply(region_pairs_species, paste, collapse = "_vs_")
save(results_species,
     file = "data/ancombc_phylum_sled_dogs_wolf.Rdata")
```


### Domestic vs sled_dog

**Enriched in domestic dogs**
```{r}
domestic <- results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
domestic %>%
  nrow() %>% 
  cat()
```

```{r}
domestic
```

**Enriched in sled dogs**

```{r}
sled <- results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)
```

```{r}
sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled
```


### Sled_dog vs wolves

**Enriched in sled dogs**
```{r}
sled <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)
```

```{r}
sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled
```

**Enriched in wolves**

```{r}
wolf <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)
```

```{r}
wolf %>%
  nrow() %>%
  cat()
```

```{r}
wolf
```

### Plots
```{r}
taxa <- as.data.frame(physeq_genome@tax_table)  %>% 
  distinct(phylum)

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxa, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (name in names(results_species)) {

  parts <- strsplit(name, "_vs_")[[1]]
  
  name1 <- parts[1]
  name2 <- parts[2]
  
  title <- paste0("Enriched in ", name1, "  ( − LFC )   vs   ", "Enriched in ", name2, "  ( + LFC )")
  
  tab <- results_species[[name]]

  coef <- grep("^lfc_dog_type", colnames(tab), value = TRUE)[1]
  qcol <- sub("^lfc_", "q_", coef)
  
  plot_data <- tab %>%
    filter(.data[[qcol]] < 0.05) %>%
    mutate(logFC = .data[[coef]])
  
  tax_color <- plot_data %>% 
    rename(phylum=1) %>% 
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()

  p <- ggplot(plot_data, aes(x = reorder(taxon, logFC), y = logFC, fill = taxon)) +
    geom_col() +
    scale_fill_manual(values = tax_color) +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    labs(
      title = title,
      x = "Phylum",
      y = "Log-Fold Change",
      fill = "Phylum"
    )

  ggsave(paste0("Figures/Fig_phylum_", name, ".png"),
         plot = p,
         width = 6,
         height = 8)
}
```



# Sled dogs vs domestic dogs vs wolves

## MAG level
```{r load_all, comment="", message=FALSE,warning=FALSE}
region_pairs_species <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
load("data/ancombc_pairwise_mag_sled_dogs_sepa.Rdata")
```

```{r eval=FALSE}
results_species <- lapply(region_pairs_species, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

  fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome),
    fix_formula = "dog_type",
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )

  # Extract main results table
  res_table <- fit$res

  # Add comparison label
  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})

# Name each table for clarity
names(results_species) <- sapply(region_pairs_species, paste, collapse = "_vs_")
save(results_species,
     file = "data/ancombc_pairwise_mag_sled_dogs_sepa.Rdata")
```


### domestic_vs_sled_dog

**Total number of enriched MAGs**
```{r}
results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  nrow
```

**Enriched in domestic dogs**
```{r}
domestic <- results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
all <- domestic %>%
  nrow()
all %>% 
  cat()
```

how many of the enriched bacteria were not annotated to species
```{r}
noanno <- domestic %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

```{r}
perc <- 100-(noanno*100/all)
perc %>% 
  cat()
```

```{r}
domestic %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, genus, family, everything()) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
domestic %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, genus, family, phylum, everything()) %>% 
  filter(species=="") %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count)
```
```{r}
domestic %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(genus, family, everything()) %>% 
  filter(genus=="") 
```

```{r}
domestic %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

```{r}
domestic %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

**Enriched in sled dogs**

```{r}
sled <- results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)
```

```{r}
all <- sled %>%
  nrow()
all %>% 
  cat()
```

how many of the enriched bacteria were not annotated to species
```{r}
noanno <- sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```
```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(genus, family, everything()) %>% 
  filter(genus=="") %>% 
  nrow()
```

Percentage of bacteria annotated to species level
```{r}
perc <- 100-(noanno*100/all)
perc %>% 
  cat()
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, genus, family, everything())
```
```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, genus, family, phylum, everything()) %>% 
  filter(species=="") %>% 
  group_by(phylum, family, genus) %>% 
  summarise(count=n()) %>% 
  arrange(-count)
```
```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)

sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```


```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(genus, family, everything()) %>% 
  filter(genus=="") 
```

### sled_dog vs wolves

**Total number of enriched MAGs**
```{r}
results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  nrow
```

**Enriched in sled dogs**
```{r}
sled <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)
```

```{r}
all <- sled %>%
  nrow()
all %>% 
  cat()
```

how many of the enriched bacteria were not annotated to species
```{r}
noanno <- sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```
```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, family, everything())
```

```{r}
sled %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

**Enriched in wolves**

```{r}
wolf <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)
```

```{r}
all <- wolf %>%
  nrow()
all %>% 
  cat()
```

how many of the enriched bacteria were not annotated to species
```{r}
noanno <- wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```

```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  select(species, family, everything())
```

```{r}
wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus, species) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)

wolf %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome"))%>% 
  group_by(family, genus) %>% 
  summarise(Mags=n()) %>% 
  arrange(-Mags)
```

### Plots
```{r}
taxonomy <- data.frame(physeq_genome@tax_table) %>%
  distinct(phylum) %>% 
  rownames_to_column(., "taxon")

colors_alphabetic <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  mutate_at(vars(phylum), ~ str_replace(., "[dpcofgs]__", ""))  %>%
  right_join(taxonomy, by = join_by(phylum == phylum)) %>%
  dplyr::select(phylum, colors) %>%
  mutate(colors = str_c(colors, "80"))  %>% 
  unique() %>%
  dplyr::arrange(phylum)

for (name in names(results_species)) {

  parts <- strsplit(name, "_vs_")[[1]]
  
  name1 <- parts[1]
  name2 <- parts[2]
  
  title <- paste0("Enriched in ", name1, "  ( − LFC )   vs   ", "Enriched in ", name2, "  ( + LFC )")
  
  tab <- results_species[[name]]

  coef <- grep("^lfc_dog_type", colnames(tab), value = TRUE)[1]
  qcol <- sub("^lfc_", "q_", coef)
  
  plot_data <- tab %>%
    filter(.data[[qcol]] < 0.05) %>%
    mutate(logFC = .data[[coef]]) %>% 
    left_join(genome_metadata, by=join_by("taxon"=="genome"))
  
  tax_color <- plot_data %>% 
    distinct(phylum) %>% 
    left_join(., colors_alphabetic, by = "phylum") %>%
    dplyr::arrange(phylum) %>%
    dplyr::select(colors) %>%
    pull()

  p <- ggplot(plot_data, aes(x = reorder(taxon, logFC), y = logFC, fill = phylum)) +
    geom_col() +
    scale_fill_manual(values = tax_color) +
    coord_flip() +
    theme(
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 6),
      axis.title = element_text(size = 14, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14, face = "bold"),
      legend.position = "right",
      legend.box = "vertical",
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    ) +
    labs(
      title = title,
      x = "MAGs",
      y = "Log-Fold Change",
      fill = "Phylum"
    )

  ggsave(paste0("Figures/Fig_mag_", name, ".png"),
         plot = p,
         width = 10,
         height = 25)
}
```

## Species level
```{r}
genome_tax <- genome_metadata %>%
  mutate(
    class  = if_else(is.na(class), paste0(phylum, "_Unclassified"), class),
    order  = if_else(is.na(order),
                     if_else(str_detect(class, "Unclassified"), class, paste0(class, "_Unclassified")),
                     order),
    family = if_else(is.na(family),
                     if_else(str_detect(order, "Unclassified"), order, paste0(order, "_Unclassified")),
                     family),
    genus  = if_else(genus=="",
                     if_else(str_detect(family, "Unclassified"), family, paste0(family, "_Unclassified")),
                     genus),
    species  = if_else(species=="",
                     if_else(str_detect(genus, "Unclassified"), genus, paste0(genus, "_Unclassified")),
                     species)
  )
```

```{r}
phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  left_join(genome_tax[c(1,3,4,6,7,8)], by = "genome") %>% 
  select(-genome) %>%
  group_by(phylum, class, family, genus, species) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  select(-c(phylum,class, family, genus))%>%
  column_to_rownames("species") %>%
  otu_table(., taxa_are_rows = TRUE)

physeq_genome_sp <- phyloseq(phylo_genome, phylo_samples)
region_pairs_species <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
load(file = "data/ancombc_species_sled_dogs_wolf.Rdata")
```

```{r pair_all_dogtype, comment="", message=FALSE,warning=FALSE, eval=FALSE}
results_species <- lapply(region_pairs_species, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

   fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome_sp),
    fix_formula = "dog_type",
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )
   # Extract main results table
  res_table <- fit$res

  # Add comparison label
  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})
names(results_species) <- sapply(region_pairs_species, paste, collapse = "_vs_")
 save(results_species,
     file = "data/ancombc_species_sled_dogs_wolf.Rdata")
```

### domestic_vs_sled_dog

**Enriched in domestic dogs**
```{r}
domestic <- results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
all <- domestic %>%
  nrow()
all %>% 
  cat()
```

how many of the enriched bacteria were not annotated to species
```{r}
noanno <- domestic %>% 
  left_join(genome_metadata, by=join_by("taxon"=="genome")) %>% 
  select(species, family, everything()) %>% 
  filter(species=="") %>% 
  nrow()
noanno %>% 
  cat()
```
```{r}
perc <- noanno*100/all 
perc %>% 
  cat()
```

```{r}
domestic
```


**Enriched in sled dogs**

```{r}
sled <- results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog>0) %>% 
  arrange(-lfc_dog_typesled_dog)
```

```{r}
all <- sled %>%
  nrow()
all %>% 
  cat()
```

```{r}
sled
```


### sled_dog vs wolves

**Enriched in sled dogs**
```{r}
sled <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf < 0) %>% 
  arrange(lfc_dog_typewolf)
```

```{r}
sled %>%
  nrow() %>% 
  cat()
```

```{r}
sled
```

**Enriched in wolves**

```{r}
wolf <- results_species$sled_dog_vs_wolf %>% 
  filter(q_dog_typewolf < 0.05) %>% 
  filter(lfc_dog_typewolf > 0) %>% 
  arrange(-lfc_dog_typewolf)
```

```{r}
wolf %>%
  nrow() %>%
  cat()
```

```{r}
wolf
```

## Phylum

```{r}
phylo_samples <- sample_metadata %>%
  column_to_rownames("sample") %>%
  sample_data()

phylo_genome <- genome_counts_filt %>% 
  select(one_of(c("genome", rownames(phylo_samples)))) %>%
  left_join(genome_metadata[c(1,3,4,6,8)], by = "genome") %>% 
  select(-genome) %>%
  group_by(phylum) %>%
  summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE)), .groups = "drop")%>% 
  as.data.frame() %>% 
  column_to_rownames("phylum") %>%
  otu_table(., taxa_are_rows = TRUE)

physeq_genome_phylum <- phyloseq(phylo_genome, phylo_samples)
region_pairs_species <- combn(unique(sample_metadata$dog_type), 2, simplify = FALSE)
load(file = "data/ancombc_phylum_sled_dogs_wolf.Rdata")
```

```{r pair_all_type2, comment="", message=FALSE,warning=FALSE, eval=FALSE}
results_species <- lapply(region_pairs_species, function(pair) {
  subset_idx <- sample_metadata$dog_type %in% pair

   fit <- ancombc2(data = prune_samples(subset_idx, physeq_genome_phylum),
    fix_formula = "dog_type",
    group = "dog_type",
    global = FALSE,
    pairwise = FALSE,
    p_adj_method = "BH",
    lib_cut = 0,
    struc_zero = TRUE,
    alpha = 0.05
  )
   # Extract main results table
  res_table <- fit$res

  # Add comparison label
  res_table$comparison <- paste(pair, collapse = "_vs_")

  return(res_table)
})
names(results_species) <- sapply(region_pairs_species, paste, collapse = "_vs_")
 save(results_species,
     file = "data/ancombc_phylum_sled_dogs_wolf.Rdata")
```

### domestic_vs_sled_dog

**Enriched in domestic dogs**
```{r}
domestic <- results_species$domestic_vs_sled_dog %>% 
  filter(q_dog_typesled_dog < 0.05) %>% 
  filter(lfc_dog_typesled_dog < 0) %>% 
  arrange(lfc_dog_typesled_dog)
```

```{r}
domestic %>%
  nrow() %>% 
  cat()
```

```{r}
domestic
```


